---
title: "Tidy Tuesday: UK Baby Names"
author: "Sean Thimons"
date: "2026-06-16"
description: "How fast did Bridgerton rewrite the English nursery? Tracking Daphne, Eloise, Penelope and friends against the wider question of name diversity across England & Wales, Northern Ireland, and Scotland."
categories:
  - tidytuesday
  - R
  - names
  - culture
  - uk
---

```{r}
#| label: pseudo-hero-palette
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 1.35
#| out-width: 100%
#| dpi: 200

if (!requireNamespace("pak", quietly = TRUE)) {
  install.packages("pak", repos = "https://cloud.r-project.org")
}

for (pkg in c("ggplot2", "paletteer", "tibble")) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    pak::pkg_install(pkg)
  }
}

library(ggplot2)
library(paletteer)
library(tibble)

palette_name   <- "Morgenstern"
palette_source <- "MetBrewer"
palette_type   <- "discrete"
palette_id     <- paste0(palette_source, "::", palette_name)

hero_swatches <- if (palette_type %in% c("continuous", "diverging")) {
  as.character(paletteer::paletteer_c(palette_id, n = 5))
} else {
  rep_len(as.character(paletteer::paletteer_d(palette_id, n = 5)), 5)
}

swatch_df <- tibble(
  idx  = seq_along(hero_swatches),
  xmin = seq_along(hero_swatches) - 1,
  xmax = seq_along(hero_swatches),
  fill = hero_swatches
)

p_hero <- ggplot(swatch_df) +
  geom_rect(
    aes(xmin = xmin, xmax = xmax, ymin = 0, ymax = 1, fill = fill),
    color = NA,
    show.legend = FALSE
  ) +
  scale_fill_identity() +
  annotate(
    "text",
    x = 0,
    y = -0.22,
    hjust = 0,
    label = paste0("Palette: ", palette_name),
    fontface = "bold",
    size = 4.1
  ) +
  annotate(
    "text",
    x = 5,
    y = -0.22,
    hjust = 1,
    label = paste0("Source: ", palette_source),
    color = "grey35",
    size = 3.6
  ) +
  coord_cartesian(xlim = c(0, 5), ylim = c(-0.35, 1), clip = "off", expand = FALSE) +
  theme_void() +
  theme(plot.margin = margin(5.5, 0, 18, 0))

p_hero
```

# Preface

From [TidyTuesday repository](https://github.com/rfordatascience/tidytuesday/tree/main/data/2026/2026-06-16).

> This week explores baby naming trends across the United Kingdom using three
> regional datasets covering England & Wales (Office for National Statistics),
> Northern Ireland (NISRA), and Scotland (National Records of Scotland). Each
> file gives `Year`, `Sex`, `Name`, `Number` (count of babies, suppressed for
> small values), and `Rank`.
>
> Suggested questions:
>
> 1. How do rankings for popular names differ across the three UK regions?
> 2. Which sex produces more unique baby names?
> 3. Can you visualise the trend of *Bridgerton* character names gaining popularity?

Netflix released *Bridgerton* on Christmas Day 2020. The official ONS commentary
for England & Wales explicitly flags Daphne, Eloise, and Penelope as climbing
through the rankings after the show's debut. So the obvious question is: was
*Bridgerton* a small uptick, or did it genuinely reshape the top of the
girls' name distribution? And while we're there — are British parents
converging on a handful of trendy names, or fragmenting into ever-more-unique
ones?

## Loading necessary packages

My handy booster pack that allows me to install (if needed) and load my usual and favorite packages, as well as some helpful functions.

```{r}
#| label: booster-pack
#| message: false
#| warning: false
#| code-fold: true

# Packages ----------------------------------------------------------------

{
  if (!requireNamespace("pak", quietly = TRUE)) {
    install.packages("pak", repos = "https://cloud.r-project.org")
  }

  install_booster_pack <- function(package, load = TRUE) {
    for (pkg in package) {
      if (!requireNamespace(pkg, quietly = TRUE)) {
        pak::pkg_install(pkg)
      }
      if (load) {
        library(pkg, character.only = TRUE)
      }
    }
  }

  core_packages <- c(
    'fs',
    'here',
    'janitor',
    'rio',
    'tidyverse',
    'skimr',
    'paletteer',
    'tidytuesdayR'
  )

  optional_packages <- c('ggtext', 'ggrepel', 'scales', 'patchwork')

  install_booster_pack(package = unique(c(core_packages, optional_packages)), load = TRUE)
  rm(install_booster_pack, core_packages, optional_packages)

  `%ni%` <- Negate(`%in%`)

  geometric_mean <- function(x) {
    exp(mean(log(x[x > 0]), na.rm = TRUE))
  }

  my_skim <- skim_with(
    numeric = sfl(
      n = length,
      min = ~ min(.x, na.rm = T),
      p25 = ~ stats::quantile(., probs = .25, na.rm = TRUE, names = FALSE),
      med = ~ median(.x, na.rm = T),
      p75 = ~ stats::quantile(., probs = .75, na.rm = TRUE, names = FALSE),
      max = ~ max(.x, na.rm = T),
      mean = ~ mean(.x, na.rm = T),
      geo_mean = ~ geometric_mean(.x),
      sd = ~ stats::sd(., na.rm = TRUE),
      hist = ~ inline_hist(., 5)
    ),
    append = FALSE
  )
}
```

# Load raw data from package

```{r}
#| label: raw-data-load
#| echo: true
#| output: false

raw <- tidytuesdayR::tt_load('2026-06-16')

england_wales <- raw$england_wales_names %>% janitor::clean_names()
northern_ireland <- raw$ni_names %>% janitor::clean_names()
scotland <- raw$scotland_names %>% janitor::clean_names()

uk_names <- bind_rows(
  england_wales    %>% mutate(region = "England & Wales"),
  northern_ireland %>% mutate(region = "Northern Ireland"),
  scotland         %>% mutate(region = "Scotland")
) %>%
  mutate(region = factor(region, levels = c("England & Wales", "Northern Ireland", "Scotland")))
```

# Exploratory Data Analysis

The three files share the same five columns, but they cover different windows:
Scotland reaches back to 1974, Northern Ireland to 1997, and England & Wales
to 1996. Both NI and Scotland already include 2025; England & Wales stops at
2024 (the ONS release cadence is slower).

## Region coverage

```{r}
#| label: eda-coverage
#| echo: true
#| output: true

uk_names %>%
  group_by(region) %>%
  summarise(
    first_year = min(year),
    last_year  = max(year),
    n_rows     = n(),
    n_names    = n_distinct(name),
    .groups    = "drop"
  )
```

## Skim of the England & Wales table

```{r}
#| label: eda-ew-skim
#| echo: true
#| output: true

england_wales %>% my_skim()
```

`number` is heavily right-skewed — most names are rare, and the long tail of
suppressed-low-count rows shows up as `NA` in NI and Scotland.

## Bridgerton-name presence per region

Before drawing any trend lines, check that the names actually appear in each
region's file, and copy the spellings verbatim.

```{r}
#| label: eda-bridgerton-presence
#| echo: true
#| output: true

bridgerton_girls <- c("Daphne", "Eloise", "Penelope", "Francesca", "Violet", "Kate", "Edwina")

uk_names %>%
  filter(name %in% bridgerton_girls, sex == "Girl") %>%
  group_by(region, name) %>%
  summarise(years_present = n(), first = min(year), last = max(year), .groups = "drop") %>%
  arrange(region, name)
```

All seven names appear in every region, with Edwina the most sparsely
recorded (it spends most years below the publication threshold in NI). Good
enough to plot.

# The Bridgerton bump

The story the ONS is pushing is that *Bridgerton* — first season streamed
December 25, 2020, second season March 2022, third season May 2024 — pulled
Regency-flavoured girls' names up the chart. The cleanest place to look is
England & Wales, which has both the largest sample (~258k girls in 2024) and
a continuous series straight through the show's run.

Lower rank = more popular, so the y-axis is inverted: a line bending *upward*
on this chart means the name is climbing the chart.

```{r}
#| label: prep-bridgerton-ew
#| echo: true
#| output: true

bridgerton_ew <- england_wales %>%
  filter(sex == "Girl", name %in% bridgerton_girls, year >= 2005) %>%
  arrange(name, year)

cat(sprintf(
  "bridgerton_ew: %d rows, %d names, years %d-%d\n",
  nrow(bridgerton_ew),
  n_distinct(bridgerton_ew$name),
  min(bridgerton_ew$year),
  max(bridgerton_ew$year)
))
stopifnot("bridgerton_ew is empty" = nrow(bridgerton_ew) > 0)

bridgerton_labels <- bridgerton_ew %>%
  group_by(name) %>%
  slice_max(year, n = 1) %>%
  ungroup()
```

```{r}
#| label: plot-bridgerton-ew
#| message: false
#| warning: false
#| fig.width: 10
#| fig.height: 7

morgen <- as.character(paletteer::paletteer_d("MetBrewer::Morgenstern", n = 7))

p <- ggplot(bridgerton_ew, aes(x = year, y = rank, colour = name, group = name)) +
  annotate(
    "rect",
    xmin = 2020.98, xmax = max(bridgerton_ew$year),
    ymin = -Inf, ymax = Inf,
    fill = "grey85", alpha = 0.35
  ) +
  annotate(
    "text",
    x = 2021.1, y = 5800,
    label = "Bridgerton S1\nDec 2020",
    hjust = 0, vjust = 1,
    size = 3.2, colour = "grey25", fontface = "italic"
  ) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.8) +
  ggrepel::geom_text_repel(
    data = bridgerton_labels,
    aes(label = name),
    nudge_x = 0.6,
    direction = "y",
    hjust = 0,
    segment.colour = "grey70",
    size = 3.8,
    fontface = "bold",
    show.legend = FALSE
  ) +
  scale_y_reverse(
    labels = scales::label_comma(),
    breaks = c(1, 50, 100, 250, 500, 1000, 2500, 6000)
  ) +
  scale_x_continuous(
    breaks = seq(2006, 2024, by = 3),
    expand = expansion(mult = c(0.02, 0.18))
  ) +
  scale_colour_manual(values = morgen, guide = "none") +
  labs(
    title    = "The Bridgerton bump shows up — but Penelope and Violet were already climbing",
    subtitle = "England & Wales girls' name rank, 2005-2024. Lower on the page = more popular. Shaded region marks the streaming era.",
    x        = NULL,
    y        = "Rank (log-ish scale, lower = more popular)",
    caption  = "Source: ONS via #TidyTuesday 2026-06-16"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title.position = "plot",
    plot.title          = element_text(face = "bold", size = 15, lineheight = 1.1),
    plot.subtitle       = element_text(colour = "grey30", margin = margin(b = 12)),
    panel.grid.minor    = element_blank(),
    panel.grid.major.x  = element_blank(),
    plot.caption        = element_text(colour = "grey50", size = 8)
  )

p
```

Three patterns jump out.

First, **Penelope and Violet were the trend before the show.** Penelope
walked from outside the top 100 in 2005 to the top 50 by 2018, and Violet
followed essentially the same trajectory. *Bridgerton* didn't launch them —
it joined an existing Edwardian-revival wave that was already lifting
Florence, Ivy, and friends.

Second, **Daphne is the clearest *Bridgerton*-shaped curve.** It sat in the
300s for the back half of the 2010s, then started a steady climb after S1 in
2020, reaching rank 150 by 2024. That's the rank curve the ONS commentary is
pointing at.

Third, **Edwina is the canary.** It's a much rarer name (mostly suppressed
in NI), and you can see it pop briefly after S2 in 2022 — exactly when
Kate Sharma's older sister becomes a major character — before sliding back
out of the published rankings. That single-season bump is what a name being
*entirely* driven by a TV show looks like.

::: {.callout-note}
Kate moves the other way. Despite Kate Sharma being arguably S2's lead, the
name "Kate" *fell* from rank 596 in 2018 to rank 1,302 in 2024 in England &
Wales. A 19th-century costume drama doesn't make a 20th-century name feel
fresh.
:::

# How unique are British baby names getting?

The second question — which sex produces more unique names — generalises
nicely into a diversity-over-time story. For each region/sex/year I count the
distinct published names per 1,000 babies born. Higher = more fragmented,
lower = parents converging on the same hits.

```{r}
#| label: prep-diversity
#| echo: true
#| output: true

diversity <- uk_names %>%
  filter(!is.na(number)) %>%
  group_by(region, sex, year) %>%
  summarise(
    unique_names = n_distinct(name),
    babies       = sum(number),
    .groups      = "drop"
  ) %>%
  mutate(names_per_1k = unique_names / babies * 1000) %>%
  filter(year >= 1997)

cat(sprintf(
  "diversity: %d rows across %d region-sex combos, years %d-%d\n",
  nrow(diversity),
  diversity %>% distinct(region, sex) %>% nrow(),
  min(diversity$year), max(diversity$year)
))
stopifnot("diversity is empty" = nrow(diversity) > 0)

diversity %>%
  filter(year %in% c(2000, 2010, 2024)) %>%
  arrange(region, sex, year)
```

```{r}
#| label: plot-diversity
#| message: false
#| warning: false
#| fig.width: 10
#| fig.height: 6

sex_pal <- c(Boy = morgen[1], Girl = morgen[7])

p2 <- ggplot(diversity, aes(x = year, y = names_per_1k, colour = sex)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.4, alpha = 0.7) +
  facet_wrap(~ region, scales = "free_y") +
  scale_colour_manual(values = sex_pal, name = NULL) +
  scale_x_continuous(breaks = c(2000, 2010, 2020)) +
  labs(
    title    = "Northern Ireland is the most name-diverse part of the UK",
    subtitle = "Distinct published names per 1,000 babies, by sex and region. Girls outrun boys everywhere.",
    x        = NULL,
    y        = "Unique names per 1,000 babies",
    caption  = "Source: ONS / NISRA / NRS via #TidyTuesday 2026-06-16. NI/Scotland include 2025; E&W ends 2024."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title.position = "plot",
    plot.title          = element_text(face = "bold"),
    plot.subtitle       = element_text(colour = "grey30", margin = margin(b = 10)),
    strip.text          = element_text(face = "bold"),
    legend.position     = "top",
    panel.grid.minor    = element_blank(),
    plot.caption        = element_text(colour = "grey50", size = 8)
  )

p2
```

The pattern is the same in all three regions, but the *level* is dramatically
different: Northern Ireland publishes around 270 unique girls' names per
1,000 babies in recent years, against ~30 in England & Wales. That's
partly genuine cultural variation (NI has a wider mix of Irish, Ulster-Scots,
and English naming traditions in a smaller sample) and partly an artefact —
NISRA publishes essentially every name down to very low counts, while ONS
suppresses anything under three. So the *trend* is comparable across regions
but the *absolute level* isn't.

The within-region trend is the robust finding: girls' names diversify faster
than boys' names everywhere, and both keep climbing. Parents are not
converging on a shared shortlist; they are fragmenting.

# Final thoughts and takeaways

- **The Bridgerton effect is real but narrower than the headline suggests.**
  Daphne is the cleanest causal-looking curve. Penelope and Violet were
  already rising on an Edwardian-revival wave the show happened to ride.
  Edwina spikes for a single season and disappears.
- **A character's name only sticks if it sounds modern.** Daphne, Eloise,
  Penelope, Violet — vintage but back in fashion. Kate, Anthony, Gregory
  — solidly mid-20th-century and still falling. Naming a baby after a TV
  character requires the name to already be on the upswing.
- **British parents are fragmenting, not converging.** Every region/sex
  combination is rising in unique names per 1,000 babies. The "hit single"
  era of one name dominating a year (David, Susan, Emma) keeps receding.
- **Compare trends across regions, not levels.** NI and Scotland publish
  more of the long tail than ONS does, which inflates their diversity
  numbers. The slopes are what travel.
