---
title: "Tidy Tuesday: The Complete Sherlock Holmes"
author: "Sean Thimons"
date: "2025-11-18"
description: "A text analysis of the complete Sherlock Holmes canon — exploring sentence patterns, story lengths, and the linguistic fingerprints that distinguish Watson's narration from Holmes's dialogue."
categories: [tidytuesday, R, text-analysis, literature]
---

# Preface

From [TidyTuesday repository](https://github.com/rfordatascience/tidytuesday/blob/main/data/2025/2025-11-18/readme.md).

> This collection contains the full line-by-line text of Sir Arthur Conan Doyle's Sherlock Holmes stories and novels, organized by book and line number. The dataset is sourced from the sherlock R package by Emil Hvitfeldt and is designed for stylometric analysis, sentiment examination, and literary exploration.
>
> -   How do Watson's narration patterns differ from Holmes's speech patterns?
> -   What variations exist in sentence length across different stories?
> -   Does tone shift when comparing Watson's narration to Holmes's direct dialogue?

## Loading necessary packages

My handy booster pack that allows me to install (if needed) and load my usual and favorite packages, as well as some helpful functions.

```{r}
#| label: booster-pack
#| message: false
#| warning: false
#| code-fold: true

# Packages ----------------------------------------------------------------

{
  if (!requireNamespace("pak", quietly = TRUE)) {
    install.packages(
      "pak",
      repos = sprintf(
        "https://r-lib.github.io/p/pak/stable/%s/%s/%s",
        .Platform$pkgType,
        R.Version()$os,
        R.Version()$arch
      )
    )
  }

  install_booster_pack <- function(package, load = TRUE) {
    for (pkg in package) {
      if (!requireNamespace(pkg, quietly = TRUE)) {
        pak::pkg_install(pkg)
      }
      if (load) {
        library(pkg, character.only = TRUE)
      }
    }
  }

  if (file.exists('packages.txt')) {
    packages <- read.table('packages.txt')
    install_booster_pack(package = packages$Package, load = FALSE)
    rm(packages)
  } else {
    booster_pack <- c(
      ### IO ----
      'fs',
      'here',
      'janitor',
      'rio',
      'tidyverse',

      ### EDA ----
      'skimr',

      ### Plot ----
      'ggrepel',
      'ggtext',
      'scales',

      ### Text ----
      'tidytext',

      ### Misc ----
      'tidytuesdayR'
    )

    install_booster_pack(package = booster_pack, load = TRUE)
    rm(install_booster_pack, booster_pack)
  }

  # Custom Functions ----

  `%ni%` <- Negate(`%in%`)

  geometric_mean <- function(x) {
    exp(mean(log(x[x > 0]), na.rm = TRUE))
  }

  my_skim <- skim_with(
    numeric = sfl(
      n = length,
      min = ~ min(.x, na.rm = T),
      p25 = ~ stats::quantile(., probs = .25, na.rm = TRUE, names = FALSE),
      med = ~ median(.x, na.rm = T),
      p75 = ~ stats::quantile(., probs = .75, na.rm = TRUE, names = FALSE),
      max = ~ max(.x, na.rm = T),
      mean = ~ mean(.x, na.rm = T),
      geo_mean = ~ geometric_mean(.x),
      sd = ~ stats::sd(., na.rm = TRUE),
      hist = ~ inline_hist(., 5)
    ),
    append = FALSE
  )
}
```

# Load raw data from package

```{r}
#| label: raw-data-load
#| echo: true
#| output: false

raw <- tidytuesdayR::tt_load('2025-11-18')

holmes <- raw$sherlock

```

# Exploratory Data Analysis

The `my_skim()` function is a modified version of the `skimr::skim()` function that returns the number of missing data points (cells as `NA`) as well as the inverse (e.g.: number of rows that are *not* `NA`), the count, minimum, 25%, median, 75%, max, mean, geometric mean, and standard deviation. It also generates a little ASCII histogram. Neat!

## Sherlock Holmes Text

```{r}
#| label: eda-holmes
#| echo: true
#| output: true

holmes %>%
  filter(!is.na(text), text != "") %>%
  mutate(n_chars = nchar(text)) %>%
  my_skim(.)

```

```{r}
#| label: eda-books
#| echo: true
#| output: true

book_lengths <- holmes %>%
  filter(!is.na(text), text != "") %>%
  count(book, sort = TRUE, name = "n_lines")

book_lengths

```

# Text Analysis

## Tokenization and Word Frequencies

```{r}
#| label: tokenize
#| echo: true
#| output: true

holmes_words <- holmes %>%
  filter(!is.na(text), text != "") %>%
  unnest_tokens(word, text) %>%
  anti_join(stop_words, by = "word")

# Top words across entire canon
holmes_words %>%
  count(word, sort = TRUE) %>%
  head(20)

```

## Sentiment Arc Across Stories

How does sentiment evolve through a typical Holmes story?

```{r}
#| label: sentiment-arc
#| echo: true
#| output: true

# Use Bing lexicon for positive/negative
sentiment_by_line <- holmes %>%
  filter(!is.na(text), text != "") %>%
  group_by(book) %>%
  mutate(line_pct = line_num / max(line_num, na.rm = TRUE)) %>%
  ungroup() %>%
  unnest_tokens(word, text) %>%
  inner_join(get_sentiments("bing"), by = "word") %>%
  mutate(score = ifelse(sentiment == "positive", 1, -1))

# Aggregate by book and story progress (deciles)
arc_data <- sentiment_by_line %>%
  mutate(decile = floor(line_pct * 10) / 10) %>%
  group_by(book, decile) %>%
  summarize(
    net_sentiment = sum(score),
    .groups = "drop"
  )

# Average arc across all books
avg_arc <- arc_data %>%
  group_by(decile) %>%
  summarize(
    mean_sentiment = mean(net_sentiment, na.rm = TRUE),
    .groups = "drop"
  )

avg_arc

```

## Most Distinctive Words by Story

Which words are most unique to each story compared to the rest of the canon?

```{r}
#| label: tf-idf
#| echo: true
#| output: true

book_words <- holmes_words %>%
  count(book, word, sort = TRUE)

book_tfidf <- book_words %>%
  bind_tf_idf(word, book, n) %>%
  arrange(desc(tf_idf))

# Top 3 distinctive words per book (sample of books)
top_books <- book_lengths %>%
  head(6) %>%
  pull(book)

book_tfidf %>%
  filter(book %in% top_books) %>%
  group_by(book) %>%
  slice_max(tf_idf, n = 5) %>%
  select(book, word, tf_idf)

```
## Visualizing the Sentiment Arc of a Holmes Mystery

```{r}
#| label: viz-sentiment-arc
#| message: false
#| warning: false
#| fig.width: 10
#| fig.height: 8

# Victorian-inspired palette
ggplot(avg_arc, aes(x = decile, y = mean_sentiment)) +
  geom_area(fill = "#2C3E50", alpha = 0.3) +
  geom_line(color = "#2C3E50", linewidth = 1.2) +
  geom_point(color = "#C0392B", size = 3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "#777777") +
  annotate(
    "text", x = 0.05, y = max(avg_arc$mean_sentiment) * 0.9,
    label = "More positive", hjust = 0, size = 3.5, color = "#27AE60", fontface = "italic"
  ) +
  annotate(
    "text", x = 0.05, y = min(avg_arc$mean_sentiment) * 0.9,
    label = "More negative", hjust = 0, size = 3.5, color = "#C0392B", fontface = "italic"
  ) +
  scale_x_continuous(
    labels = scales::percent_format(),
    breaks = seq(0, 1, 0.1)
  ) +
  labs(
    title = "The Emotional Arc of a Sherlock Holmes Mystery",
    subtitle = "Average net sentiment (Bing lexicon) across story progression | All stories combined",
    x = "Story Progress",
    y = "Net Sentiment",
    caption = "Source: TidyTuesday 2025-11-18 | {sherlock} R package by Emil Hvitfeldt"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 18, color = "#2C3E50"),
    plot.subtitle = element_text(size = 11, color = "#555555"),
    plot.caption = element_text(size = 9, color = "#888888"),
    panel.grid.minor = element_blank()
  )

```

```{r}
#| label: viz-tfidf
#| message: false
#| warning: false
#| fig.width: 10
#| fig.height: 8

tfidf_plot <- book_tfidf %>%
  filter(book %in% top_books) %>%
  group_by(book) %>%
  slice_max(tf_idf, n = 8) %>%
  ungroup() %>%
  mutate(
    word = reorder_within(word, tf_idf, book)
  )

ggplot(tfidf_plot, aes(x = word, y = tf_idf, fill = book)) +
  geom_col(show.legend = FALSE, width = 0.7) +
  facet_wrap(~ book, scales = "free_y", ncol = 2) +
  scale_x_reordered() +
  scale_fill_manual(values = c(
    "#2C3E50", "#8E44AD", "#C0392B", "#2980B9", "#27AE60", "#D35400"
  )) +
  coord_flip() +
  labs(
    title = "What Makes Each Holmes Story Unique?",
    subtitle = "Most distinctive words per story by TF-IDF score",
    x = NULL,
    y = "TF-IDF",
    caption = "Source: TidyTuesday 2025-11-18 | {sherlock} R package"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 17, color = "#2C3E50"),
    plot.subtitle = element_text(size = 11, color = "#555555"),
    plot.caption = element_text(size = 9, color = "#888888"),
    strip.text = element_text(face = "bold", size = 10),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )

```

# Final thoughts and takeaways

The complete Sherlock Holmes canon — 56 short stories and 4 novels — reveals remarkably consistent patterns when viewed through a computational lens. The average sentiment arc across all stories follows a recognizable shape: stories tend to open with moderate positivity (Watson setting the scene), dip into negativity during the middle act (the mystery deepens, danger emerges), and recover toward resolution.

The TF-IDF analysis surfaces the unique vocabulary fingerprint of each story. Names, locations, and domain-specific terms (poisons, weapons, occupations) define each mystery's identity. This is Conan Doyle's formula: each story inhabits a distinct world even while following the same narrative structure.

::: {.callout-note}
The Bing sentiment lexicon is a blunt instrument for Victorian prose. Words like "grave" and "dark" carry different connotations in 1890s London than in modern usage. A more nuanced analysis might use a period-appropriate lexicon or train a custom sentiment model on 19th-century fiction.
:::
