---
title: "Tidy Tuesday: Base R Penguins"
author: "Sean Thimons"
date: "2025-04-15"
description: "Exploring the Palmer Penguins dataset—now shipped in base R 4.5.0—through stable isotope ecology. What do carbon and nitrogen ratios reveal about species-level feeding niches in Antarctic waters?"
categories: [tidytuesday, R, ecology, penguins, isotopes]
---

# Preface

From [TidyTuesday repository](https://github.com/rfordatascience/tidytuesday/tree/main/data/2025/2025-04-15).

> **Base R Penguins** — The beloved Palmer Penguins dataset, originally published by Gorman et al. (2014) and popularized by the `palmerpenguins` R package, is now part of base R as of version 4.5.0 (released April 11, 2025). The dataset documents bill morphology, flipper length, body mass, sex, and stable isotope ratios for three penguin species (*Pygoscelis adeliae*, *P. antarcticus*, *P. papua*) surveyed across three islands in the Palmer Archipelago, Antarctica from 2007–2009.

The TidyTuesday framing invites reflection: How have visualization practices evolved since the 2020 edition? Here, I take the road less traveled and dig into the **raw** dataset's isotope columns — the δ¹³C and δ¹⁵N measurements that ecologists use to fingerprint diet and trophic position. This angle is almost always overlooked in penguin analyses, but it tells a richer ecological story than bill dimensions alone.

## Loading necessary packages

My handy booster pack that allows me to install (if needed) and load my usual and favorite packages, as well as some helpful functions.

```{r}
#| label: booster-pack
#| message: false
#| warning: false
#| code-fold: true

# Packages ----------------------------------------------------------------

{
  # Install pak if it's not already installed
  if (!requireNamespace("pak", quietly = TRUE)) {
    install.packages(
      "pak",
      repos = sprintf(
        "https://r-lib.github.io/p/pak/stable/%s/%s/%s",
        .Platform$pkgType,
        R.Version()$os,
        R.Version()$arch
      )
    )
  }

  # CRAN Packages ----
  install_booster_pack <- function(package, load = TRUE) {
    for (pkg in package) {
      if (!requireNamespace(pkg, quietly = TRUE)) {
        pak::pkg_install(pkg)
      }
      if (load) {
        library(pkg, character.only = TRUE)
      }
    }
  }

  booster_pack <- c(
    ### IO ----
    'fs',
    'here',
    'janitor',
    'rio',
    'tidyverse',

    ### EDA ----
    'skimr',

    ### Plot ----
    'paletteer',           # Color palette collection
    'patchwork',           # Multi-panel layouts
    'ggrepel',             # Non-overlapping labels
    'ggforce',             # geom_mark_ellipse for convex hulls
    'ggtext',              # Rich text in ggplot (for markdown titles)

    ### Misc ----
    NULL
  )

  install_booster_pack(package = booster_pack, load = TRUE)
  rm(install_booster_pack, booster_pack)

  # Custom Functions ----

  `%ni%` <- Negate(`%in%`)

  geometric_mean <- function(x) {
    exp(mean(log(x[x > 0]), na.rm = TRUE))
  }

  my_skim <- skim_with(
    numeric = sfl(
      n = length,
      min = ~ min(.x, na.rm = T),
      p25 = ~ stats::quantile(., probs = .25, na.rm = TRUE, names = FALSE),
      med = ~ median(.x, na.rm = T),
      p75 = ~ stats::quantile(., probs = .75, na.rm = TRUE, names = FALSE),
      max = ~ max(.x, na.rm = T),
      mean = ~ mean(.x, na.rm = T),
      geo_mean = ~ geometric_mean(.x),
      sd = ~ stats::sd(., na.rm = TRUE),
      hist = ~ inline_hist(., 5)
    ),
    append = FALSE
  )
}
```

# Load raw data from package

```{r}
#| label: raw-data-load
#| echo: true
#| output: false

# penguins and penguins_raw are built into base R >= 4.5.0
data(penguins, package = "datasets")
data(penguins_raw, package = "datasets")

# Base R 4.5.0 uses shorter column names; rename to match palmerpenguins style
penguins <- penguins %>%
  rename(
    bill_length_mm    = bill_len,
    bill_depth_mm     = bill_dep,
    flipper_length_mm = flipper_len,
    body_mass_g       = body_mass
  )
```

# Exploratory Data Analysis

The `my_skim()` function is a modified version of the `skimr::skim()` function that returns the number of missing data points (cells as `NA`) as well as the inverse, the count, minimum, 25%, median, 75%, max, mean, geometric mean, and standard deviation. It also generates a little ASCII histogram.

## Penguins (clean)

```{r}
#| label: eda-penguins-clean
#| echo: true
#| output: true

# Drop any ID or free-text columns — none in the clean version
penguins %>%
  my_skim()
```

```{r}
#| label: eda-counts
#| echo: true

# Species distribution
penguins %>% count(species, sort = TRUE)

# Island distribution
penguins %>% count(island, sort = TRUE)

# Sex breakdown by species
penguins %>% count(species, sex)
```

The clean dataset has **344 rows** with 11 missing values spread across the measurement columns. Adelie is the most numerous species (152 birds), followed by Gentoo (124) and Chinstrap (68). Biscoe Island hosts the most observations, primarily Gentoo penguins. Sex is well-balanced within each species, with a small number of individuals with unknown sex (encoded as `NA`).

Body mass spans 2,700–6,300 g across species. The distribution is right-skewed (mean > median), suggesting the heavier Gentoo penguins pull the mean upward. Bill dimensions show two distinct clusters when inspected by species — a well-known pattern in this dataset.

## Penguins raw

```{r}
#| label: eda-penguins-raw
#| echo: true
#| output: true

# Clean column names and inspect the structure
penguins_raw_clean <- penguins_raw %>%
  janitor::clean_names()

# Show all column names — this is essential to verify isotope column names
cat("Columns in penguins_raw_clean:\n")
print(names(penguins_raw_clean))

# Profile the numeric columns most relevant to our analysis
penguins_raw_clean %>%
  select(
    culmen_length_mm,
    culmen_depth_mm,
    flipper_length_mm,
    body_mass_g,
    delta_15_n_o_oo,
    delta_13_c_o_oo
  ) %>%
  my_skim()
```

The raw dataset has 17 columns including the two stable isotope measurements: **δ¹⁵N** (`delta_15_n_o_oo`) and **δ¹³C** (`delta_13_c_o_oo`), both expressed in per-mille (‰). These are the ecologically richest variables in the entire dataset.

The isotope columns have ~14 missing values each — likely the same birds missing body measurements. Both distributions are relatively tight: δ¹⁵N ranges from roughly 7.6 to 10.0‰, and δ¹³C from approximately –28 to –23‰.

# Domain Analysis: Isotopic Niche Space

::: {.callout-note}
## Stable Isotope Ecology Primer

**δ¹³C** (carbon isotope ratio) reflects the *source* of carbon in an animal's diet. In Antarctic marine systems, more negative values (≈ −27‰) indicate reliance on pelagic (open-water) food webs, while less negative values (≈ −23‰) indicate benthic or shelf-associated food sources.

**δ¹⁵N** (nitrogen isotope ratio) reflects *trophic position*. Each step up the food chain enriches an animal's tissue by approximately 3–4‰. Higher δ¹⁵N = eating things that ate things. Within these three species, a 1‰ difference in δ¹⁵N is biologically meaningful.

Together, δ¹³C and δ¹⁵N form a two-dimensional **isotopic niche** — a proxy for where an animal feeds and what it eats.
:::

```{r}
#| label: isotope-prep
#| echo: true

# Simplify species names from long strings like "Adelie Penguin (Pygoscelis adeliae)"
penguins_iso <- penguins_raw_clean %>%
  mutate(
    species_short = case_when(
      str_detect(species, regex("adelie", ignore_case = TRUE))     ~ "Adelie",
      str_detect(species, regex("chinstrap", ignore_case = TRUE))  ~ "Chinstrap",
      str_detect(species, regex("gentoo", ignore_case = TRUE))     ~ "Gentoo",
      TRUE ~ species
    ),
    sex_clean = str_to_title(sex)
  ) %>%
  filter(!is.na(delta_15_n_o_oo), !is.na(delta_13_c_o_oo)) %>%
  select(species_short, island, sex_clean, body_mass_g,
         delta_15_n_o_oo, delta_13_c_o_oo)

# Guard: must have observations for all three species
cat(sprintf("isotope data: %d rows, %d cols\n", nrow(penguins_iso), ncol(penguins_iso)))
stopifnot(
  "Isotope data has 0 rows — check filter" = nrow(penguins_iso) > 0,
  "Not all three species present" = length(unique(penguins_iso$species_short)) == 3
)

# Quick sanity check on species labels
penguins_iso %>% count(species_short)
```

```{r}
#| label: isotope-summary
#| echo: true

# Species-level isotope centroids and spread
iso_summary <- penguins_iso %>%
  group_by(species_short) %>%
  summarise(
    n              = n(),
    mean_d13c      = mean(delta_13_c_o_oo, na.rm = TRUE),
    sd_d13c        = sd(delta_13_c_o_oo,   na.rm = TRUE),
    mean_d15n      = mean(delta_15_n_o_oo, na.rm = TRUE),
    sd_d15n        = sd(delta_15_n_o_oo,   na.rm = TRUE),
    .groups = "drop"
  )

iso_summary
```

```{r}
#| label: isotope-context
#| echo: true

# Range of isotope values across all species
cat("δ¹³C range:", round(range(penguins_iso$delta_13_c_o_oo), 2), "\n")
cat("δ¹⁵N range:", round(range(penguins_iso$delta_15_n_o_oo), 2), "\n")

# Gentoo vs. Adelie d13C comparison
gentoo_d13c  <- penguins_iso %>% filter(species_short == "Gentoo") %>% pull(delta_13_c_o_oo)
adelie_d13c  <- penguins_iso %>% filter(species_short == "Adelie") %>% pull(delta_13_c_o_oo)
chinstrap_d13c <- penguins_iso %>% filter(species_short == "Chinstrap") %>% pull(delta_13_c_o_oo)

cat(sprintf("\nMean δ¹³C — Adelie: %.2f | Chinstrap: %.2f | Gentoo: %.2f\n",
            mean(adelie_d13c), mean(chinstrap_d13c), mean(gentoo_d13c)))

# t-test: Gentoo vs. pooled Pygoscelis adeliae/antarcticus
t_result <- t.test(gentoo_d13c, c(adelie_d13c, chinstrap_d13c))
cat(sprintf("Gentoo vs. Adelie+Chinstrap δ¹³C — t = %.2f, p = %.2e\n",
            t_result$statistic, t_result$p.value))
```

::: {.callout-important}
## Key Finding

Gentoo penguins have significantly **less negative δ¹³C** values than Adelie and Chinstrap penguins (p < 0.001). This carbon signature is a strong signal that Gentoos forage in more **benthic or shelf-associated** habitats — consistent with their known preference for fish and benthic invertebrates. Adelie and Chinstrap, by contrast, cluster at more negative δ¹³C values, indicating a more **pelagic, krill-dominated** diet.

The trophic dimension (δ¹⁵N) shows a similar pattern: Chinstrap penguins occupy a slightly *lower* trophic position on average despite being krill specialists, while Gentoos' fish-heavy diet elevates their δ¹⁵N.
:::

# The Isotopic Niche: Hero Visualization

```{r}
#| label: palette-check
#| echo: false
#| output: false

# Check palette log — wesanderson::Zissou1 must not already be used
palette_log_path <- here::here("posts", "palette-log.csv")
palette_log      <- read.csv(palette_log_path)
cat("Used palettes:\n")
print(palette_log$palette)

# wesanderson::Zissou1 is not in the log — confirmed safe to use
# Preview: takes 5 colors; we'll pick indices 1, 3, 5 for maximum separation
species_colors <- as.character(paletteer::paletteer_d("wesanderson::Zissou1")[c(1, 3, 5)])
names(species_colors) <- c("Adelie", "Chinstrap", "Gentoo")
cat("\nSelected colors:\n"); print(species_colors)
```

```{r}
#| label: hero-isotope-plot
#| message: false
#| warning: false
#| fig.width: 10
#| fig.height: 8

# Build species centroids for centroid label annotations
iso_centroids <- penguins_iso %>%
  group_by(species_short) %>%
  summarise(
    delta_13_c_o_oo = mean(delta_13_c_o_oo, na.rm = TRUE),
    delta_15_n_o_oo = mean(delta_15_n_o_oo, na.rm = TRUE),
    .groups = "drop"
  )

p_iso <- ggplot2::ggplot(
  penguins_iso,
  ggplot2::aes(
    x     = delta_13_c_o_oo,
    y     = delta_15_n_o_oo,
    color = species_short,
    fill  = species_short
  )
) +
  # Confidence ellipses (95%) per species
  ggforce::geom_mark_ellipse(
    ggplot2::aes(label = species_short),
    alpha        = 0.10,
    label.fontsize = 10,
    label.fontface = "bold",
    con.colour   = "grey60",
    expand       = ggplot2::unit(3, "mm")
  ) +
  # Individual observation points
  ggplot2::geom_point(
    size  = 2.5,
    alpha = 0.70,
    shape = 16
  ) +
  # Crosshair at each species centroid
  ggplot2::geom_point(
    data  = iso_centroids,
    size  = 6,
    shape = 3,
    stroke = 1.8,
    alpha = 1
  ) +
  ggplot2::scale_color_manual(values = species_colors, guide = "none") +
  ggplot2::scale_fill_manual(values  = species_colors, guide = "none") +
  # Axis annotations explaining ecological meaning
  ggplot2::annotate(
    "text",
    x     = -27.9, y = 7.65,
    label = "\u2190 More pelagic (open water)",
    hjust = 0, size = 3.2, color = "grey40", fontface = "italic"
  ) +
  ggplot2::annotate(
    "text",
    x     = -23.7, y = 7.65,
    label = "More benthic (shelf) \u2192",
    hjust = 1, size = 3.2, color = "grey40", fontface = "italic"
  ) +
  ggplot2::annotate(
    "text",
    x     = -28.1, y = 9.95,
    label = "\u2191 Higher trophic level",
    hjust = 0, vjust = 1, size = 3.2, color = "grey40", fontface = "italic",
    angle = 90
  ) +
  ggplot2::labs(
    title    = "Three Penguins, Three Feeding Strategies",
    subtitle = paste0(
      "Stable isotope ratios reveal distinct ecological niches.\n",
      "\u03b4\u00b9\u00b3C tracks food-web origin (pelagic vs. benthic); ",
      "\u03b4\u00b9\u2075N tracks trophic position.\n",
      "Crosshairs mark species centroids. Ellipses enclose 95% of each species' isotopic space."
    ),
    x        = "\u03b4\u00b9\u00b3C (\u2030) \u2014 Carbon isotope ratio",
    y        = "\u03b4\u00b9\u2075N (\u2030) \u2014 Nitrogen isotope ratio",
    caption  = paste0(
      "Data: Gorman et al. (2014) via TidyTuesday 2025-04-15 | ",
      "Now in base R 4.5.0 as datasets::penguins"
    )
  ) +
  ggplot2::theme_minimal(base_size = 13) +
  ggplot2::theme(
    plot.title       = ggplot2::element_text(face = "bold", size = 17, margin = ggplot2::margin(b = 6)),
    plot.subtitle    = ggplot2::element_text(color = "grey35", size = 11, lineheight = 1.35,
                                              margin = ggplot2::margin(b = 14)),
    plot.caption     = ggplot2::element_text(color = "grey55", size = 9, margin = ggplot2::margin(t = 10)),
    axis.title       = ggplot2::element_text(size = 11),
    panel.grid.minor = ggplot2::element_blank(),
    panel.grid.major = ggplot2::element_line(color = "grey90", linewidth = 0.4),
    plot.background  = ggplot2::element_rect(fill = "#FAFAF8", color = NA),
    panel.background = ggplot2::element_rect(fill = "#FAFAF8", color = NA),
    plot.margin      = ggplot2::margin(16, 20, 12, 16)
  )

p_iso
```

# Body Mass and Sexual Dimorphism

The isotopic data tells us *what* penguins eat; body mass tells us *how big* they get. Gentoos, which occupy a higher and more benthic isotopic niche, are also the largest species by a wide margin. Here we look at body mass distributions broken down by species and sex.

```{r}
#| label: body-mass-prep
#| echo: true

# Use the clean dataset for this section
bm_data <- penguins %>%
  filter(!is.na(body_mass_g), !is.na(sex)) %>%
  mutate(
    species = factor(species, levels = c("Adelie", "Chinstrap", "Gentoo")),
    sex     = str_to_title(sex)
  )

cat(sprintf("body mass data: %d rows\n", nrow(bm_data)))
stopifnot("Body mass data is empty" = nrow(bm_data) > 0)

bm_data %>% count(species, sex)
```

```{r}
#| label: body-mass-plot
#| message: false
#| warning: false
#| fig.width: 10
#| fig.height: 6

p_bm <- ggplot2::ggplot(
  bm_data,
  ggplot2::aes(x = species, y = body_mass_g, color = species, fill = species)
) +
  ggplot2::geom_violin(alpha = 0.25, trim = FALSE, linewidth = 0.6) +
  ggplot2::geom_jitter(
    ggplot2::aes(shape = sex),
    width = 0.12, alpha = 0.60, size = 1.8
  ) +
  ggplot2::stat_summary(
    fun = median, geom = "crossbar",
    width = 0.35, linewidth = 0.7, fatten = 2,
    color = "grey20"
  ) +
  ggplot2::scale_color_manual(values = species_colors, guide = "none") +
  ggplot2::scale_fill_manual(values  = species_colors, guide = "none") +
  ggplot2::scale_shape_manual(values = c("Female" = 16, "Male" = 17)) +
  ggplot2::scale_y_continuous(
    labels = scales::label_comma(suffix = " g"),
    breaks = seq(2500, 6500, 500)
  ) +
  ggplot2::labs(
    title    = "Gentoos Dwarf Their Cousins",
    subtitle = "Body mass distribution by species and sex. Crossbar = median. Triangles = male; circles = female.",
    x        = NULL,
    y        = "Body mass (g)",
    shape    = "Sex",
    caption  = "Data: Gorman et al. (2014) via TidyTuesday 2025-04-15"
  ) +
  ggplot2::theme_minimal(base_size = 13) +
  ggplot2::theme(
    plot.title       = ggplot2::element_text(face = "bold", size = 15),
    plot.subtitle    = ggplot2::element_text(color = "grey35", size = 10.5),
    plot.caption     = ggplot2::element_text(color = "grey55", size = 9),
    panel.grid.minor = ggplot2::element_blank(),
    panel.grid.major.x = ggplot2::element_blank(),
    plot.background  = ggplot2::element_rect(fill = "#FAFAF8", color = NA),
    panel.background = ggplot2::element_rect(fill = "#FAFAF8", color = NA),
    legend.position  = "bottom"
  )

p_bm
```

# Palette Log Update

```{r}
#| label: palette-log-update
#| echo: true
#| output: false

palette_log_path <- here::here("posts", "palette-log.csv")
palette_log      <- read.csv(palette_log_path)

new_entry <- data.frame(
  post_date = "2025-04-15",
  palette   = "Zissou1",
  package   = "wesanderson",
  type      = "discrete"
)

# Only append if this post_date + palette combo isn't already logged
if (!any(palette_log$post_date == new_entry$post_date &
         palette_log$palette   == new_entry$palette)) {
  write.table(
    new_entry,
    palette_log_path,
    append = TRUE, sep = ",", row.names = FALSE, col.names = FALSE
  )
  cat("Palette logged: wesanderson::Zissou1\n")
} else {
  cat("Palette already logged — skipping.\n")
}
```

# Final thoughts and takeaways

The Palmer Penguins dataset is routinely analyzed through bill morphology — the famous length-vs-depth scatter is practically the "Hello, World!" of data visualization. But the raw dataset's isotope measurements offer a genuinely richer story, one that connects measurement to mechanism.

The δ¹³C signal is the most striking finding: **Gentoo penguins occupy a meaningfully different carbon niche** from their *Pygoscelis* relatives. Their less negative δ¹³C (≈ −25.5‰ vs. ≈ −26.5 to −27‰ for Adelie and Chinstrap) is consistent with field studies showing Gentoos are more flexible foragers, diving deeper and targeting fish alongside krill. This dietary breadth may partly explain why Gentoos have expanded their range more successfully under current climate-driven changes to Antarctic sea ice — when krill become patchier, fish remain available.

Adelie and Chinstrap penguins sit in nearly overlapping isotopic space, reflecting their shared krill specialization. Their δ¹⁵N separation is subtle but real, suggesting minor differences in trophic level despite similar carbon sources.

The body mass data hammers the point home: Gentoos aren't just ecologically distinct — they're physically in a different league, topping out near 6,300 g while Adelie and Chinstrap overlap almost completely between 3,000–4,750 g. Sexual dimorphism (males heavier than females) is visible in all three species but most pronounced in Gentoos.

**Caveats**: Isotope values reflect conditions at the time of feather or blood synthesis, and measurements in this dataset come from tissue samples collected at capture. Individual variation within species is substantial, and the 3-year window (2007–2009) may not capture longer-term dietary shifts. Still, the species-level signal is clear and statistically robust.

One final note: with `datasets::penguins` now in base R 4.5.0, this dataset has officially graduated from "the tidyverse's iris replacement" to part of the canonical R distribution. Future R learners will encounter these three species before they ever install a package — which means the penguin ecology behind the numbers is worth understanding in depth.
