---
title: "Tidy Tuesday: World Food Day"
author: "Sean Thimons"
date: "2025-10-14"
description: "Marking the FAO's 80th anniversary with a deep dive into global food security indicators — tracking undernourishment, dietary adequacy, and food affordability across regions and time."
categories: [tidytuesday, R, food-security, global-health]
---

# Preface

From [TidyTuesday repository](https://github.com/rfordatascience/tidytuesday/blob/main/data/2025/2025-10-14/readme.md).

> This dataset contains the FAO's Suite of Food Security Indicators, compiled to commemorate the 80th anniversary of the Food and Agriculture Organization of the United Nations. The data captures various aspects of food insecurity through indicators selected by expert judgment, covering observations from 2005 onward.
>
> -   Which indicators tend to vary together?
> -   Do any indicators appear to be leading indicators of others?
> -   How have confidence intervals changed over time or by region?

## Loading necessary packages

My handy booster pack that allows me to install (if needed) and load my usual and favorite packages, as well as some helpful functions.

```{r}
#| label: booster-pack
#| message: false
#| warning: false
#| code-fold: true

# Packages ----------------------------------------------------------------

{
  if (!requireNamespace("pak", quietly = TRUE)) {
    install.packages(
      "pak",
      repos = sprintf(
        "https://r-lib.github.io/p/pak/stable/%s/%s/%s",
        .Platform$pkgType,
        R.Version()$os,
        R.Version()$arch
      )
    )
  }

  install_booster_pack <- function(package, load = TRUE) {
    for (pkg in package) {
      if (!requireNamespace(pkg, quietly = TRUE)) {
        pak::pkg_install(pkg)
      }
      if (load) {
        library(pkg, character.only = TRUE)
      }
    }
  }

  if (file.exists('packages.txt')) {
    packages <- read.table('packages.txt')
    install_booster_pack(package = packages$Package, load = FALSE)
    rm(packages)
  } else {
    booster_pack <- c(
      ### IO ----
      'fs',
      'here',
      'janitor',
      'rio',
      'tidyverse',

      ### EDA ----
      'skimr',

      ### Plot ----
      'ggrepel',
      'ggtext',
      'scales',

      ### Misc ----
      'tidytuesdayR'
    )

    install_booster_pack(package = booster_pack, load = TRUE)
    rm(install_booster_pack, booster_pack)
  }

  # Custom Functions ----

  `%ni%` <- Negate(`%in%`)

  geometric_mean <- function(x) {
    exp(mean(log(x[x > 0]), na.rm = TRUE))
  }

  my_skim <- skim_with(
    numeric = sfl(
      n = length,
      min = ~ min(.x, na.rm = T),
      p25 = ~ stats::quantile(., probs = .25, na.rm = TRUE, names = FALSE),
      med = ~ median(.x, na.rm = T),
      p75 = ~ stats::quantile(., probs = .75, na.rm = TRUE, names = FALSE),
      max = ~ max(.x, na.rm = T),
      mean = ~ mean(.x, na.rm = T),
      geo_mean = ~ geometric_mean(.x),
      sd = ~ stats::sd(., na.rm = TRUE),
      hist = ~ inline_hist(., 5)
    ),
    append = FALSE
  )
}
```

# Load raw data from package

```{r}
#| label: raw-data-load
#| echo: true
#| output: false

raw <- tidytuesdayR::tt_load('2025-10-14')

food_security <- raw$food_security

```

# Exploratory Data Analysis

The `my_skim()` function is a modified version of the `skimr::skim()` function that returns the number of missing data points (cells as `NA`) as well as the inverse (e.g.: number of rows that are *not* `NA`), the count, minimum, 25%, median, 75%, max, mean, geometric mean, and standard deviation. It also generates a little ASCII histogram. Neat!

## Food Security Indicators

```{r}
#| label: eda-food
#| echo: true
#| output: true

food_security %>%
  my_skim(.)

```

```{r}
#| label: eda-overview
#| echo: true
#| output: true

food_security %>%
  count(Item, sort = TRUE) %>%
  head(15)

food_security %>%
  count(Unit, sort = TRUE)

food_security %>%
  count(Flag, sort = TRUE)

# How many countries?
n_distinct(food_security$Area)

```

# Undernourishment and Food Affordability

## Prevalence of Undernourishment Over Time

```{r}
#| label: undernourishment-trend
#| echo: true
#| output: true

undernourishment <- food_security %>%
  filter(str_detect(Item, regex("prevalence of undernourishment", ignore_case = TRUE))) %>%
  filter(!is.na(Value))

# Global regions (filter out individual countries for trend)
global_regions <- c("World", "Africa", "Asia", "Latin America and the Caribbean",
                    "Northern America and Europe", "Oceania")

under_regions <- undernourishment %>%
  filter(Area %in% global_regions)

under_regions %>%
  filter(Year_Start == max(Year_Start)) %>%
  arrange(desc(Value))

```

## Dietary Energy Supply Adequacy

```{r}
#| label: dietary-adequacy
#| echo: true
#| output: true

dietary_adequacy <- food_security %>%
  filter(str_detect(Item, regex("dietary energy supply adequacy", ignore_case = TRUE))) %>%
  filter(!is.na(Value), Area %in% global_regions)

dietary_adequacy %>%
  filter(Year_Start == max(Year_Start)) %>%
  arrange(Value)

```

## Countries with Highest Food Insecurity

```{r}
#| label: worst-countries
#| echo: true
#| output: true

latest_under <- undernourishment %>%
  filter(Area %ni% global_regions) %>%
  group_by(Area) %>%
  slice_max(Year_Start, n = 1) %>%
  ungroup() %>%
  arrange(desc(Value)) %>%
  head(15)

latest_under %>%
  select(Area, Year_Start, Year_End, Value, CI_Lower, CI_Upper)

```
## Visualizing Global Food Security

```{r}
#| label: viz-undernourishment
#| message: false
#| warning: false
#| fig.width: 10
#| fig.height: 8

# Earth-tone palette for food/agriculture
region_cols <- c(
  "World"                               = "#1B1B1B",
  "Africa"                              = "#9B2226",
  "Asia"                                = "#CA6702",
  "Latin America and the Caribbean"     = "#588157",
  "Northern America and Europe"         = "#005F73",
  "Oceania"                             = "#AE2012"
)

ggplot(under_regions, aes(x = Year_Start, y = Value, color = Area)) +
  geom_line(linewidth = 1.1) +
  geom_point(
    data = under_regions %>%
      group_by(Area) %>%
      slice_max(Year_Start, n = 1),
    size = 2.5
  ) +
  geom_ribbon(
    aes(ymin = CI_Lower, ymax = CI_Upper, fill = Area),
    alpha = 0.1,
    color = NA
  ) +
  geom_text_repel(
    data = under_regions %>%
      group_by(Area) %>%
      slice_max(Year_Start, n = 1),
    aes(label = Area),
    nudge_x = 1,
    size = 3.3,
    direction = "y",
    segment.color = "#BBBBBB"
  ) +
  scale_color_manual(values = region_cols) +
  scale_fill_manual(values = region_cols) +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  labs(
    title = "The Uneven Fight Against Hunger",
    subtitle = "Prevalence of undernourishment by world region | Shading = confidence interval",
    x = "Year",
    y = "Prevalence of Undernourishment (%)",
    caption = "Source: TidyTuesday 2025-10-14 | FAO Suite of Food Security Indicators"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 18, color = "#1B1B1B"),
    plot.subtitle = element_text(size = 11, color = "#555555"),
    plot.caption = element_text(size = 9, color = "#888888"),
    legend.position = "none",
    panel.grid.minor = element_blank()
  )

```

```{r}
#| label: viz-worst-countries
#| message: false
#| warning: false
#| fig.width: 10
#| fig.height: 7

latest_under_plot <- latest_under %>%
  mutate(Area = fct_reorder(Area, Value))

ggplot(latest_under_plot, aes(x = Area, y = Value)) +
  geom_col(fill = "#9B2226", width = 0.7) +
  geom_errorbar(
    aes(ymin = CI_Lower, ymax = CI_Upper),
    width = 0.2,
    color = "#333333"
  ) +
  geom_text(
    aes(label = paste0(round(Value, 1), "%")),
    hjust = -0.2,
    size = 3.5,
    color = "#333333"
  ) +
  scale_y_continuous(
    labels = scales::percent_format(scale = 1),
    expand = expansion(mult = c(0, 0.15))
  ) +
  coord_flip() +
  labs(
    title = "Countries with Highest Undernourishment",
    subtitle = "Most recent available data | Error bars = confidence interval",
    x = NULL,
    y = "Prevalence of Undernourishment (%)",
    caption = "Source: TidyTuesday 2025-10-14 | FAO"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 17, color = "#1B1B1B"),
    plot.subtitle = element_text(size = 11, color = "#555555"),
    plot.caption = element_text(size = 9, color = "#888888"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )

```

# Final thoughts and takeaways

The FAO's food security indicators paint a picture of progress interrupted. Global undernourishment declined steadily from the mid-2000s through 2015, but has since stalled or reversed in several regions — driven by conflict, climate shocks, and economic disruption (including the COVID-19 pandemic).

Africa remains the most food-insecure continent by a wide margin, with undernourishment prevalence far exceeding other regions. Asia has made remarkable progress in reducing hunger, but still carries the largest absolute number of undernourished people due to population size. The confidence intervals on these estimates are wide, reminding us that measuring hunger at a population level is inherently imprecise.

::: {.callout-important}
Food security is not just about calories. The FAO framework captures multiple dimensions — availability, access, utilization, and stability. A country can have adequate caloric supply on paper while still having significant populations who can't afford a nutritious diet. The "cost of a healthy diet" indicator is increasingly seen as a more meaningful measure than caloric adequacy alone.
:::
