---
title: "Tidy Tuesday: API Specs"
author: "Sean Thimons"
date: "2025-06-17"
description: "Mapping the web API ecosystem through APIs.guru's catalog — who provides the most, what categories dominate, and how modern is the spec landscape?"
categories: [tidytuesday, R, apis, web, technology, openapi]
---

# Preface

From the [TidyTuesday repository](https://github.com/rfordatascience/tidytuesday/blob/main/data/2025/2025-06-17/readme.md).

> This week we're exploring Web APIs! The dataset was curated by Jon Harmon while developing tools for creating API-wrapping R packages. The data comes from [APIs.guru](https://apis.guru), whose goal is to create a "machine-readable Wikipedia for Web APIs in the OpenAPI Specification format." Five tables cover API metadata, categories, logos, origin formats, and core specification details.

Suggested questions from the repo:

- What API specs are provided by APIs.guru? Are these the same as the origin specs?
- How many different APIs ("services") do providers provide?
- What licenses do APIs use?
- Are any APIs listed more than once in the dataset?

## Loading necessary packages

My handy booster pack that allows me to install (if needed) and load my usual and favorite packages, as well as some helpful functions.

```{r}
#| label: booster-pack
#| message: false
#| warning: false
#| code-fold: true

# Packages ----------------------------------------------------------------

{
  # Install pak if it's not already installed
  if (!requireNamespace("pak", quietly = TRUE)) {
    install.packages(
      "pak",
      repos = sprintf(
        "https://r-lib.github.io/p/pak/stable/%s/%s/%s",
        .Platform$pkgType,
        R.Version()$os,
        R.Version()$arch
      )
    )
  }

  # CRAN Packages ----
  install_booster_pack <- function(package, load = TRUE) {
    for (pkg in package) {
      if (!requireNamespace(pkg, quietly = TRUE)) {
        pak::pkg_install(pkg)
      }
      if (load) {
        library(pkg, character.only = TRUE)
      }
    }
  }

  booster_pack <- c(
    ### IO ----
    'fs',
    'here',
    'janitor',
    'rio',
    'tidyverse',

    ### EDA ----
    'skimr',

    ### Plot ----
    'paletteer',           # Color palette collection
    'patchwork',           # Multi-panel layouts
    'ggtext',              # Rich text in ggplot
    'ggrepel',             # Non-overlapping labels

    ### Misc ----
    'tidytuesdayR'
  )

  install_booster_pack(package = booster_pack, load = TRUE)
  rm(install_booster_pack, booster_pack)

  # Custom Functions ----

  `%ni%` <- Negate(`%in%`)

  geometric_mean <- function(x) {
    exp(mean(log(x[x > 0]), na.rm = TRUE))
  }

  my_skim <- skim_with(
    numeric = sfl(
      n = length,
      min = ~ min(.x, na.rm = T),
      p25 = ~ stats::quantile(., probs = .25, na.rm = TRUE, names = FALSE),
      med = ~ median(.x, na.rm = T),
      p75 = ~ stats::quantile(., probs = .75, na.rm = TRUE, names = FALSE),
      max = ~ max(.x, na.rm = T),
      mean = ~ mean(.x, na.rm = T),
      geo_mean = ~ geometric_mean(.x),
      sd = ~ stats::sd(., na.rm = TRUE),
      hist = ~ inline_hist(., 5)
    ),
    append = FALSE
  )
}
```

# Load raw data from package

```{r}
#| label: raw-data-load
#| echo: true
#| output: false

raw <- tidytuesdayR::tt_load('2025-06-17')

apisguru_apis <- raw$apisguru_apis %>% janitor::clean_names()
api_categories <- raw$api_categories %>% janitor::clean_names()
api_info       <- raw$api_info       %>% janitor::clean_names()
api_logos      <- raw$api_logos      %>% janitor::clean_names()
api_origins    <- raw$api_origins    %>% janitor::clean_names()
```

# Exploratory Data Analysis

The `my_skim()` function returns count, min, percentiles, mean, geometric mean, standard deviation, and an ASCII histogram.

## `apisguru_apis` — the core catalog

This table has one row per API (filtered to the preferred version only), with timing metadata and spec version info. I drop `swagger_url`, `link`, `external_docs_url`, and `external_docs_description` since they are mostly URLs and free-text that won't add to the numeric profile.

```{r}
#| label: eda-apisguru-apis
#| echo: true
#| output: true

apisguru_apis %>%
  select(-swagger_url, -link, -external_docs_url, -external_docs_description) %>%
  my_skim()
```

## `api_info` — provider, title, and license metadata

This table carries the richest semantic information. I drop free-text columns (`description`, `contact_url`, `license_url`, `terms_of_service`) that won't contribute to the statistical profile, keeping the categorically interesting fields.

```{r}
#| label: eda-api-info
#| echo: true
#| output: true

api_info %>%
  select(name, contact_name, title, provider_name, service_name,
         license_name) %>%
  my_skim()
```

## `api_categories` — many-to-many category assignments

```{r}
#| label: eda-api-categories
#| echo: true
#| output: true

cat(sprintf("api_categories: %d rows, %d cols\n",
            nrow(api_categories), ncol(api_categories)))

api_categories %>%
  count(apisguru_category, sort = TRUE) %>%
  head(20)
```

## `api_origins` — source spec formats

```{r}
#| label: eda-api-origins
#| echo: true
#| output: true

api_origins %>%
  count(format, sort = TRUE)
```

The EDA reveals several things at a glance. The `apisguru_apis` catalog spans thousands of APIs with `added` dates spread over many years. The `api_info` table confirms that `provider_name` is the key grouping variable — providers supply anywhere from one to hundreds of services. Licenses are sparsely populated (many NAs), suggesting the API ecosystem is not yet mature in its licensing practices. In `api_origins`, **swagger** dominates as the original spec format, followed by **openapi** and **google** (Google Discovery).

# The Web API Ecosystem

## Provider concentration: the long tail of API supply

How many APIs does each provider offer? The hypothesis is a classic **power-law distribution**: a handful of mega-providers (Amazon, Google, Microsoft) supply hundreds of APIs, while the vast majority of providers supply just one.

```{r}
#| label: provider-concentration
#| echo: true
#| output: true

provider_counts <- api_info %>%
  filter(!is.na(provider_name)) %>%
  count(provider_name, sort = TRUE, name = "n_apis")

cat(sprintf("provider_counts: %d rows, %d cols\n",
            nrow(provider_counts), ncol(provider_counts)))
stopifnot("provider_counts is empty" = nrow(provider_counts) > 0)

# Distribution summary
provider_counts %>%
  summarise(
    total_providers = n(),
    single_api      = sum(n_apis == 1),
    pct_single      = round(100 * mean(n_apis == 1), 1),
    median_apis     = median(n_apis),
    max_apis        = max(n_apis),
    top_provider    = provider_name[which.max(n_apis)]
  )
```

```{r}
#| label: top-providers-table
#| echo: true
#| output: true

provider_counts %>% head(20)
```

::: {.callout-note}
The concentration is striking. The top provider alone offers tens to hundreds of API versions, while the majority of providers in the catalog contribute just a single spec. This is a textbook Pareto distribution — the "API ecosystem" looks like a busy marketplace only from the outside.
:::

## Category landscape

APIs.guru categorizes APIs with a many-to-many mapping — one API can belong to multiple categories. The counts below reflect total category assignments, not unique APIs.

```{r}
#| label: category-landscape
#| echo: true
#| output: true

category_counts <- api_categories %>%
  filter(!is.na(apisguru_category)) %>%
  count(apisguru_category, sort = TRUE, name = "n_apis")

cat(sprintf("category_counts: %d rows, %d cols\n",
            nrow(category_counts), ncol(category_counts)))
stopifnot("category_counts is empty" = nrow(category_counts) > 0)

category_counts %>% head(20)
```

## Licensing ecosystem

```{r}
#| label: license-landscape
#| echo: true
#| output: true

license_counts <- api_info %>%
  mutate(license_name = if_else(is.na(license_name), "(none / unlisted)", license_name)) %>%
  count(license_name, sort = TRUE, name = "n_apis")

license_counts %>% head(15)
```

::: {.callout-important}
The large proportion of APIs without a stated license reflects a recurring gap in the API ecosystem — many providers publish specs without explicit licensing terms, creating ambiguity for developers building on top of them.
:::

## Spec format: origin vs. APIs.guru's served version

APIs.guru converts all ingested specs to OpenAPI format. This means an API originally written in Postman, WADL, or Google Discovery gets normalized and served in a common format. The `api_origins` table captures the original format, while `apisguru_apis$openapi_ver` reflects what APIs.guru actually serves.

```{r}
#| label: spec-format-analysis
#| echo: true
#| output: true

# Original spec formats
origin_formats <- api_origins %>%
  filter(!is.na(format)) %>%
  count(format, sort = TRUE, name = "n_specs") %>%
  mutate(pct = round(100 * n_specs / sum(n_specs), 1))

cat("Origin formats:\n")
print(origin_formats)

# What APIs.guru serves (openapi version)
served_versions <- apisguru_apis %>%
  filter(!is.na(openapi_ver)) %>%
  mutate(
    major_ver = case_when(
      str_starts(openapi_ver, "3.1") ~ "OpenAPI 3.1.x",
      str_starts(openapi_ver, "3.0") ~ "OpenAPI 3.0.x",
      str_starts(openapi_ver, "2")   ~ "Swagger 2.0",
      TRUE                           ~ paste("Other:", openapi_ver)
    )
  ) %>%
  count(major_ver, sort = TRUE, name = "n_specs") %>%
  mutate(pct = round(100 * n_specs / sum(n_specs), 1))

cat("\nAPIs.guru served versions (grouped):\n")
print(served_versions)
```

::: {.callout-tip}
**Swagger → OpenAPI**: The original format breakdown shows swagger-dominated origins, but APIs.guru also serves swagger 2.0 specs directly. The migration to OpenAPI 3.x is underway but not yet the majority — the ecosystem carries significant legacy.
:::

# Hero visualization: The Web API Ecosystem at a Glance

```{r}
#| label: hero-viz
#| message: false
#| warning: false
#| fig.width: 14
#| fig.height: 8

# --- palette --------------------------------------------------------------
# rcartocolor::Bold — 12 distinct colors, not previously used
cat_palette <- as.character(paletteer::paletteer_d("rcartocolor::Bold"))
provider_accent <- cat_palette[1]  # single accent for provider bars

# --- panel 1: top 20 providers -------------------------------------------
top20_providers <- provider_counts %>%
  head(20) %>%
  mutate(provider_name = fct_reorder(provider_name, n_apis))

cat(sprintf("top20_providers: %d rows\n", nrow(top20_providers)))
stopifnot("top20_providers is empty" = nrow(top20_providers) > 0)

p1 <- ggplot(top20_providers,
             aes(x = n_apis, y = provider_name)) +
  geom_segment(aes(x = 0, xend = n_apis,
                   y = provider_name, yend = provider_name),
               color = "grey80", linewidth = 0.6) +
  geom_point(size = 4, color = provider_accent) +
  geom_text(aes(label = n_apis),
            hjust = -0.4, size = 2.8, color = "grey30", family = "sans") +
  scale_x_continuous(expand = expansion(mult = c(0.01, 0.2))) +
  labs(
    title    = "Top 20 API Providers",
    subtitle = "Number of API specs catalogued on APIs.guru\n(preferred versions only)",
    x        = "Number of APIs",
    y        = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title       = element_text(face = "bold", size = 13),
    plot.subtitle    = element_text(color = "grey50", size = 9),
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    axis.text.y      = element_text(size = 9),
    axis.text.x      = element_text(size = 8)
  )

# --- panel 2: top 10 categories ------------------------------------------
top10_cats <- category_counts %>%
  head(10) %>%
  mutate(
    apisguru_category = fct_reorder(apisguru_category, n_apis),
    fill_color        = cat_palette[seq_len(n())]
  )

cat(sprintf("top10_cats: %d rows\n", nrow(top10_cats)))
stopifnot("top10_cats is empty" = nrow(top10_cats) > 0)

p2 <- ggplot(top10_cats,
             aes(x = n_apis, y = apisguru_category, fill = apisguru_category)) +
  geom_col(width = 0.7, show.legend = FALSE) +
  geom_text(aes(label = n_apis),
            hjust = -0.3, size = 2.8, color = "grey30", family = "sans") +
  scale_fill_manual(values = setNames(top10_cats$fill_color,
                                      top10_cats$apisguru_category)) +
  scale_x_continuous(expand = expansion(mult = c(0.01, 0.2))) +
  labs(
    title    = "Top 10 API Categories",
    subtitle = "Total category assignments\n(one API may appear in multiple categories)",
    x        = "Number of category assignments",
    y        = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title       = element_text(face = "bold", size = 13),
    plot.subtitle    = element_text(color = "grey50", size = 9),
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    axis.text.y      = element_text(size = 9),
    axis.text.x      = element_text(size = 8)
  )

# --- combine with patchwork ----------------------------------------------
p <- p1 + p2 +
  plot_annotation(
    title    = "The Web API Ecosystem",
    subtitle = "Provider concentration and category landscape from APIs.guru's machine-readable API catalog",
    caption  = "Source: APIs.guru via TidyTuesday (2025-06-17) · Visualization: seanthimons.github.io",
    theme    = theme(
      plot.title    = element_text(face = "bold", size = 18, margin = margin(b = 4)),
      plot.subtitle = element_text(color = "grey45", size = 11, margin = margin(b = 12)),
      plot.caption  = element_text(color = "grey60", size = 8, hjust = 1),
      plot.background = element_rect(fill = "white", color = NA)
    )
  )

p
```

```{r}
#| label: palette-log-update
#| echo: false
#| output: false

palette_log_path <- here::here("posts", "palette-log.csv")
palette_log      <- read.csv(palette_log_path)

new_entry <- data.frame(
  post_date = "2025-06-17",
  palette   = "Bold",
  package   = "rcartocolor",
  type      = "discrete"
)

if (!any(palette_log$post_date == new_entry$post_date &
         palette_log$palette   == new_entry$palette)) {
  write.table(
    new_entry,
    palette_log_path,
    append    = TRUE,
    sep       = ",",
    row.names = FALSE,
    col.names = FALSE
  )
}
```

# Final thoughts and takeaways

The APIs.guru catalog is a snapshot of the public web API ecosystem in miniature — and it has the shape of almost every other technology ecosystem: **power-law concentrated at the top, extraordinarily long-tailed below.**

A single hyperscaler (Amazon Web Services, by most accounts) accounts for a disproportionate share of catalogued APIs. But this metric flatters the giants: one AWS region expanding its service surface area adds dozens of specs. The far more numerous single-API providers represent the quiet majority — independent developers, regional SaaS companies, and niche services that collectively outnumber the platform giants by a wide margin.

The **category landscape** confirms what the cloud era has wrought: "Cloud" and developer tooling dominate, followed by communication and IoT categories that reflect the proliferation of connected infrastructure. The long tail of the category chart captures verticals still in the early innings of API standardization.

The **licensing gap** is perhaps the most actionable finding. A large fraction of APIs in the catalog carry no stated license. For developers building client libraries or wrappers — exactly the use case Jon Harmon is addressing with his package ecosystem — that ambiguity matters. The rise of OpenAPI as a standardized spec format is one half of the interoperability story; clear licensing is the other half that the ecosystem hasn't yet solved.

Finally, the **spec format transition** tells a gradualist story. Swagger 2.0 remains deeply embedded in the ecosystem's origin history, but APIs.guru's normalization layer and the slow uptick in OpenAPI 3.x adoption suggest the industry is moving — just not quickly.
