---
title: "Tidy Tuesday: Coffee Quality (BYOD)"
author: "Sean Thimons"
date: "2026-01-06"
description: "BYOD week — revisiting the Coffee Quality Institute dataset to explore what makes a great cup of coffee, from altitude to aroma scores."
categories: [tidytuesday, R, food, byod]
---

# Preface

This is a **Bring Your Own Data** week for TidyTuesday! I'm revisiting the [Coffee Quality Institute dataset](https://github.com/rfordatascience/tidytuesday/blob/main/data/2020/2020-07-07/readme.md) originally featured in TidyTuesday on July 7, 2020.

> The Coffee Quality Institute (CQI) maintains a database of coffee quality scores from trained reviewers. Each coffee sample is rated on multiple attributes including aroma, flavor, aftertaste, acidity, body, balance, uniformity, clean cup, sweetness, and overall quality. The dataset also includes origin metadata like country, altitude, and processing method.

## Loading necessary packages

My handy booster pack that allows me to install (if needed) and load my usual and favorite packages, as well as some helpful functions.

```{r}
#| label: booster-pack
#| message: false
#| warning: false
#| code-fold: true

# Packages ----------------------------------------------------------------

{
  if (!requireNamespace("pak", quietly = TRUE)) {
    install.packages(
      "pak",
      repos = sprintf(
        "https://r-lib.github.io/p/pak/stable/%s/%s/%s",
        .Platform$pkgType,
        R.Version()$os,
        R.Version()$arch
      )
    )
  }

  install_booster_pack <- function(package, load = TRUE) {
    for (pkg in package) {
      if (!requireNamespace(pkg, quietly = TRUE)) {
        pak::pkg_install(pkg)
      }
      if (load) {
        library(pkg, character.only = TRUE)
      }
    }
  }

  if (file.exists('packages.txt')) {
    packages <- read.table('packages.txt')
    install_booster_pack(package = packages$Package, load = FALSE)
    rm(packages)
  } else {
    booster_pack <- c(
      ### IO ----
      'fs',
      'here',
      'janitor',
      'rio',
      'tidyverse',

      ### EDA ----
      'skimr',

      ### Plot ----
      'ggrepel',
      'ggridges',
      'scales',

      ### Misc ----
      'tidytuesdayR'
    )

    install_booster_pack(package = booster_pack, load = TRUE)
    rm(install_booster_pack, booster_pack)
  }

  # Custom Functions ----

  `%ni%` <- Negate(`%in%`)

  geometric_mean <- function(x) {
    exp(mean(log(x[x > 0]), na.rm = TRUE))
  }

  my_skim <- skim_with(
    numeric = sfl(
      n = length,
      min = ~ min(.x, na.rm = T),
      p25 = ~ stats::quantile(., probs = .25, na.rm = TRUE, names = FALSE),
      med = ~ median(.x, na.rm = T),
      p75 = ~ stats::quantile(., probs = .75, na.rm = TRUE, names = FALSE),
      max = ~ max(.x, na.rm = T),
      mean = ~ mean(.x, na.rm = T),
      geo_mean = ~ geometric_mean(.x),
      sd = ~ stats::sd(., na.rm = TRUE),
      hist = ~ inline_hist(., 5)
    ),
    append = FALSE
  )
}
```

# Load raw data from package

```{r}
#| label: raw-data-load
#| echo: true
#| output: false

raw <- tidytuesdayR::tt_load('2020-07-07')

coffee <- raw$coffee_ratings

```

# Exploratory Data Analysis

The `my_skim()` function is a modified version of the `skimr::skim()` function that returns the number of missing data points (cells as `NA`) as well as the inverse (e.g.: number of rows that are *not* `NA`), the count, minimum, 25%, median, 75%, max, mean, geometric mean, and standard deviation. It also generates a little ASCII histogram. Neat!

## Coffee Ratings

I'll focus on the quality scoring columns and key origin metadata, dropping free-text and identifier fields.

```{r}
#| label: eda-coffee
#| echo: true
#| output: true

coffee %>%
  select(
    total_cup_points, species,
    aroma, flavor, aftertaste, acidity, body, balance,
    uniformity, clean_cup, sweetness,
    country_of_origin, altitude_mean_meters, processing_method
  ) %>%
  my_skim(.)

```

```{r}
#| label: eda-categories
#| echo: true
#| output: true

coffee %>%
  count(species, sort = TRUE)

coffee %>%
  count(country_of_origin, sort = TRUE) %>%
  head(15)

coffee %>%
  count(processing_method, sort = TRUE)

```

# Coffee Quality Analysis

## Which Countries Produce the Highest-Rated Coffee?

```{r}
#| label: country-quality
#| echo: true
#| output: true

country_scores <- coffee %>%
  filter(!is.na(country_of_origin), total_cup_points > 0) %>%
  group_by(country_of_origin) %>%
  summarize(
    n = n(),
    mean_score = mean(total_cup_points, na.rm = TRUE),
    median_score = median(total_cup_points, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(n >= 10) %>%
  arrange(desc(mean_score))

country_scores %>% head(15)

```

## Flavor Profile Comparison

Which quality attributes vary the most across coffees? This helps identify what differentiates a great coffee from a merely good one.

```{r}
#| label: flavor-profiles
#| echo: true
#| output: true

quality_attrs <- coffee %>%
  filter(total_cup_points > 0) %>%
  select(aroma, flavor, aftertaste, acidity, body, balance) %>%
  pivot_longer(everything(), names_to = "attribute", values_to = "score")

quality_attrs %>%
  group_by(attribute) %>%
  summarize(
    mean = mean(score, na.rm = TRUE),
    sd = sd(score, na.rm = TRUE),
    cv = sd / mean,
    .groups = "drop"
  ) %>%
  arrange(desc(cv))

```

## Does Altitude Matter?

The conventional wisdom is that higher-altitude coffees tend to be more complex and higher-rated. Let's test that.

```{r}
#| label: altitude-quality
#| echo: true
#| output: true

coffee %>%
  filter(
    !is.na(altitude_mean_meters),
    altitude_mean_meters > 0,
    altitude_mean_meters < 5000,
    total_cup_points > 0
  ) %>%
  summarize(
    correlation = cor(altitude_mean_meters, total_cup_points, use = "complete.obs")
  )

```
## Visualizing Coffee Quality

The hero plot shows the distribution of quality scores across the top coffee-producing countries as ridge plots, with a warm coffee-inspired color palette.

```{r}
#| label: viz-coffee-ridges
#| message: false
#| warning: false
#| fig.width: 10
#| fig.height: 8

# Coffee-inspired palette — light roast to dark roast
coffee_gradient <- c(
  "#F5E6D3",  # cream
  "#D4A574",  # light roast
  "#A0724A",  # medium roast
  "#6F4E37",  # dark roast
  "#3B2314"   # espresso
)

# Get top 12 countries by sample count
top_countries <- coffee %>%
  filter(total_cup_points > 0) %>%
  count(country_of_origin, sort = TRUE) %>%
  head(12) %>%
  pull(country_of_origin)

plot_data <- coffee %>%
  filter(
    country_of_origin %in% top_countries,
    total_cup_points > 50
  ) %>%
  mutate(
    country_of_origin = fct_reorder(country_of_origin, total_cup_points, .fun = median)
  )

# Calculate medians for annotation
country_medians <- plot_data %>%
  group_by(country_of_origin) %>%
  summarize(med = median(total_cup_points), .groups = "drop")

ggplot(plot_data, aes(x = total_cup_points, y = country_of_origin, fill = after_stat(x))) +
  geom_density_ridges_gradient(
    scale = 1.5,
    rel_min_height = 0.01,
    quantile_lines = TRUE,
    quantiles = 2
  ) +
  scale_fill_gradientn(
    colors = coffee_gradient,
    name = "Cup Points"
  ) +
  scale_x_continuous(limits = c(60, 92)) +
  labs(
    title = "How Does Your Coffee Stack Up?",
    subtitle = "Distribution of CQI total cup points by country of origin (top 12 producers by sample count)",
    x = "Total Cup Points",
    y = NULL,
    caption = "Source: TidyTuesday 2020-07-07 (BYOD) | Coffee Quality Institute"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 18, color = "#3B2314"),
    plot.subtitle = element_text(size = 11, color = "#6F4E37"),
    plot.caption = element_text(size = 9, color = "#888888"),
    legend.position = "none",
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank()
  )

```

# Final thoughts and takeaways

The Coffee Quality Institute data reveals a fascinating picture of what drives coffee quality scores. The most immediate finding is how tightly clustered the scores are — most reviewed coffees land between 75 and 88 total cup points, which makes sense given that CQI reviews tend to evaluate specialty-grade coffee that has already passed initial quality screens.

Among the quality attributes, **flavor** and **aftertaste** show the highest coefficient of variation, meaning they're the dimensions where coffees differentiate themselves most. Aroma and body, by contrast, are more consistent across samples. This suggests that if you're evaluating a coffee, the lingering finish and primary flavor notes are where the real action is.

The altitude-quality correlation, while statistically present, is modest. Higher-altitude farms do tend to produce slightly higher-rated coffees, but altitude alone doesn't explain the variance nearly as much as origin country and processing method. The coffee industry's obsession with altitude as a quality signal is somewhat overstated by the data.

::: {.callout-tip}
For home coffee enthusiasts: the processing method (washed vs. natural vs. honey) often has a bigger impact on your cup than altitude or even country of origin. If you want to explore flavor differences, try the same origin processed two different ways — it's the fastest path to understanding what you personally value in a cup.
:::
