---
title: "Tidy Tuesday: Wreck Inventory of Ireland"
author: "Sean Thimons"
date: "2026-06-30"
description: "Inventory records turn Irish shipwrecks into a coastline-shaped archive of vessel types, dates, and places of loss."
categories:
  - tidytuesday
  - maritime
  - history
  - ireland
---


# Preface

From the [TidyTuesday repository](https://github.com/rfordatascience/tidytuesday/tree/main/data/2026/2026-06-30).

The Irish wreck inventory is not just a catalogue of ship names. It is a coastline-shaped memory system: 17,981 records, a long tail of vessel classifications, narrative descriptions of weather and cargo, and enough coordinates to show where archive work becomes geography.

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(tidytuesdayR)
library(scales)

raw <- tidytuesdayR::tt_load("2026-06-30")
wreck_inventory <- raw$wreck_inventory
```

## What is in the file

The dataset contains a loss date, parsed year, wreck name and number, classification, a free-text description, place of loss, coordinates, and coordinate source. I treated it as an inventory rather than a casualty model: the important first pass is coverage, not causal inference.

| Measure | Value |
|---|---:|
| Records | 17,981 |
| Distinct classifications | 138 |
| Records with usable coordinates | 3,564 |
| Earliest parsed year | 1300s |
| Latest parsed year | 2010s |

## Vessel types

```{r}
#| label: fig-classification
#| fig-cap: "Most common wreck classifications in the Irish inventory."
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 5

wreck_inventory %>%
  filter(!is.na(classification), classification != "") %>%
  count(classification, sort = TRUE) %>%
  slice_head(n = 15) %>%
  mutate(classification = fct_reorder(classification, n)) %>%
  ggplot(aes(n, classification)) +
  geom_col(fill = "#2f6f73") +
  scale_x_continuous(labels = comma) +
  labs(x = "Wreck records", y = NULL) +
  theme_minimal(base_size = 12)
```

The top of the classification table is practical rather than romantic: schooners, steamers, trawlers, and other working vessels dominate the named categories. That matters because it frames the inventory as commercial and coastal infrastructure history, not only spectacular disasters.

## Time

```{r}
#| label: fig-decades
#| fig-cap: "Recorded wreck losses by decade."
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 4.5

wreck_inventory %>%
  filter(!is.na(year)) %>%
  mutate(decade = floor(year / 10) * 10) %>%
  count(decade) %>%
  ggplot(aes(decade, n)) +
  geom_col(fill = "#8b5e34", width = 8) +
  scale_x_continuous(breaks = breaks_width(50)) +
  scale_y_continuous(labels = comma) +
  labs(x = "Decade", y = "Wreck records") +
  theme_minimal(base_size = 12)
```

The decade curve should be read as recorded inventory intensity, not a clean rate of maritime danger. Still, the broad shape is useful. The 19th and early 20th century concentration reflects the age when Irish coastal trade, sail-to-steam transition, and documentary survival overlap most visibly.

## Place

```{r}
#| label: fig-locations
#| fig-cap: "Coordinates in the register trace the Irish coastline even without a basemap."
#| message: false
#| warning: false
#| fig-width: 6
#| fig-height: 6

wreck_inventory %>%
  filter(!is.na(longitude), !is.na(latitude)) %>%
  ggplot(aes(longitude, latitude)) +
  geom_point(color = "#12355b", alpha = 0.28, size = 0.7) +
  coord_fixed() +
  labs(x = "Longitude", y = "Latitude") +
  theme_minimal(base_size = 12)
```

The coordinate plot is deliberately plain: longitude against latitude. Even without a basemap, the island appears through absence and edge density. That is the signature of a place-based register — the data are about named events, but the spatial pattern is the coast doing what coasts do to vessels.

## Takeaway

This week rewards humble analysis. The wreck inventory is rich, but the mixed descriptions and uneven historical coverage make it risky to over-fit. The honest story is that administrative memory has a shape: categories reveal working vessels, dates reveal eras of documentation, and coordinates reveal where loss became record.

