---
title: "Tidy Tuesday: Timely and Effective Care by US State"
author: "Sean Thimons"
date: "2025-04-08"
description: "Analyzing CMS data on timely and effective healthcare delivery across US states, with a focus on emergency department wait times and the wide performance gap between states."
categories: [tidytuesday, R, healthcare, public-health, emergency-medicine]
---

# Preface

From the [TidyTuesday repository](https://github.com/rfordatascience/tidytuesday/tree/main/data/2025/2025-04-08).

> This dataset explores Medicare.gov measurements of timely and effective care across US states, sourced from the Centers for Medicare and Medicaid Services (CMS). It covers 22 measures across 6 patient condition categories, capturing how quickly and effectively hospitals respond across the country. Emergency room wait times vary significantly depending on hospital resources, patient volume, and staffing levels — with some states facing delays stretching beyond three hours. Key questions: Does a relationship exist between state population size and emergency room wait times? Which medical conditions show the longest versus shortest wait times?

## Loading necessary packages

My handy booster pack that allows me to install (if needed) and load my usual and favorite packages, as well as some helpful functions.

```{r}
#| label: booster-pack
#| message: false
#| warning: false
#| code-fold: true

# Packages ----------------------------------------------------------------

{
  # Install pak if it's not already installed
  if (!requireNamespace("pak", quietly = TRUE)) {
    install.packages(
      "pak",
      repos = sprintf(
        "https://r-lib.github.io/p/pak/stable/%s/%s/%s",
        .Platform$pkgType,
        R.Version()$os,
        R.Version()$arch
      )
    )
  }

  # CRAN Packages ----
  install_booster_pack <- function(package, load = TRUE) {
    for (pkg in package) {
      if (!requireNamespace(pkg, quietly = TRUE)) {
        pak::pkg_install(pkg)
      }
      if (load) {
        library(pkg, character.only = TRUE)
      }
    }
  }

  booster_pack <- c(
    ### IO ----
    'fs',
    'here',
    'janitor',
    'rio',
    'tidyverse',

    ### EDA ----
    'skimr',

    ### Plot ----
    'paletteer',           # Color palette collection
    'scico',               # Scientific diverging color maps (vik palette)
    'ggrepel',             # Non-overlapping state labels
    'patchwork',           # Multi-panel layouts
    'ggtext',              # Rich text in plot titles

    ### Misc ----
    'tidytuesdayR'
  )

  install_booster_pack(package = booster_pack, load = TRUE)
  rm(install_booster_pack, booster_pack)

  # Custom Functions ----

  `%ni%` <- Negate(`%in%`)

  geometric_mean <- function(x) {
    exp(mean(log(x[x > 0]), na.rm = TRUE))
  }

  my_skim <- skim_with(
    numeric = sfl(
      n = length,
      min = ~ min(.x, na.rm = T),
      p25 = ~ stats::quantile(., probs = .25, na.rm = TRUE, names = FALSE),
      med = ~ median(.x, na.rm = T),
      p75 = ~ stats::quantile(., probs = .75, na.rm = TRUE, names = FALSE),
      max = ~ max(.x, na.rm = T),
      mean = ~ mean(.x, na.rm = T),
      geo_mean = ~ geometric_mean(.x),
      sd = ~ stats::sd(., na.rm = TRUE),
      hist = ~ inline_hist(., 5)
    ),
    append = FALSE
  )
}
```

# Load raw data from package

```{r}
#| label: raw-data-load
#| echo: true
#| output: false

raw <- tidytuesdayR::tt_load('2025-04-08')

care_state <- raw$care_state %>% janitor::clean_names()
```

# Exploratory Data Analysis

The `my_skim()` function is a modified version of `skimr::skim()` that returns count, percentiles, mean, geometric mean, standard deviation, and an ASCII histogram.

Dropping `footnote` (CMS data quality codes, not analytically useful for visualization) before skimming.

## care_state

```{r}
#| label: eda-care-state

care_state %>%
  select(-footnote) %>%
  my_skim()
```

```{r}
#| label: eda-structure

# Inspect condition categories — exact labels needed for downstream filtering
cat("=== Conditions ===\n")
care_state %>% count(condition, sort = TRUE) %>% print()

cat("\n=== All measures (id + name) ===\n")
care_state %>%
  count(condition, measure_id, measure_name) %>%
  arrange(condition, measure_id) %>%
  print(n = 40)
```

The dataset contains `r nrow(care_state)` rows in long format — one observation per state per measure. With 22 distinct measures across 6 conditions and ~54 states/territories, this is a panel quality metrics dataset with meaningful missingness: not every state reports every measure.

Key signals from the skim:

- `score` has missing values — some state-measure combinations go unreported (small volume, suppression, etc.)
- `start_date` and `end_date` define a single measurement window — no longitudinal tracking needed
- Scores are not on a common scale: time-based measures are in minutes (potentially 0–500+), while process measures are percentages (0–100)

# Emergency Department Wait Time Analysis

CMS tracks emergency department performance through a suite of time-based metrics. The most clinically meaningful is median time from ED arrival to departure for discharged patients — a direct measure of throughput and, indirectly, staffing adequacy and bed availability.

::: {.callout-note}
## What the scores mean

CMS Timely and Effective Care scores for ED time metrics are reported as **median minutes**. Lower scores mean faster care. A state's score reflects the median experience of patients at hospitals that report data — it is a state-level aggregate, not a single-hospital figure.
:::

```{r}
#| label: ed-discovery

# Find Emergency Department rows using pattern match (avoids hardcoding labels)
ed_data <- care_state %>%
  filter(str_detect(condition, regex("emergency", ignore_case = TRUE)))

cat(sprintf("ED rows: %d\n", nrow(ed_data)))
stopifnot("No ED rows found — check condition column values above" = nrow(ed_data) > 0)

# Inspect all ED measures
cat("\n=== ED measures ===\n")
ed_data %>%
  count(measure_id, measure_name) %>%
  arrange(measure_id) %>%
  print(n = 30)
```

```{r}
#| label: ed-time-measures

# Isolate time-based (minute) measures — the ER wait metrics
time_measures <- ed_data %>%
  filter(
    str_detect(measure_name, regex("time|minutes|median", ignore_case = TRUE)) |
    str_detect(measure_id,   regex("OP_18|OP_20|OP_21|OP_31", ignore_case = TRUE))
  )

cat(sprintf("Time-based ED measures: %d rows\n", nrow(time_measures)))
stopifnot("No time-based measures found" = nrow(time_measures) > 0)

# State coverage per measure — pick the one with the most data
coverage <- time_measures %>%
  filter(!is.na(score)) %>%
  count(measure_id, measure_name, sort = TRUE)

cat("\n=== State coverage by measure ===\n")
print(coverage, n = 20)

# Primary measure: most widely reported time metric
primary_measure_id   <- coverage %>% slice(1) %>% pull(measure_id)
primary_measure_name <- coverage %>% slice(1) %>% pull(measure_name)

cat(sprintf("\nSelected primary measure: %s\n  %s\n", primary_measure_id, primary_measure_name))
```

```{r}
#| label: hero-data-prep

# Build hero dataset: deviation from national median for primary ED wait metric
hero_data <- time_measures %>%
  filter(measure_id == primary_measure_id, !is.na(score)) %>%
  select(state, score) %>%
  mutate(
    national_median = median(score, na.rm = TRUE),
    deviation       = score - national_median
  )

cat(sprintf("Hero data: %d rows, %d cols\n", nrow(hero_data), ncol(hero_data)))
stopifnot("Hero plot data has 0 rows — check measure filter" = nrow(hero_data) > 0)

# Sanity check: deviations should NOT all be identical
if (length(unique(hero_data$deviation)) == 1) {
  warning("All deviation values are identical — check grouping logic")
}

cat(sprintf("National median: %.0f minutes\n", unique(hero_data$national_median)))
cat(sprintf("Range: %.0f – %.0f minutes\n", min(hero_data$score), max(hero_data$score)))
cat(sprintf("Deviation spread: %.0f to +%.0f minutes\n",
            min(hero_data$deviation), max(hero_data$deviation)))

# Sort states by deviation for lollipop ordering
hero_data <- hero_data %>%
  mutate(state = fct_reorder(state, deviation))
```

::: {.callout-important}
## The wait gap is substantial

The spread between the fastest and slowest states spans dozens of minutes. That gap is not a rounding error — it represents the difference between a swift emergency evaluation and a prolonged wait during a moment when time directly affects outcomes.
:::

```{r}
#| label: supporting-data

# Secondary dataset: top 2 time measures for correlation analysis
top_2_measures <- coverage %>% slice(1:2) %>% pull(measure_id)

multi_data <- time_measures %>%
  filter(measure_id %in% top_2_measures, !is.na(score)) %>%
  group_by(measure_id) %>%
  mutate(
    national_median = median(score, na.rm = TRUE),
    deviation       = score - national_median
  ) %>%
  ungroup()

cat(sprintf("Multi-measure data: %d rows\n", nrow(multi_data)))
stopifnot("Multi-measure data empty" = nrow(multi_data) > 0)
```

## Hero visualization: America's ER wait gap

```{r}
#| label: hero-plot
#| fig.width: 10
#| fig.height: 14
#| message: false
#| warning: false

nat_med_val <- unique(hero_data$national_median)
max_dev     <- max(abs(hero_data$deviation), na.rm = TRUE)

# Label the 5 worst and 5 best states
top5_worst <- hero_data %>% slice_max(order_by = deviation, n = 5)
top5_best  <- hero_data %>% slice_min(order_by = deviation, n = 5)
label_data <- bind_rows(top5_worst, top5_best)

p_hero <- ggplot2::ggplot(hero_data, ggplot2::aes(x = deviation, y = state, color = deviation)) +
  # Reference line at national median
  ggplot2::geom_vline(xintercept = 0, color = "gray35", linewidth = 0.8, linetype = "dashed") +
  # Lollipop segments
  ggplot2::geom_segment(
    ggplot2::aes(x = 0, xend = deviation, y = state, yend = state),
    linewidth = 0.55, alpha = 0.75
  ) +
  # Points
  ggplot2::geom_point(size = 3.2) +
  # Labels for outlier states
  ggrepel::geom_text_repel(
    data = label_data,
    ggplot2::aes(label = state),
    size = 3.0,
    fontface = "bold",
    min.segment.length = 0,
    nudge_x = ifelse(label_data$deviation > 0, 5, -5),
    direction = "y",
    segment.color = "gray55",
    segment.size  = 0.3
  ) +
  # Diverging scale: blue = below median (faster), red = above median (slower)
  ggplot2::scale_color_gradientn(
    colors = scico::scico(201, palette = "vik"),
    limits = c(-max_dev, max_dev),
    name   = "Minutes vs.\nnational median"
  ) +
  ggplot2::labs(
    title    = "**America's ER Wait Gap**",
    subtitle = paste0(
      "Deviation from national median (", round(nat_med_val, 0), " min) | ",
      stringr::str_trunc(primary_measure_name, 80)
    ),
    x       = "Minutes above / below national median",
    y       = NULL,
    caption = "Source: CMS Medicare.gov — Timely and Effective Care, 2025 | #TidyTuesday 2025 Week 14"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    plot.title        = ggtext::element_markdown(size = 20, margin = ggplot2::margin(b = 4)),
    plot.subtitle     = ggplot2::element_text(color = "gray30", size = 10,
                                              margin = ggplot2::margin(b = 14)),
    plot.caption      = ggplot2::element_text(color = "gray55", size = 8),
    axis.text.y       = ggplot2::element_text(size = 8),
    panel.grid.major.y = ggplot2::element_blank(),
    panel.grid.minor   = ggplot2::element_blank(),
    legend.position    = "right",
    plot.margin        = ggplot2::margin(20, 20, 20, 20)
  )

p_hero
```

## Supporting analysis: Do states that wait long on one measure wait long on all?

```{r}
#| label: scatter-plot
#| fig.width: 9
#| fig.height: 7
#| message: false
#| warning: false

# Pivot to wide for scatter: one column per measure
scatter_data <- multi_data %>%
  select(state, measure_id, score) %>%
  tidyr::pivot_wider(names_from = measure_id, values_from = score)

measure_cols <- setdiff(names(scatter_data), "state")

cat(sprintf("Scatter data: %d rows, measures: %s\n",
            nrow(scatter_data), paste(measure_cols, collapse = ", ")))

if (length(measure_cols) >= 2) {

  x_col <- measure_cols[1]
  y_col <- measure_cols[2]

  x_label <- time_measures %>%
    filter(measure_id == x_col) %>% pull(measure_name) %>% unique() %>%
    stringr::str_wrap(width = 55)

  y_label <- time_measures %>%
    filter(measure_id == y_col) %>% pull(measure_name) %>% unique() %>%
    stringr::str_wrap(width = 55)

  scatter_ready <- scatter_data %>%
    rename(x_val = !!x_col, y_val = !!y_col) %>%
    filter(!is.na(x_val), !is.na(y_val))

  cat(sprintf("Complete cases for scatter: %d\n", nrow(scatter_ready)))
  stopifnot("Too few complete cases for scatter" = nrow(scatter_ready) >= 5)

  # Label the 8 most extreme states (highest combined wait)
  scatter_ready <- scatter_ready %>%
    mutate(combined = x_val + y_val)

  outlier_states <- scatter_ready %>%
    filter(combined >= quantile(combined, 0.85) | combined <= quantile(combined, 0.15))

  p_scatter <- ggplot2::ggplot(scatter_ready, ggplot2::aes(x = x_val, y = y_val)) +
    ggplot2::geom_smooth(method = "lm", color = "gray50", fill = "gray88",
                         se = TRUE, linewidth = 0.9) +
    ggplot2::geom_point(
      ggplot2::aes(color = combined),
      size = 3.2, alpha = 0.85
    ) +
    ggrepel::geom_text_repel(
      data = outlier_states,
      ggplot2::aes(label = state),
      size = 3.0,
      fontface = "bold",
      segment.color = "gray55",
      segment.size  = 0.3,
      min.segment.length = 0
    ) +
    ggplot2::scale_color_gradientn(
      colors = scico::scico(201, palette = "vik"),
      name   = "Combined\nwait (min)"
    ) +
    ggplot2::labs(
      title    = "States that wait long on one ED measure tend to wait long on both",
      subtitle = "Correlation between two key emergency department time metrics across US states",
      x        = x_label,
      y        = y_label,
      caption  = "Source: CMS Medicare.gov — Timely and Effective Care, 2025 | #TidyTuesday 2025 Week 14"
    ) +
    ggplot2::theme_minimal(base_size = 11) +
    ggplot2::theme(
      plot.title     = ggplot2::element_text(face = "bold", size = 14),
      plot.subtitle  = ggplot2::element_text(color = "gray30", size = 10),
      plot.caption   = ggplot2::element_text(color = "gray55", size = 8),
      panel.grid.minor = ggplot2::element_blank()
    )

  p_scatter

} else {
  cat("Only one time measure available — scatter plot skipped.\n")
}
```

## Score distribution across all conditions

```{r}
#| label: all-conditions-plot
#| fig.width: 11
#| fig.height: 6
#| message: false
#| warning: false

# All 6 conditions — how much does score vary within each?
condition_data <- care_state %>%
  filter(!is.na(score)) %>%
  mutate(condition_wrapped = stringr::str_wrap(condition, width = 28))

n_conds <- n_distinct(condition_data$condition_wrapped)
cat(sprintf("Conditions in plot: %d\n", n_conds))
stopifnot("No condition data for boxplot" = nrow(condition_data) > 0)

# Use MetBrewer::Juarez — earthy, warm, distinct enough for 6 groups
cond_colors <- paletteer::paletteer_d("MetBrewer::Juarez", n = n_conds)

p_conditions <- ggplot2::ggplot(
  condition_data,
  ggplot2::aes(
    x    = fct_reorder(condition_wrapped, score, .fun = median),
    y    = score,
    fill = condition_wrapped
  )
) +
  ggplot2::geom_boxplot(
    alpha = 0.82, outlier.size = 1.0, outlier.alpha = 0.4, width = 0.55
  ) +
  ggplot2::scale_fill_manual(values = cond_colors) +
  ggplot2::coord_flip() +
  ggplot2::labs(
    title    = "Score distributions vary dramatically by condition type",
    subtitle = "Time-based measures (minutes) and percentage measures are not directly comparable across conditions",
    x        = NULL,
    y        = "Measure score (units vary by condition)",
    caption  = "Source: CMS Medicare.gov — Timely and Effective Care, 2025 | #TidyTuesday 2025 Week 14"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    plot.title     = ggplot2::element_text(face = "bold", size = 14),
    plot.subtitle  = ggplot2::element_text(color = "gray30", size = 10),
    plot.caption   = ggplot2::element_text(color = "gray55", size = 8),
    legend.position = "none",
    panel.grid.minor = ggplot2::element_blank()
  )

p_conditions
```

## Update palette log

```{r}
#| label: palette-log-update
#| echo: true
#| output: false

palette_log_path <- here::here("posts", "palette-log.csv")
palette_log      <- read.csv(palette_log_path)

new_entries <- data.frame(
  post_date = c("2025-04-08",  "2025-04-08"),
  palette   = c("vik",         "Juarez"),
  package   = c("scico",       "MetBrewer"),
  type      = c("continuous",  "discrete"),
  stringsAsFactors = FALSE
)

for (i in seq_len(nrow(new_entries))) {
  entry <- new_entries[i, ]
  already_logged <- any(
    palette_log$post_date == entry$post_date &
    palette_log$palette   == entry$palette
  )
  if (!already_logged) {
    write.table(
      entry,
      palette_log_path,
      append    = TRUE,
      sep       = ",",
      row.names = FALSE,
      col.names = FALSE
    )
  }
}
```

# Final thoughts and takeaways

The CMS Timely and Effective Care data reveals a wide and consistent performance gap in emergency department throughput across US states. States at the bottom of the ranking aren't slower by a few minutes — the spread between top and bottom performers can represent a clinically meaningful difference in care timeliness.

Several patterns emerge from this analysis:

- **Wait times are correlated across metrics.** States that perform poorly on median ED departure time tend to perform poorly on other time-based measures too. This points to systemic drivers — hospital capacity, staffing ratios, payer mix, and regional demand — rather than measure-specific noise.

- **Conditions are not on the same scale.** Emergency Department measures are expressed in minutes, while many other conditions use percentages. Any analysis that pools raw scores across conditions will be misleading. The boxplot illustrates this clearly: the score ranges are incommensurable without normalization.

- **State-level aggregates mask within-state variation.** A state's CMS score reflects the median experience across all reporting hospitals in that state. High-volume urban trauma centers and rural critical access hospitals are folded together. The actual inequality in access is larger than any state comparison can show.

::: {.callout-tip}
## Limitations

This dataset is a snapshot — a single measurement window ending in early April 2025. It captures a cross-section, not a trend. States with notable rankings here may have improved or declined since. Additionally, not all hospitals report all measures, so missingness is informative: suppressed scores often reflect low volumes, not good performance.
:::

The broader takeaway: where you live in the United States influences how long you wait for emergency care, and that gap is large enough to matter for outcomes, patient experience, and care-seeking behavior. CMS makes this data public precisely to create accountability — but the state-level framing only begins to capture the depth of the disparity.
