---
title: "Tidy Tuesday: Australian Frogs"
author: "Sean Thimons"
date: "2025-09-02"
description: "Mapping the seasonal calling calendar of Australia's frogs using 136,000 citizen-science observations from the 2023 FrogID survey."
categories: [tidytuesday, R, ecology, citizen-science, australia, wildlife]
---

# Preface

From the [TidyTuesday repository](https://github.com/rfordatascience/tidytuesday/tree/main/data/2025/2025-09-02).

> FrogID is a citizen science initiative allowing Australians to record and submit frog calls for expert identification. The 2023 dataset represents the sixth annual data release, contributing to over 30 scientific papers on frog ecology, taxonomy, and conservation since 2017. Australia hosts 257 native frog species found nowhere else on Earth — yet almost one in five species are threatened with extinction due to climate change, urbanisation, disease, and invasive species.

Suggested research questions:

- Which frog species are endemic to specific Australian regions?
- Do different species exhibit distinct seasonal calling patterns?
- Which species has the broadest geographic distribution versus the most limited range?

## Loading necessary packages

My handy booster pack that allows me to install (if needed) and load my usual and favorite packages, as well as some helpful functions.

```{r}
#| label: booster-pack
#| message: false
#| warning: false
#| code-fold: true

# Packages ----------------------------------------------------------------

{
  # Install pak if it's not already installed
  if (!requireNamespace("pak", quietly = TRUE)) {
    install.packages(
      "pak",
      repos = sprintf(
        "https://r-lib.github.io/p/pak/stable/%s/%s/%s",
        .Platform$pkgType,
        R.Version()$os,
        R.Version()$arch
      )
    )
  }

  # CRAN Packages ----
  install_booster_pack <- function(package, load = TRUE) {
    for (pkg in package) {
      if (!requireNamespace(pkg, quietly = TRUE)) {
        pak::pkg_install(pkg)
      }
      if (load) {
        library(pkg, character.only = TRUE)
      }
    }
  }

  booster_pack <- c(
    ### IO ----
    'fs',
    'here',
    'janitor',
    'rio',
    'tidyverse',

    ### EDA ----
    'skimr',

    ### Plot ----
    'paletteer',           # Color palette collection
    'ggtext',              # Rich text (italic species names)
    'ggrepel',             # Non-overlapping labels
    'scales',              # Axis formatting

    ### Misc ----
    'tidytuesdayR'
  )

  install_booster_pack(package = booster_pack, load = TRUE)
  rm(install_booster_pack, booster_pack)

  # Custom Functions ----

  `%ni%` <- Negate(`%in%`)

  geometric_mean <- function(x) {
    exp(mean(log(x[x > 0]), na.rm = TRUE))
  }

  my_skim <- skim_with(
    numeric = sfl(
      n = length,
      min = ~ min(.x, na.rm = T),
      p25 = ~ stats::quantile(., probs = .25, na.rm = TRUE, names = FALSE),
      med = ~ median(.x, na.rm = T),
      p75 = ~ stats::quantile(., probs = .75, na.rm = TRUE, names = FALSE),
      max = ~ max(.x, na.rm = T),
      mean = ~ mean(.x, na.rm = T),
      geo_mean = ~ geometric_mean(.x),
      sd = ~ stats::sd(., na.rm = TRUE),
      hist = ~ inline_hist(., 5)
    ),
    append = FALSE
  )
}
```

# Load raw data from package

```{r}
#| label: raw-data-load
#| echo: true
#| output: false

raw <- tidytuesdayR::tt_load('2025-09-02')

frog_data  <- raw$frogID_data   # 136,621 citizen-science observation records
frog_names <- raw$frog_names    # 294-row taxonomic lookup (subfamily, common names)
```

# Exploratory Data Analysis

The `my_skim()` function returns count, min, p25, median, p75, max, mean, geometric mean, and standard deviation, plus an ASCII histogram.

## frogID_data — occurrence records

```{r}
#| label: eda-frogid-data
#| echo: true
#| output: true

# Drop identifier and recorder columns — not analytically useful
frog_data %>%
  select(-occurrenceID, -eventID, -recordedBy) %>%
  my_skim()
```

```{r}
#| label: eda-categorical-overview
#| echo: true
#| output: true

# State distribution
cat("=== Observations by State ===\n")
frog_data %>% count(stateProvince, sort = TRUE) %>% print(n = 10)

# Top observed species
cat("\n=== Top 15 Species ===\n")
frog_data %>% count(scientificName, sort = TRUE) %>% head(15) %>% print(n = 15)

# Monthly rhythm
cat("\n=== Observations by Month ===\n")
frog_data %>%
  mutate(month = month(eventDate, label = TRUE, abbr = TRUE)) %>%
  count(month) %>%
  print(n = 12)
```

The `frogID_data` table contains **136,621 records** spanning January–November 2023, covering **186 unique species** across all Australian states and territories. A few immediate structural features stand out:

- **New South Wales dominates the record count** (58,749 — 43% of all observations), likely reflecting both its large population of citizen scientists and the density of the FrogID app's user base along the eastern seaboard.
- **Queensland leads in species richness** (96 unique species), consistent with its tropical biodiversity, even though NSW contributes far more raw observations.
- **The seasonal signal is unmistakable.** November alone accounts for nearly 23,000 observations — over 16% of the annual total — while April–June are quietest. This spring peak reflects the southern hemisphere breeding season for most temperate species.
- **Crinia signifera (Common Eastern Froglet)** accounts for 33,630 records — nearly 25% of all observations — making it by far the most-recorded species in this dataset.

## frog_names — taxonomic lookup

```{r}
#| label: eda-frog-names
#| echo: true
#| output: true

frog_names %>%
  select(subfamily, tribe, scientificName, commonName) %>%
  my_skim()

# Subfamily breakdown
cat("\n=== Species count by subfamily ===\n")
frog_names %>%
  filter(str_detect(scientificName, " ")) %>%  # binomial names only
  count(subfamily, sort = TRUE)
```

The `frog_names` table is a 294-row taxonomic dictionary. Most entries belong to **Myobatrachidae** (ground frogs, endemic to Australia and New Guinea) and **Hylidae** (tree frogs). A handful of rows are genus-level entries without a species epithet — these need to be filtered out before joining to avoid many-to-many matches.

# Seasonal Calling Phenology: When Do Australia's Frogs Sing?

Frog calling is tightly coupled to breeding: males call to attract females, so call frequency spikes during the breeding season. In Australia, this creates a stark biogeographic divide:

::: {.callout-note}
**The north-south seasonal split.** Tropical frogs in Queensland and the Northern Territory breed during the wet season (November–April), triggered by monsoon rains. Temperate frogs in NSW, Victoria, and South Australia breed in spring and early summer (August–November), responding to warming temperatures. The FrogID data, collected across 2023, captures both signals simultaneously.
:::

## Species richness vs. observation effort by state

```{r}
#| label: analysis-state-richness
#| echo: true
#| output: true

state_richness <- frog_data %>%
  group_by(stateProvince) %>%
  summarise(
    n_species = n_distinct(scientificName),
    n_obs     = n(),
    .groups   = "drop"
  ) %>%
  arrange(desc(n_species))

cat("Species richness and observation effort by state:\n")
print(state_richness)
```

Queensland has 96 unique species — more than triple Victoria's 28 — yet NSW generates 2.5× more observations. This gap between *where frogs are* and *where people are looking* is the classic citizen-science sampling bias. The FrogID team explicitly accounts for this in their conservation analyses.

## Constructing the seasonal calling calendar

```{r}
#| label: analysis-seasonal-calendar
#| echo: true
#| output: true

# Filter frog_names to binomial species only to avoid genus-level many-to-many join
frog_names_sp <- frog_names %>%
  filter(str_detect(scientificName, " ")) %>%
  select(scientificName, commonName) %>%
  distinct(scientificName, .keep_all = TRUE)

# Identify top 15 species by observation count
top15_species <- frog_data %>%
  count(scientificName, sort = TRUE) %>%
  head(15) %>%
  pull(scientificName)

cat("Top 15 species selected for heatmap:\n")
print(top15_species)

# Build monthly proportion matrix for top 15 species
monthly_matrix <- frog_data %>%
  filter(scientificName %in% top15_species) %>%
  mutate(month_num = month(eventDate)) %>%
  count(scientificName, month_num) %>%
  group_by(scientificName) %>%
  mutate(pct = n / sum(n)) %>%
  ungroup()

# Verify data integrity
cat(sprintf("\nmonthly_matrix: %d rows, %d cols\n", nrow(monthly_matrix), ncol(monthly_matrix)))
stopifnot("Plot data has 0 rows" = nrow(monthly_matrix) > 0)

# Quick sanity check on proportions
cat("Row sums per species (should all be ~1.0):\n")
monthly_matrix %>%
  group_by(scientificName) %>%
  summarise(total_pct = sum(pct)) %>%
  print(n = 15)

# Determine peak calling month for each species (for ordering)
peak_month <- monthly_matrix %>%
  group_by(scientificName) %>%
  slice_max(pct, n = 1, with_ties = FALSE) %>%
  select(scientificName, peak_month = month_num)

# Join common names
monthly_matrix <- monthly_matrix %>%
  left_join(frog_names_sp, by = "scientificName") %>%
  left_join(peak_month, by = "scientificName") %>%
  mutate(
    # Use common name where available; fall back to italicised scientific name
    display_name = case_when(
      !is.na(commonName) & commonName != "—" ~
        paste0(commonName, "\n*", scientificName, "*"),
      TRUE ~
        paste0("*", scientificName, "*")
    ),
    # Order species by peak calling month
    display_name = fct_reorder(display_name, peak_month),
    month_abbr = factor(
      month.abb[month_num],
      levels = month.abb
    )
  )

cat("\nmonthly_matrix after join and mutate: ", nrow(monthly_matrix), "rows\n")
```

# Visualisation: The Frog Calling Calendar

```{r}
#| label: hero-plot-heatmap
#| message: false
#| warning: false
#| fig.width: 11
#| fig.height: 8

# Palette: scico::bamako (sequential, unused)
# Low = dark navy/purple (quiet month), High = warm cream/gold (peak calling)

p <- ggplot2::ggplot(
  monthly_matrix,
  ggplot2::aes(x = month_abbr, y = display_name, fill = pct)
) +
  ggplot2::geom_tile(colour = "white", linewidth = 0.4) +
  # Annotate the spring and wet-season peaks
  ggplot2::annotate(
    "rect",
    xmin = 8.5, xmax = 11.5,
    ymin = 0.5, ymax = 15.5,
    fill = NA, colour = "#F5D76E", linewidth = 0.7, linetype = "dashed",
    alpha = 0.5
  ) +
  ggplot2::annotate(
    "text",
    x = 10, y = 15.7,
    label = "Southern spring\nbreeding season",
    colour = "#F5D76E", size = 3, fontface = "bold", hjust = 0.5
  ) +
  paletteer::scale_fill_paletteer_c(
    "scico::bamako",
    direction  = -1,
    labels     = scales::label_percent(accuracy = 1),
    name       = "% of annual\nobservations",
    guide      = ggplot2::guide_colorbar(
      barwidth  = 0.6,
      barheight = 8,
      title.position = "top"
    )
  ) +
  ggplot2::scale_x_discrete(expand = ggplot2::expansion(0, 0)) +
  ggplot2::scale_y_discrete(expand = ggplot2::expansion(0, 0)) +
  ggplot2::labs(
    title    = "When Australia's Frogs Call",
    subtitle = "Monthly calling activity (as % of annual observations) for the 15 most-recorded species in the 2023 FrogID citizen-science survey.\nSpecies ordered by peak calling month. Dashed box = southern hemisphere spring.",
    x        = NULL,
    y        = NULL,
    caption  = "Source: FrogID 2023 Annual Data Release via TidyTuesday (2025-09-02) · Palette: scico::bamako"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    plot.title         = ggplot2::element_text(face = "bold", size = 17, margin = ggplot2::margin(b = 4)),
    plot.subtitle      = ggtext::element_textbox_simple(
                           size = 9, colour = "grey40",
                           margin = ggplot2::margin(b = 12), lineheight = 1.3
                         ),
    plot.caption       = ggplot2::element_text(colour = "grey55", size = 7.5, margin = ggplot2::margin(t = 8)),
    axis.text.y        = ggtext::element_markdown(size = 8.5, hjust = 1),
    axis.text.x        = ggplot2::element_text(size = 9, face = "bold"),
    panel.grid         = ggplot2::element_blank(),
    legend.title       = ggplot2::element_text(size = 8),
    legend.text        = ggplot2::element_text(size = 7.5),
    plot.margin        = ggplot2::margin(12, 16, 12, 12)
  )

p
```

```{r}
#| label: palette-log-update
#| echo: false
#| output: false

palette_log_path <- here::here("posts", "palette-log.csv")
palette_log      <- read.csv(palette_log_path)

new_entry <- data.frame(
  post_date = "2025-09-02",
  palette   = "bamako",
  package   = "scico",
  type      = "continuous"
)

if (!any(palette_log$post_date == new_entry$post_date &
         palette_log$palette   == new_entry$palette)) {
  write.table(
    new_entry,
    palette_log_path,
    append    = TRUE,
    sep       = ",",
    row.names = FALSE,
    col.names = FALSE
  )
  cat("Palette log updated: scico::bamako added for 2025-09-02\n")
} else {
  cat("Palette log entry already exists — skipping.\n")
}
```

# Final thoughts and takeaways

The heatmap makes a few things immediately clear:

**1. Spring is frog season — for most species.** The dashed box covering August–November captures the peak calling period for the majority of the top-15 species. This aligns with the southern hemisphere spring, when temperatures warm and rains arrive across eastern Australia. *Crinia signifera* (Common Eastern Froglet), which alone accounts for 25% of all observations, is remarkably evenly spread across the year but shows a sharp November peak.

**2. The tropical outliers stand out.** *Uperoleia* species and Queensland-heavy frogs (appearing lower in the ordered list) show relatively flat or early-year distributions, hinting at the wet-season breeding calendar of tropical Australia — though the citizen-science sampling bias toward NSW blunts this signal in the aggregate data.

**3. Citizen science has a seasonality of its own.** The November surge isn't only frogs calling more — it's also people going outside more. Disentangling phenological signal from observer effort is a persistent challenge in FrogID analyses. The research team publishes annual corrections for this bias, which is partly why FrogID requires expert acoustic verification rather than relying solely on observer labels.

**4. 186 species in a single year is remarkable — and not enough.** The 2023 dataset captures less than three-quarters of Australia's 257 native frog species. Many threatened species are either in remote areas with few citizen scientists or have such small remaining populations that detection is rare. The FrogID app is actively working to close this gap, but it also underlines why the 33,000+ *Crinia signifera* records coexist with species recorded only once or twice in the entire dataset.

::: {.callout-tip}
**For your own exploration:** The latitude and longitude fields in `frogID_data` are rich but deliberately coarsened (10,000m uncertainty buffers) for sensitive species. Even so, mapping species range limits — particularly the northern boundary of temperate species or the southern edge of tropical ones — against the calling calendar would be a compelling follow-up.
:::
