---
title: "Tidy Tuesday: Repair Cafes Worldwide"
author: "Sean Thimons"
date: "2026-04-07"
description: "Exploring repair success rates and failure modes across product categories at repair cafes worldwide."
categories:
  - tidytuesday
  - R
  - sustainability
  - repair
---

```{r}
#| label: pseudo-hero-palette
#| echo: false
#| message: false
#| warning: false
#| fig-width: 14
#| fig-height: 2.5
#| dpi: 300
#| out-width: 100%

if (!requireNamespace("pak", quietly = TRUE)) {
  install.packages("pak", repos = "https://cloud.r-project.org")
}

for (pkg in c("ggplot2", "paletteer", "tibble")) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    pak::pkg_install(pkg)
  }
}

library(ggplot2)
library(paletteer)
library(tibble)

palette_name   <- "OskarSchlemmer"
palette_source <- "lisa"
palette_type   <- "discrete"
palette_id     <- paste0(palette_source, "::", palette_name)

hero_swatches <- if (palette_type %in% c("continuous", "diverging")) {
  as.character(paletteer::paletteer_c(palette_id, n = 5))
} else {
  rep_len(as.character(paletteer::paletteer_d(palette_id, n = 5)), 5)
}

swatch_df <- tibble(
  idx  = seq_along(hero_swatches),
  xmin = seq_along(hero_swatches) - 1,
  xmax = seq_along(hero_swatches),
  fill = hero_swatches
)

p_hero <- ggplot(swatch_df) +
  geom_rect(
    aes(xmin = xmin, xmax = xmax, ymin = 0, ymax = 1, fill = fill),
    color = NA,
    show.legend = FALSE
  ) +
  scale_fill_identity() +
  annotate(
    "text",
    x = 0,
    y = -0.22,
    hjust = 0,
    label = paste0("Palette: ", palette_name),
    fontface = "bold",
    size = 4.1
  ) +
  annotate(
    "text",
    x = 5,
    y = -0.22,
    hjust = 1,
    label = paste0("Source: ", palette_source),
    color = "grey35",
    size = 3.6
  ) +
  coord_cartesian(xlim = c(0, 5), ylim = c(-0.35, 1), clip = "off", expand = FALSE) +
  theme_void() +
  theme(plot.margin = margin(5.5, 0, 18, 0))

p_hero
```

# Preface

From [TidyTuesday repository](https://github.com/rfordatascience/tidytuesday/tree/main/data/2026/2026-04-07).

> Repair Cafes are free meeting places where visitors can bring their broken items to be repaired by skilled volunteers. This dataset comes from the Repair Cafe Foundation and covers repairs logged from hundreds of cafes worldwide, including the item category, repair outcome, and failure reasons when a repair could not be completed.

## Loading necessary packages

My handy booster pack that allows me to install (if needed) and load my usual and favorite packages, as well as some helpful functions.

```{r}
#| label: booster-pack
#| message: false
#| warning: false
#| code-fold: true

# Packages ----------------------------------------------------------------

{
  # Install pak if it's not already installed
  if (!requireNamespace("pak", quietly = TRUE)) {
    install.packages("pak", repos = "https://cloud.r-project.org")
  }

  # CRAN Packages ----
  install_booster_pack <- function(package, load = TRUE) {
    for (pkg in package) {
      if (!requireNamespace(pkg, quietly = TRUE)) {
        pak::pkg_install(pkg)
      }
      if (load) {
        library(pkg, character.only = TRUE)
      }
    }
  }

  core_packages <- c(
    'fs',
    'here',
    'janitor',
    'rio',
    'tidyverse',
    'skimr',
    'paletteer',
    'tidytuesdayR'
  )

  optional_packages <- c('ggtext', 'glue')

  install_booster_pack(package = unique(c(core_packages, optional_packages)), load = TRUE)
  rm(install_booster_pack, core_packages, optional_packages)

  # Custom Functions ----

  `%ni%` <- Negate(`%in%`)

  geometric_mean <- function(x) {
    exp(mean(log(x[x > 0]), na.rm = TRUE))
  }

  my_skim <- skim_with(
    numeric = sfl(
      n = length,
      min = ~ min(.x, na.rm = T),
      p25 = ~ stats::quantile(., probs = .25, na.rm = TRUE, names = FALSE),
      med = ~ median(.x, na.rm = T),
      p75 = ~ stats::quantile(., probs = .75, na.rm = TRUE, names = FALSE),
      max = ~ max(.x, na.rm = T),
      mean = ~ mean(.x, na.rm = T),
      geo_mean = ~ geometric_mean(.x),
      sd = ~ stats::sd(., na.rm = TRUE),
      hist = ~ inline_hist(., 5)
    ),
    append = FALSE
  )
}


```

# Load raw data from package

```{r}
#| label: raw-data-load
#| echo: true
#| output: false

raw <- tidytuesdayR::tt_load('2026-04-07')

repairs      <- raw$repairs
repairs_text <- raw$repairs_text

```

# Exploratory Data Analysis

The `my_skim()` function is a modified version of the `skimr::skim()` function that returns the number of missing data points (cells as `NA`) as well as the inverse (e.g.: number of rows that are *not* `NA`), the count, minimum, 25%, median, 75%, max, mean, geometric mean, and standard deviation. It also generates a little ASCII histogram. Neat!

## repairs

```{r}
#| label: eda-repairs
#| echo: true
#| output: true

my_skim(repairs)

# Check exact values in key categorical columns before any filtering
cat("\n--- repaired column values ---\n")
repairs |> count(repaired, sort = TRUE) |> print()

cat("\n--- category column values ---\n")
repairs |> count(category, sort = TRUE) |> print()

cat("\n--- country column values ---\n")
repairs |> count(country, sort = TRUE) |> print(n = 30)

```

## repairs_text

```{r}
#| label: eda-repairs-text
#| echo: true
#| output: true

my_skim(repairs_text)

cat("\n--- failure_reasons column values ---\n")
repairs_text |> count(failure_reasons, sort = TRUE) |> print()

```

The `repairs` table has ~179,000 repair records spanning hundreds of repair cafes worldwide. The dominant countries are the Netherlands (NL, ~91k records), Great Britain (GB, ~38k), and France (FR, ~20k), reflecting where the Repair Cafe movement originated. The `repaired` column uses "yes", "no", and "half" as its primary outcome values, with a small number of "ja" entries (Dutch for "yes"). The `repairability` score runs 0–10 and is missing for roughly a third of records.

# Repair Success by Product Category

The core question for any repair cafe visitor is: "What are my chances of walking out with a fixed item?" I look at repair outcomes across all product categories, normalizing for the mix of "yes", "half", and "no" results. The "half" outcome — a partial fix — is treated as a mid-point rather than a failure.

```{r}
#| label: prep-category-outcomes
#| echo: true
#| output: true
#| message: false
#| warning: false

# Normalize repaired values: treat "ja" (Dutch yes) as "yes"
repairs_clean <- repairs |>
  mutate(
    repaired_norm = case_when(
      repaired %in% c("yes", "ja") ~ "Repaired",
      repaired == "half"           ~ "Partial",
      repaired == "no"             ~ "Not repaired",
      TRUE                         ~ NA_character_
    )
  ) |>
  filter(!is.na(category), !is.na(repaired_norm))

# Verify we have data
stopifnot(nrow(repairs_clean) > 0)
cat("Rows after cleaning:", nrow(repairs_clean), "\n")

# Count outcomes per category
category_outcomes <- repairs_clean |>
  count(category, repaired_norm) |>
  group_by(category) |>
  mutate(
    total      = sum(n),
    pct        = n / total
  ) |>
  ungroup()

# Verify
stopifnot(nrow(category_outcomes) > 0)
cat("Categories x outcomes:\n")
print(category_outcomes)

# Order categories by % fully repaired
cat_order <- category_outcomes |>
  filter(repaired_norm == "Repaired") |>
  arrange(pct) |>
  pull(category)

category_outcomes <- category_outcomes |>
  mutate(category = factor(category, levels = cat_order))

```

```{r}
#| label: hero-plot-category-outcomes
#| message: false
#| warning: false
#| fig-width: 11
#| fig-height: 8
#| out-width: 100%

pal_colors <- as.character(paletteer::paletteer_d("lisa::OskarSchlemmer", n = 5))
# Assign outcome colors: Repaired = warm gold, Partial = mid blue-purple, Not repaired = dark red
outcome_colors <- c(
  "Repaired"     = pal_colors[3],   # DABD61 warm gold
  "Partial"      = pal_colors[2],   # 8785B2 muted purple
  "Not repaired" = pal_colors[5]    # BE3428 warm red
)

p <- ggplot(
  category_outcomes,
  aes(x = pct, y = category, fill = repaired_norm)
) +
  geom_col(width = 0.7, position = "stack") +
  scale_x_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0.02))
  ) +
  scale_fill_manual(
    values = outcome_colors,
    breaks = c("Repaired", "Partial", "Not repaired"),
    name   = "Outcome"
  ) +
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "grey40", linewidth = 0.5) +
  labs(
    title    = "Repair success rates at repair cafes worldwide",
    subtitle = "Share of items fully repaired, partially fixed, or not repaired — by product category",
    x        = "Share of repairs",
    y        = NULL,
    caption  = "Source: Repair Cafe Foundation via TidyTuesday 2026-04-07 | 'half' outcomes treated as partial fixes"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title       = element_text(face = "bold", size = 15),
    plot.subtitle    = element_text(color = "grey30", size = 11, margin = margin(b = 10)),
    plot.caption     = element_text(color = "grey50", size = 9, hjust = 0),
    legend.position  = "top",
    legend.title     = element_text(face = "bold"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    axis.text.y      = element_text(size = 11)
  )

p

```

```{r}
#| label: failure-reasons
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 6
#| out-width: 100%

# Join repairs with repairs_text to get failure reasons
repairs_with_text <- repairs |>
  inner_join(repairs_text |> select(repair_id, failure_reasons), by = "repair_id") |>
  filter(!is.na(failure_reasons), !is.na(category))

stopifnot(nrow(repairs_with_text) > 0)
cat("Rows with failure reasons:", nrow(repairs_with_text), "\n")

# Top failure reasons overall
failure_summary <- repairs_with_text |>
  count(failure_reasons, sort = TRUE) |>
  filter(failure_reasons != "NA") |>
  slice_max(n, n = 9) |>
  mutate(
    failure_reasons = str_wrap(failure_reasons, width = 35),
    failure_reasons = fct_reorder(failure_reasons, n)
  )

stopifnot(nrow(failure_summary) > 0)

pal_fail <- as.character(paletteer::paletteer_d("lisa::OskarSchlemmer", n = 5))

p2 <- ggplot(failure_summary, aes(x = n, y = failure_reasons, fill = n)) +
  geom_col(width = 0.7, show.legend = FALSE) +
  scale_fill_gradient(low = pal_fail[2], high = pal_fail[4]) +
  scale_x_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(
    title    = "Why do repairs fail at repair cafes?",
    subtitle = "Most common logged reasons when a repair could not be completed",
    x        = "Number of repair records",
    y        = NULL,
    caption  = "Source: Repair Cafe Foundation via TidyTuesday 2026-04-07"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title       = element_text(face = "bold", size = 15),
    plot.subtitle    = element_text(color = "grey30", size = 11, margin = margin(b = 10)),
    plot.caption     = element_text(color = "grey50", size = 9, hjust = 0),
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    axis.text.y      = element_text(size = 11)
  )

p2

```

# Final thoughts and takeaways

The repair cafe dataset paints an encouraging picture of the repair movement. Across roughly 179,000 logged repair attempts:

- **Textiles and non-electric tools** tend to have the highest full-repair rates. These categories often involve straightforward fixes — sewing a seam, sharpening a blade — that skilled volunteers can reliably complete within a session.

- **Bicycles and clocks/alarm clocks** also perform well. The mechanical nature of these items, combined with the specialized volunteers often present at repair cafes, makes successful repair relatively common.

- **Electronic and computer equipment** sits in the middle. While volunteers can fix many issues, component-level repairs require parts and equipment that aren't always on hand.

- The dominant failure modes are structural rather than skill-related: **"No way to fix the product"** (e.g., items that are sealed or glued shut) and **"Spare parts not available"** together account for the largest share of unsuccessful repairs. This points directly at a design and supply chain problem — products engineered without repairability in mind.

The repair cafe model is fundamentally optimistic: bring what's broken, and skilled community members will try their best. But the data reveals where that optimism runs into systemic barriers. The right-to-repair movement's emphasis on access to spare parts and service documentation maps directly onto the failure reasons most common in this dataset.