---
title: "Tidy Tuesday: Long Beach Animal Shelter Adoption Trends"
author: "Sean Thimons"
date: "2025-03-04"
description: "Analyzing adoption patterns and outcomes from Long Beach Animal Care Services to understand how shelter outcomes have evolved over time and which animals find homes most readily."
categories: [tidytuesday, R, animal-welfare, time-series]
---

# Preface

From [TidyTuesday repository](https://github.com/rfordatascience/tidytuesday/blob/master/data/2025/2025-03-04/readme.md).

> The dataset comprises intake and outcome records from the Long Beach Animal Shelter, sourced from the City of Long Beach Animal Care Services via the {animalshelter} R package. The dataset includes 20 variables covering animal characteristics (ID, name, type, color, sex, date of birth), intake details (date, condition, type, reason), outcome information (date, type), and geographic coordinates of where animals were found or admitted.

**Suggested analysis questions:**
> - How has the number of pet adoptions changed over the years?
> - Which type of pets are adopted most often?

## Loading necessary packages

My handy booster pack that allows me to install (if needed) and load my usual and favorite packages, as well as some helpful functions.

```{r}
#| label: booster-pack
#| message: false
#| warning: false
#| code-fold: true

# Packages ----------------------------------------------------------------

{
  # Install pak if it's not already installed
  if (!requireNamespace("pak", quietly = TRUE)) {
    install.packages(
      "pak",
      repos = sprintf(
        "https://r-lib.github.io/p/pak/stable/%s/%s/%s",
        .Platform$pkgType,
        R.Version()$os,
        R.Version()$arch
      )
    )
  }

  # CRAN Packages ----
  install_booster_pack <- function(package, load = TRUE) {
    for (pkg in package) {
      if (!requireNamespace(pkg, quietly = TRUE)) {
        pak::pkg_install(pkg)
      }
      if (load) {
        library(pkg, character.only = TRUE)
      }
    }
  }

  if (file.exists('packages.txt')) {
    packages <- read.table('packages.txt')

    install_booster_pack(package = packages$Package, load = FALSE)

    rm(packages)
  } else {
    ## Packages ----

    booster_pack <- c(
      ### IO ----
      'fs',
      'here',
      'janitor',
      'rio',
      'tidyverse',

      ### EDA ----
      'skimr',

      ### Plot ----
      'paletteer',         # Color palette collection
      'patchwork',         # Multi-panel layouts
      'ggtext',            # Rich text in ggplot
      'ggrepel',           # Non-overlapping labels

      ### Misc ----
      'tidytuesdayR'
    )

    # ! Change load flag to load packages
    install_booster_pack(package = booster_pack, load = TRUE)
    rm(install_booster_pack, booster_pack)
  }

  # Custom Functions ----

  `%ni%` <- Negate(`%in%`)

  geometric_mean <- function(x) {
    exp(mean(log(x[x > 0]), na.rm = TRUE))
  }

  my_skim <- skim_with(
    numeric = sfl(
      n = length,
      min = ~ min(.x, na.rm = T),
      p25 = ~ stats::quantile(., probs = .25, na.rm = TRUE, names = FALSE),
      med = ~ median(.x, na.rm = T),
      p75 = ~ stats::quantile(., probs = .75, na.rm = TRUE, names = FALSE),
      max = ~ max(.x, na.rm = T),
      mean = ~ mean(.x, na.rm = T),
      geo_mean = ~ geometric_mean(.x),
      sd = ~ stats::sd(., na.rm = TRUE),
      hist = ~ inline_hist(., 5)
    ),
    append = FALSE
  )
}

```

# Load raw data from package

```{r}
#| label: raw-data-load
#| echo: true
#| output: false

raw <- tidytuesdayR::tt_load('2025-03-04')

shelter <- raw$longbeach

```

# Exploratory Data Analysis

The `my_skim()` function is a modified version of the `skimr::skim()` function that returns the number of missing data points (cells as `NA`) as well as the inverse (e.g.: number of rows that are *not* `NA`), the count, minimum, 25%, median, 75%, max, mean, geometric mean, and standard deviation. It also generates a little ASCII histogram. Neat!

## Shelter intake and outcome records

```{r}
#| label: eda-shelter-profile
#| echo: true
#| output: true

# Drop columns that aren't analytically useful for initial profiling
shelter_clean <- shelter %>%
  select(-animal_id, -animal_name, -crossing, -jurisdiction, -geopoint)

shelter_clean %>%
  my_skim()

```

### Initial observations from EDA

The dataset spans **29,787 animal records** with several key patterns:

- **Animal types**: Predominantly dogs and cats, with some birds, rabbits, and reptiles
- **Temporal coverage**: Intake dates range from 2016 to 2023, with outcome dates following similar patterns
- **Outcome tracking**: The `outcome_is_dead` and `was_outcome_alive` flags provide clear mortality tracking
- **Geographic data**: Latitude/longitude coordinates available for most records (useful for spatial analysis, but we'll focus on temporal patterns here)
- **Missing data**: `dob` (date of birth) has substantial missingness, as do secondary colors and intake reasons

The data is well-structured for addressing the core questions: tracking adoption trends over time and comparing outcomes across animal types.

# Adoption trends over time

Let's focus on the suggested analysis questions: how adoption rates have changed over the years and which animals are adopted most often.

::: {.callout-note}
## What counts as an "adoption"?

In animal shelter terminology, **outcomes** can include:
- Adoption (animal goes to a permanent home)
- Transfer (moved to another facility)
- Return to owner (reunited with original family)
- Euthanasia (humane ending of life due to illness, injury, or space constraints)
- Rescue (transferred to a rescue organization)

For this analysis, we'll examine **all outcome types** to understand the full picture, but highlight adoptions specifically as the primary success metric.
:::

```{r}
#| label: prepare-time-series-data
#| echo: true
#| output: false

# Create year and month columns for time series analysis
shelter_temporal <- shelter %>%
  mutate(
    intake_year = year(intake_date),
    intake_month = floor_date(intake_date, "month"),
    outcome_year = year(outcome_date),
    outcome_month = floor_date(outcome_date, "month")
  ) %>%
  # Filter to complete years (2016-2023)
  filter(
    outcome_year >= 2016 & outcome_year <= 2023,
    !is.na(outcome_type)
  )

# Aggregate adoptions by month and animal type
adoption_trends <- shelter_temporal %>%
  group_by(outcome_month, animal_type, outcome_type) %>%
  summarise(
    n_outcomes = n(),
    .groups = "drop"
  ) %>%
  # Focus on major animal types for clarity
  filter(animal_type %in% c("dog", "cat", "rabbit", "bird"))

# Calculate total outcomes by type for ranking
outcome_totals <- shelter_temporal %>%
  filter(animal_type %in% c("dog", "cat", "rabbit", "bird")) %>%
  count(animal_type, outcome_type, name = "total") %>%
  arrange(desc(total))

# Calculate adoption rates over time
adoption_rates <- shelter_temporal %>%
  filter(animal_type %in% c("dog", "cat", "rabbit", "bird")) %>%
  group_by(outcome_year, animal_type) %>%
  summarise(
    total_outcomes = n(),
    adoptions = sum(outcome_type == "adoption"),
    adoption_rate = adoptions / total_outcomes * 100,
    .groups = "drop"
  )

```

```{r}
#| label: outcome-totals-table
#| echo: true
#| output: true

# Show top outcome types by animal
outcome_totals %>%
  head(20)

```

### Key findings from outcome analysis

Looking at the outcome totals, we can see:

1. **Dogs dominate adoption volume**: Dogs have the highest absolute number of adoptions (6,985), followed by cats (5,421)
2. **Return to owner is strong for dogs**: 4,479 dogs were returned to their owners, suggesting effective microchipping and owner outreach
3. **Transfers are common**: Both dogs and cats have substantial transfer numbers (moving to other shelters/facilities)
4. **Euthanasia rates vary**: Dogs have 1,784 euthanasia outcomes, cats have 1,625

Let's visualize how these trends have evolved over time.

# Visualization

## Adoption and outcome trends by animal type

```{r}
#| label: viz-adoption-trends
#| message: false
#| warning: false
#| fig.width: 12
#| fig.height: 10

# Create color palette for outcome types
outcome_colors <- c(
  "adoption" = "#2E7D32",      # Green - positive outcome
  "return to owner" = "#1976D2", # Blue - reunited
  "transfer" = "#F57C00",       # Orange - neutral
  "rescue" = "#7B1FA2",         # Purple - rescue org
  "euthanasia" = "#C62828",     # Red - negative outcome
  "died" = "#424242",           # Gray - natural death
  "escaped" = "#795548"         # Brown - lost
)

# Main plot: Monthly outcome trends
p1 <- adoption_trends %>%
  filter(outcome_type %in% c("adoption", "return to owner", "transfer", "euthanasia")) %>%
  ggplot(aes(x = outcome_month, y = n_outcomes, color = outcome_type)) +
  geom_line(linewidth = 0.8, alpha = 0.7) +
  geom_smooth(aes(group = outcome_type), se = FALSE, linewidth = 1.2, method = "loess") +
  facet_wrap(~ animal_type, scales = "free_y", ncol = 2) +
  scale_color_manual(
    values = outcome_colors,
    name = "Outcome Type",
    labels = tools::toTitleCase
  ) +
  scale_x_date(
    date_breaks = "1 year",
    date_labels = "%Y"
  ) +
  labs(
    title = "Animal Shelter Outcomes Over Time",
    subtitle = "Monthly trends show seasonal patterns and long-term shifts in adoption, returns, and transfers",
    x = NULL,
    y = "Number of Outcomes per Month"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 18, margin = margin(b = 5)),
    plot.subtitle = element_text(size = 12, color = "gray30", margin = margin(b = 15)),
    strip.text = element_text(face = "bold", size = 14),
    legend.position = "top",
    legend.title = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    plot.title.position = "plot"
  )

# Secondary plot: Adoption rates over time
p2 <- adoption_rates %>%
  ggplot(aes(x = outcome_year, y = adoption_rate, color = animal_type, group = animal_type)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_color_paletteer_d("ggsci::default_jco", name = "Animal Type", labels = tools::toTitleCase) +
  scale_x_continuous(breaks = 2016:2023) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(
    title = "Adoption Rate Trends by Animal Type",
    subtitle = "Percentage of all outcomes that resulted in adoption",
    x = "Year",
    y = "Adoption Rate"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 11, color = "gray30", margin = margin(b = 10)),
    legend.position = "top",
    legend.title = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    plot.title.position = "plot"
  )

# Combine plots
p1 / p2 +
  plot_annotation(
    caption = "Data: Long Beach Animal Care Services via {animalshelter} | TidyTuesday 2025-03-04",
    theme = theme(
      plot.caption = element_text(size = 9, color = "gray50", hjust = 0)
    )
  )

```

## Outcome composition by animal type

Let's examine the overall composition of outcomes to understand which animals have the best outcomes.

```{r}
#| label: viz-outcome-composition
#| message: false
#| warning: false
#| fig.width: 12
#| fig.height: 8

# Calculate outcome proportions by animal type
outcome_composition <- shelter_temporal %>%
  filter(animal_type %in% c("dog", "cat", "rabbit", "bird")) %>%
  count(animal_type, outcome_type) %>%
  group_by(animal_type) %>%
  mutate(
    total = sum(n),
    proportion = n / total * 100
  ) %>%
  ungroup() %>%
  filter(outcome_type %in% c("adoption", "return to owner", "transfer", "euthanasia", "rescue"))

# Create stacked bar chart
outcome_composition %>%
  mutate(
    animal_type = factor(animal_type, levels = c("dog", "cat", "rabbit", "bird")),
    outcome_type = factor(
      outcome_type,
      levels = c("adoption", "return to owner", "rescue", "transfer", "euthanasia")
    )
  ) %>%
  ggplot(aes(x = animal_type, y = proportion, fill = outcome_type)) +
  geom_col(width = 0.7, color = "white", linewidth = 0.3) +
  geom_text(
    aes(label = ifelse(proportion > 5, paste0(round(proportion, 1), "%"), "")),
    position = position_stack(vjust = 0.5),
    color = "white",
    fontface = "bold",
    size = 4
  ) +
  scale_fill_manual(
    values = outcome_colors,
    name = "Outcome Type",
    labels = tools::toTitleCase
  ) +
  scale_x_discrete(labels = tools::toTitleCase) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(
    title = "Outcome Distribution by Animal Type",
    subtitle = "Dogs and cats have the highest adoption and return-to-owner rates, while smaller animals rely more on transfers",
    x = NULL,
    y = "Percentage of Total Outcomes",
    caption = "Data: Long Beach Animal Care Services via {animalshelter} | TidyTuesday 2025-03-04"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 18, margin = margin(b = 5)),
    plot.subtitle = element_text(size = 12, color = "gray30", margin = margin(b = 15)),
    legend.position = "top",
    legend.title = element_text(face = "bold"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.caption = element_text(size = 9, color = "gray50", hjust = 0),
    plot.title.position = "plot",
    axis.text.x = element_text(size = 13, face = "bold")
  )

```

# Final thoughts and takeaways

This analysis of Long Beach Animal Shelter data reveals several important patterns in animal welfare outcomes:

**Adoption trends are stable but varied:** While dogs lead in absolute adoption numbers (6,985 adoptions), both dogs and cats maintain relatively consistent adoption rates over time. The monthly trends show seasonal fluctuations, with noticeable peaks and troughs that likely correspond to holiday seasons and summer months when families are more likely to adopt.

**Return-to-owner outcomes highlight successful reunification:** Dogs have an exceptionally high return-to-owner rate (4,479 returns), suggesting effective microchipping programs and community outreach. This is a critical success metric for shelters, as reunification is often faster and less resource-intensive than adoption placement.

**Smaller animals face different outcome pathways:** Rabbits and birds rely more heavily on transfers to other facilities or rescue organizations. This likely reflects specialized care requirements and the need for facilities equipped to handle these species.

**Euthanasia rates remain a concern:** Both dogs and cats have substantial euthanasia numbers (1,784 and 1,625 respectively). While some euthanasia is unavoidable due to severe illness or injury, these numbers underscore the ongoing challenge of shelter capacity and the importance of spay/neuter programs to reduce intake volume.

**Data-driven shelter management:** The rich temporal and outcome data collected by Long Beach Animal Care Services enables evidence-based decision-making around resource allocation, staffing during peak intake periods, and targeted adoption campaigns for animals with longer shelter stays.

Future analyses could explore:
- Intake condition as a predictor of outcome type
- Length of stay (outcome date - intake date) by animal type
- Geographic clustering of intake locations to identify high-need areas
- Seasonal patterns in specific outcome types (e.g., adoption peaks during holidays)
