---
title: "Tidy Tuesday: Coastal Ocean Temperature by Depth"
author: "Sean Thimons"
date: "2026-04-02"
description: "Exploring vertical temperature profiles in Lunenburg County's coastal waters."
categories:
  - tidytuesday
  - R
  - oceanography
  - climate
---

```{r}
#| label: pseudo-hero-palette
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 1.35
#| out-width: 100%

if (!requireNamespace("pak", quietly = TRUE)) {
  install.packages("pak", repos = "https://cloud.r-project.org")
}

for (pkg in c("ggplot2", "paletteer", "tibble")) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    pak::pkg_install(pkg)
  }
}

library(ggplot2)
library(paletteer)
library(tibble)

palette_name   <- "oslo"
palette_source <- "scico"
palette_type   <- "continuous"
palette_id     <- paste0(palette_source, "::", palette_name)

hero_swatches <- if (palette_type %in% c("continuous", "diverging")) {
  as.character(paletteer::paletteer_c(palette_id, n = 5))
} else {
  rep_len(as.character(paletteer::paletteer_d(palette_id, n = 5)), 5)
}

swatch_df <- tibble(
  idx  = seq_along(hero_swatches),
  xmin = seq_along(hero_swatches) - 1,
  xmax = seq_along(hero_swatches),
  fill = hero_swatches
)

p_hero <- ggplot(swatch_df) +
  geom_rect(
    aes(xmin = xmin, xmax = xmax, ymin = 0, ymax = 1, fill = fill),
    color = NA,
    show.legend = FALSE
  ) +
  scale_fill_identity() +
  annotate(
    "text",
    x = 0,
    y = -0.22,
    hjust = 0,
    label = paste0("Palette: ", palette_name),
    fontface = "bold",
    size = 4.1
  ) +
  annotate(
    "text",
    x = 5,
    y = -0.22,
    hjust = 1,
    label = paste0("Source: ", palette_source),
    color = "grey35",
    size = 3.6
  ) +
  coord_cartesian(xlim = c(0, 5), ylim = c(-0.35, 1), clip = "off", expand = FALSE) +
  theme_void() +
  theme(plot.margin = margin(5.5, 0, 18, 0))

p_hero
```

# Preface

From [TidyTuesday repository](https://github.com/rfordatascience/tidytuesday/blob/main/data/2026/2026-03-31/readme.md).

> This dataset explores the complex relationship between ocean depth and mean water temperature in coastal monitoring sites.

## Loading necessary packages

My handy booster pack that allows me to install (if needed) and load my usual and favorite packages, as well as some helpful functions.

```{r}
#| label: booster-pack
#| message: false
#| warning: false
#| code-fold: true

# Packages ----------------------------------------------------------------

{
  # Install pak if it's not already installed
  if (!requireNamespace("pak", quietly = TRUE)) {
    install.packages("pak", repos = "https://cloud.r-project.org")
  }

  # CRAN Packages ----
  install_booster_pack <- function(package, load = TRUE) {
    for (pkg in package) {
      if (!requireNamespace(pkg, quietly = TRUE)) {
        pak::pkg_install(pkg)
      }
      if (load) {
        library(pkg, character.only = TRUE)
      }
    }
  }

  core_packages <- c(
    'fs',
    'here',
    'janitor',
    'rio',
    'tidyverse',
    'skimr',
    'paletteer',
    'tidytuesdayR'
  )

  optional_packages <- c('gghighlight', 'ggthemes', 'patchwork')

  install_booster_pack(package = unique(c(core_packages, optional_packages)), load = TRUE)
  rm(install_booster_pack, core_packages, optional_packages)

  # Custom Functions ----

  `%ni%` <- Negate(`%in%`)

  geometric_mean <- function(x) {
    exp(mean(log(x[x > 0]), na.rm = TRUE))
  }

  my_skim <- skim_with(
    numeric = sfl(
      n = length,
      min = ~ min(.x, na.rm = T),
      p25 = ~ stats::quantile(., probs = .25, na.rm = TRUE, names = FALSE),
      med = ~ median(.x, na.rm = T),
      p75 = ~ stats::quantile(., probs = .75, na.rm = TRUE, names = FALSE),
      max = ~ max(.x, na.rm = T),
      mean = ~ mean(.x, na.rm = T),
      geo_mean = ~ geometric_mean(.x),
      sd = ~ stats::sd(., na.rm = TRUE),
      hist = ~ inline_hist(., 5)
    ),
    append = FALSE
  )
}
```

# Load raw data from package

```{r}
#| label: raw-data-load
#| echo: true
#| output: false

raw <- tidytuesdayR::tt_load('2026-03-31')

ocean_temp <- raw$ocean_temperature
deployments <- raw$ocean_temperature_deployments

```

# Exploratory Data Analysis

The `my_skim()` function is a modified version of the `skimr::skim()` function that returns the number of missing data points (cells as `NA`) as well as the inverse (e.g.: number of rows that are *not* `NA`), the count, minimum, 25%, median, 75%, max, mean, geometric mean, and standard deviation. It also generates a little ASCII histogram. Neat!

## Ocean Temperature Data

```{r}
#| label: eda-ocean-temp
#| echo: true
#| output: true

my_skim(ocean_temp)

```

The dataset contains vertical temperature measurements across multiple depths. We see a clear depth-temperature gradient.

# Seasonal Stability vs Depth

How does the variance (standard deviation) of temperature behave as we go deeper? We might expect deeper waters to be more thermally stable throughout the year.

```{r}
#| label: depth-stability-plot
#| message: false
#| warning: false
#| fig.width: 10
#| fig.height: 8

p <- ocean_temp %>%
  mutate(month = month(date, label = TRUE)) %>%
  group_by(sensor_depth_at_low_tide_m, month) %>%
  summarize(
    mean_temp = mean(mean_temperature_degree_c, na.rm = TRUE),
    sd_temp = mean(sd_temperature_degree_c, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = month, y = mean_temp, group = factor(sensor_depth_at_low_tide_m))) +
  geom_line(aes(color = sensor_depth_at_low_tide_m), linewidth = 1.2) +
  scale_color_paletteer_c("scico::oslo", direction = -1) +
  labs(
    title = "Mean Ocean Temperature Cycles by Sensor Depth",
    subtitle = "Deepest sensors (darker) show more stability and slightly warmer winter mins",
    x = "Month",
    y = "Mean Temperature (°C)",
    color = "Depth (m)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

p
```

# Final thoughts and takeaways

The data confirms that while the surface waters experience the most extreme swings, the deeper strata provide a more stable thermal environment, which is critical for local sea life during the transition seasons.
