---
title: "Tidy Tuesday: Near-Death Experiences (NDERF)"
author: "Sean Thimons"
date: "2026-07-21"
description: "The NDERF archive is best read as sensitive public self-report data, where classifications and AI-coded motifs summarize narratives without replacing them."
categories:
  - tidytuesday
  - health
  - narratives
  - archive
---


# Preface

From the [TidyTuesday repository](https://github.com/rfordatascience/tidytuesday/tree/main/data/2026/2026-07-21).

The NDERF file is sensitive data in the ordinary human sense, even when it is public and tabular. Each row points back to a narrated experience. I treated the dataset as a catalog of self-reports and labels, not as evidence for or against anyone's metaphysics.

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(tidytuesdayR)
library(scales)

raw <- tidytuesdayR::tt_load("2026-07-21")
nde_experiences <- raw$nde_experiences
```

## Coverage

| Measure | Value |
|---|---:|
| Rows | 589 |
| Countries/regions represented | 65 |
| Classifications | 8 |
| AI motif columns | 8 |

## Geography of the archive

```{r}
#| label: fig-countries
#| fig-cap: "Most common reported countries in the NDERF table."
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 5

nde_experiences %>%
  filter(!is.na(country), country != "") %>%
  count(country, sort = TRUE) %>%
  slice_head(n = 12) %>%
  mutate(country = fct_reorder(country, n)) %>%
  ggplot(aes(n, country)) +
  geom_col(fill = "#355070") +
  scale_x_continuous(labels = comma) +
  labs(x = "Reports", y = NULL) +
  theme_minimal(base_size = 12)
```

The country distribution is mostly an archive-access pattern. English-language web forms, internet access, and willingness to submit a narrative all sit between experience and row. That makes geographic interpretation tricky: this is where reports came from, not where near-death experiences "happen."

## Classifications

```{r}
#| label: fig-classes
#| fig-cap: "Experience classifications in the archive."
#| message: false
#| warning: false
#| fig-width: 7
#| fig-height: 4.5

nde_experiences %>%
  filter(!is.na(classification), classification != "") %>%
  count(classification, sort = TRUE) %>%
  mutate(classification = fct_reorder(classification, n)) %>%
  ggplot(aes(n, classification)) +
  geom_col(fill = "#6d597a") +
  scale_x_continuous(labels = comma) +
  labs(x = "Reports", y = NULL) +
  theme_minimal(base_size = 12)
```

The classification field is helpful because it acknowledges that the archive is not one homogeneous phenomenon. Near-death reports sit beside related categories, and analysis should preserve that distinction unless there is a clear reason to pool them.

## Coded motifs

```{r}
#| label: fig-ai-tags
#| fig-cap: "AI-coded motifs summarize rough narrative themes, not final readings."
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 5

nde_experiences %>%
  select(starts_with("ai_")) %>%
  summarise(across(everything(), ~ sum(.x, na.rm = TRUE))) %>%
  pivot_longer(everything(), names_to = "motif", values_to = "reports") %>%
  mutate(motif = str_remove(motif, "^ai_"), motif = str_replace_all(motif, "_", " "), motif = fct_reorder(motif, reports)) %>%
  ggplot(aes(reports, motif)) +
  # A lollipop preserves the honest global ranking without implying stable subgroup rates.
  geom_segment(aes(x = 0, xend = reports, yend = motif), color = "#dec4cc", linewidth = 0.8) +
  geom_point(color = "#b56576", size = 3.2) +
  geom_text(aes(label = comma(reports)), hjust = -0.5, size = 3.3) +
  scale_x_continuous(labels = comma, expand = expansion(mult = c(0, 0.12))) +
  labs(x = "Reports tagged TRUE", y = NULL) +
  theme_minimal(base_size = 12)
```

The AI motif columns are useful as a rough index, not as a final reading. They invite questions about out-of-body experience, unity, clinical details, ESP claims, and other themes, but the labels should be checked against source narratives before any strong claim.

## Takeaway

This is a dataset where restraint is part of the method. Counts can summarize the archive, but they cannot carry the meaning of the stories alone. The right first pass is provenance-aware: who submitted, how the record was classified, and which computational labels are strong enough to support follow-up reading.
