---
title: "Tidy Tuesday: Many Penguins"
author: "Sean Thimons"
date: "2026-07-14"
description: "A compact AVONET penguin trait table uses genus, sex labels, and morphology measurements to show how small biological datasets still carry structure."
categories:
  - tidytuesday
  - penguins
  - biology
  - morphology
---


# Preface

From the [TidyTuesday repository](https://github.com/rfordatascience/tidytuesday/tree/main/data/2026/2026-07-14).

The `many_penguins` file is a small morphology slice from AVONET. It is compact enough to understand by inspection, but it carries the classic biological-data warning: the tidy table is not a sampling design. Species, genus, and sex labels are measurements and metadata, not guarantees of balanced coverage.

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(tidytuesdayR)

raw <- tidytuesdayR::tt_load("2026-07-14")
many_penguins <- raw$many_penguins
```

## Coverage

| Measure | Value |
|---|---:|
| Rows | 93 |
| Species | 18 |
| Genera | 6 |
| Usable beak/wing pairs | 79 |

## Taxonomic shape

```{r}
#| label: fig-genus
#| fig-cap: "Penguin morphology records by genus."
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 4.5

many_penguins %>%
  count(genus, sort = TRUE) %>%
  mutate(genus = fct_reorder(genus, n)) %>%
  ggplot(aes(n, genus)) +
  geom_col(fill = "#1f6f8b") +
  labs(x = "Records", y = NULL) +
  theme_minimal(base_size = 12)
```

The genus counts show the smallness of the file. This is not a citizen-science flood of repeated observations; it is a trait table. The unit of attention is the species/sex morphology record, so aggregation choices matter more than row count.

## Sex labels

```{r}
#| label: fig-sex
#| fig-cap: "Sex labels preserve unknown or unsexed records rather than hiding them."
#| message: false
#| warning: false
#| fig-width: 6
#| fig-height: 4

many_penguins %>%
  count(sex, sort = TRUE) %>%
  mutate(sex = fct_reorder(sex, n)) %>%
  ggplot(aes(n, sex, fill = sex)) +
  geom_col(width = 0.72, show.legend = FALSE) +
  geom_text(aes(label = n), hjust = -0.25, size = 4) +
  scale_fill_manual(values = c("F" = "#b56576", "M" = "#355070", "U" = "#6d597a")) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(x = "Records", y = NULL, title = "Unknown labels are visible evidence, not cleanup debris") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

The `U` category is not a nuisance to delete. Unknown or unsexed records are part of the evidence trail. Dropping them would make the table look cleaner while silently changing the biological coverage.

## Shape space

```{r}
#| label: fig-beak-wing
#| fig-cap: "Culmen length and wing length sketch a quick morphology map."
#| message: false
#| warning: false
#| fig-width: 7
#| fig-height: 5

many_penguins %>%
  filter(!is.na(beak.length_culmen), !is.na(wing.length)) %>%
  ggplot(aes(beak.length_culmen, wing.length, color = genus)) +
  geom_point(size = 2.4, alpha = 0.85) +
  labs(x = "Culmen length", y = "Wing length", color = "Genus") +
  theme_minimal(base_size = 12)
```

The beak-wing scatter is a quick morphology map. It shows separation in body plan without pretending that two measurements summarize a bird. As a first pass, though, it catches what makes penguins fun data objects: they are familiar animals with enough morphological spread to make simple plots legible.

## Takeaway

This week is a good antidote to overbuilt dashboards. A few rows, clear trait columns, and a modest scatterplot are enough to ask better questions: Which genera are morphologically distinct? Which sex labels are comparable? Which measurements scale together, and which describe different ecological constraints?

