---
title: "Tidy Tuesday: The 2026 Winter Olympics"
author: "Sean Thimons"
date: "2026-02-12"
description: "Mapping the geographic pulse of the Milano-Cortina 2026 Winter Olympics — how 1,866 events spread across 13 venues and 6 host cities over 19 days."
categories: [tidytuesday, R, sports, olympics, geography]
---

# Preface

From [TidyTuesday repository](https://github.com/rfordatascience/tidytuesday/blob/main/data/2026/2026-02-10/readme.md).

> This week's dataset focuses on the Milano-Cortina 2026 Winter Olympics schedule. The data contains 1,866 Olympic events including competition and training sessions across 16 winter sports disciplines. The schedule spans February 4–22, 2026 and features timezone conversions, venue details, and medal event classifications.

## Loading necessary packages

My handy booster pack that allows me to install (if needed) and load my usual and favorite packages, as well as some helpful functions.

```{r}
#| label: booster-pack
#| message: false
#| warning: false
#| code-fold: true

# Packages ----------------------------------------------------------------

{
  # Install pak if it's not already installed
  if (!requireNamespace("pak", quietly = TRUE)) {
    install.packages(
      "pak",
      repos = sprintf(
        "https://r-lib.github.io/p/pak/stable/%s/%s/%s",
        .Platform$pkgType,
        R.Version()$os,
        R.Version()$arch
      )
    )
  }

  # CRAN Packages ----
  install_booster_pack <- function(package, load = TRUE) {
    for (pkg in package) {
      if (!requireNamespace(pkg, quietly = TRUE)) {
        pak::pkg_install(pkg)
      }
      if (load) {
        library(pkg, character.only = TRUE)
      }
    }
  }

  if (file.exists('packages.txt')) {
    packages <- read.table('packages.txt')

    install_booster_pack(package = packages$Package, load = FALSE)

    rm(packages)
  } else {
    ## Packages ----

    booster_pack <- c(
      ### IO ----
      'fs',
      'here',
      'janitor',
      'rio',
      'tidyverse',

      ### EDA ----
      'skimr',

      ### Plot ----
      'camcorder',
      'paletteer',
      'patchwork',
      'ggtext',
      'ggrepel',

      ### Reporting ----
      'gt',

      ### Misc ----
      'tidytuesdayR'
    )

    # ! Change load flag to load packages
    install_booster_pack(package = booster_pack, load = TRUE)
    rm(install_booster_pack, booster_pack)
  }

  # Custom Functions ----

  `%ni%` <- Negate(`%in%`)

  geometric_mean <- function(x) {
    exp(mean(log(x[x > 0]), na.rm = TRUE))
  }

  my_skim <- skim_with(
    numeric = sfl(
      n = length,
      min = ~ min(.x, na.rm = T),
      p25 = ~ stats::quantile(., probs = .25, na.rm = TRUE, names = FALSE),
      med = ~ median(.x, na.rm = T),
      p75 = ~ stats::quantile(., probs = .75, na.rm = TRUE, names = FALSE),
      max = ~ max(.x, na.rm = T),
      mean = ~ mean(.x, na.rm = T),
      geo_mean = ~ geometric_mean(.x),
      sd = ~ stats::sd(., na.rm = TRUE),
      hist = ~ inline_hist(., 5)
    ),
    append = FALSE
  )
}


```

# Load raw data from package

```{r}
#| label: raw-data-load
#| echo: true
#| output: false

raw <- tidytuesdayR::tt_load('2026-02-10')

schedule <- raw$schedule %>%
  clean_names() %>%
  mutate(
    date = as.Date(date),
    is_medal_event = as.logical(is_medal_event),
    is_training = as.logical(is_training),
    estimated_start = as.logical(estimated_start),
    start_datetime_local = as.POSIXct(start_datetime_local),
    end_datetime_local = as.POSIXct(end_datetime_local),
    start_datetime_utc = as.POSIXct(start_datetime_utc, tz = "UTC"),
    end_datetime_utc = as.POSIXct(end_datetime_utc, tz = "UTC")
  )

```

# Exploratory Data Analysis

The `my_skim()` function is a modified version of the `skimr::skim()` function that returns the number of missing data points (cells as `NA`) as well as the inverse (e.g.: number of rows that are *not* `NA`), the count, minimum, 25%, median, 75%, max, mean, geometric mean, and standard deviation. It also generates a little ASCII histogram. Neat!

## Schedule overview

```{r}
#| label: eda-schedule
#| echo: true
#| output: true

schedule %>%
  select(
    -session_code,
    -event_code,
    -discipline_code,
    -venue_code,
    -venue_slug,
    -location_code,
    -start_datetime_utc,
    -end_datetime_utc
  ) %>%
  my_skim()

```

The schedule contains `r nrow(schedule)` total events across `r n_distinct(schedule$discipline_name)` disciplines and `r n_distinct(schedule$venue_name)` venues. Events span from `r min(schedule$date)` to `r max(schedule$date)` — a 19-day window that includes both training and competition sessions.

Key observations from the skim:

- **Training vs. competition**: `r sum(schedule$is_training)` training sessions and `r sum(!schedule$is_training)` competition events
- **Medal events**: `r sum(schedule$is_medal_event)` medal sessions out of the full schedule
- **Venue concentration**: Some venues host hundreds of events (Cortina Curling Olympic Stadium alone accounts for a huge share), while others are more specialized

## Host city extraction

The 2026 Games are uniquely distributed across northern Italy. Let's extract the host city from each venue name and see how the Games are geographically structured.

```{r}
#| label: eda-cities
#| echo: true
#| output: true

schedule <- schedule %>%
  mutate(
    host_city = case_when(
      str_detect(venue_name, "^Milano") ~ "Milano",
      str_detect(venue_name, "^Cortina") ~ "Cortina d'Ampezzo",
      str_detect(venue_name, "^Livigno") ~ "Livigno",
      str_detect(venue_name, "^Predazzo") ~ "Predazzo",
      str_detect(venue_name, "^Tesero") ~ "Tesero",
      str_detect(venue_name, "^Anterselva") ~ "Anterselva",
      str_detect(venue_name, "^Stelvio") ~ "Bormio",
      str_detect(venue_name, "^Tofane") ~ "Cortina d'Ampezzo",
      TRUE ~ "Other"
    )
  )

schedule %>%
  filter(!is_training) %>%
  count(host_city, sort = TRUE) %>%
  mutate(pct = scales::percent(n / sum(n))) %>%
  gt() %>%
  cols_label(
    host_city = "Host City",
    n = "Competition Events",
    pct = "Share"
  ) %>%
  tab_header(title = "Competition events by host city")

```

::: {.callout-note}
## A distributed Games

Unlike many Winter Olympics concentrated in a single mountain region, Milano-Cortina 2026 spreads across the Italian Alps and into the Po Valley. Milano hosts the ice sports (figure skating, speed skating, short track, hockey), while the mountain towns — Cortina, Livigno, Bormio — handle the snow and sliding disciplines. This geographic spread is a defining feature of these Games.
:::

# Venue Geography Analysis

## Disciplines per venue

Each venue is purpose-built or adapted for specific sports. Let's map which disciplines live where.

```{r}
#| label: venue-discipline-map
#| echo: true
#| output: true

venue_disciplines <- schedule %>%
  filter(!is_training) %>%
  count(venue_name, host_city, discipline_name, sort = TRUE)

venue_disciplines %>%
  group_by(venue_name, host_city) %>%
  summarise(
    disciplines = paste(discipline_name, collapse = ", "),
    total_events = sum(n),
    .groups = "drop"
  ) %>%
  arrange(desc(total_events)) %>%
  gt() %>%
  cols_label(
    venue_name = "Venue",
    host_city = "City",
    disciplines = "Disciplines",
    total_events = "Events"
  ) %>%
  tab_header(title = "Venue utilization and discipline mapping")

```

## Medal density by venue

Not all events are created equal. Some venues host dozens of rounds leading to a handful of medal moments; others are medal-dense. The **medal density** — the share of competition events that award medals — reveals which venues are built for the climactic moments.

```{r}
#| label: medal-density
#| echo: true
#| output: true

venue_medal_density <- schedule %>%
  filter(!is_training) %>%
  group_by(venue_name, host_city) %>%
  summarise(
    total_events = n(),
    medal_events = sum(is_medal_event),
    medal_density = medal_events / total_events,
    .groups = "drop"
  ) %>%
  arrange(desc(medal_density))

venue_medal_density %>%
  mutate(medal_density = scales::percent(medal_density, accuracy = 0.1)) %>%
  gt() %>%
  cols_label(
    venue_name = "Venue",
    host_city = "City",
    total_events = "Total Events",
    medal_events = "Medal Events",
    medal_density = "Medal Density"
  ) %>%
  tab_header(title = "Medal density across Olympic venues")

```

::: {.callout-important}
## Curling: the endurance sport of scheduling

Cortina Curling Olympic Stadium hosts a staggering 436 competition events — more than double any other venue — yet produces only 8 medal sessions (a medal density under 2%). Curling's round-robin format demands this marathon of matches, making it the most venue-intensive discipline by far.
:::

## Daily venue activity

```{r}
#| label: daily-venue-prep
#| echo: true
#| output: true

daily_venue <- schedule %>%
  filter(!is_training) %>%
  count(date, venue_name, host_city, name = "events")

# Create ordered factor for venues grouped by city
venue_order <- schedule %>%
  filter(!is_training) %>%
  mutate(
    city_order = factor(
      host_city,
      levels = c(
        "Milano",
        "Cortina d'Ampezzo",
        "Livigno",
        "Bormio",
        "Predazzo",
        "Tesero",
        "Anterselva"
      )
    )
  ) %>%
  arrange(city_order, venue_name) %>%
  pull(venue_name) %>%
  unique()

daily_venue <- daily_venue %>%
  mutate(venue_name = factor(venue_name, levels = rev(venue_order)))

```

```{r}
#| label: camcorder
#| warning: false

if (!dir.exists(here::here('posts', '2026-02-12', 'output'))) {
  dir.create(here::here('posts', '2026-02-12', 'output'), recursive = TRUE)
}

gg_record(
  here::here('posts', '2026-02-12', 'output'),
  device = "png",
  width = 10,
  height = 7,
  units = "in",
  dpi = 320
)


```

```{r}
#| label: hero-heatmap
#| message: false
#| warning: false
#| fig.width: 10
#| fig.height: 8

# Olympic-inspired color palette
olympic_gradient <- c("#FFFFFF", "#0085C7", "#00247D")

# Key ceremony dates
ceremonies <- tibble(
  date = as.Date(c("2026-02-06", "2026-02-22")),
  label = c("Opening\nCeremony", "Closing\nCeremony")
)

# City labels for venue grouping
city_labels <- daily_venue %>%
  distinct(venue_name, host_city) %>%
  mutate(
    venue_name = factor(venue_name, levels = levels(daily_venue$venue_name))
  ) %>%
  group_by(host_city) %>%
  summarise(
    y_pos = mean(as.numeric(venue_name)),
    .groups = "drop"
  )

# Build the heatmap
p <- ggplot(daily_venue, aes(x = date, y = venue_name, fill = events)) +
  geom_tile(color = "white", linewidth = 0.4) +
  scale_fill_gradient2(
    low = "#FFFFFF",
    mid = "#0085C7",
    high = "#00247D",
    midpoint = 15,
    name = "Events",
    na.value = "#F5F5F5"
  ) +
  scale_x_date(
    date_breaks = "2 days",
    date_labels = "%b %d",
    expand = expansion(mult = c(0.01, 0.05))
  ) +
  # Opening & closing ceremony markers
  geom_vline(
    data = ceremonies,
    aes(xintercept = date),
    linetype = "dashed",
    color = "#D4AF37",
    linewidth = 0.6
  ) +
  geom_label(
    data = ceremonies,
    aes(x = date, y = 0.5, label = label),
    inherit.aes = FALSE,
    fill = "#D4AF37",
    color = "white",
    fontface = "bold",
    size = 2.5,
    label.padding = unit(0.2, "lines"),
    label.size = 0
  ) +
  # City group annotations on the right
  annotate(
    "text",
    x = max(daily_venue$date) + 1.5,
    y = city_labels$y_pos,
    label = city_labels$host_city,
    hjust = 0,
    fontface = "italic",
    size = 3,
    color = "#555555"
  ) +
  coord_cartesian(clip = "off") +
  labs(
    title = "The Geographic Pulse of Milano-Cortina 2026",
    subtitle = "Daily competition events across 13 Olympic venues, grouped by host city",
    x = NULL,
    y = NULL,
    caption = "Source: TidyTuesday 2026-02-10 · Milano-Cortina 2026 Winter Olympics Schedule"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 16, color = "#00247D"),
    plot.subtitle = element_text(
      size = 11,
      color = "#555555",
      margin = margin(b = 12)
    ),
    plot.caption = element_text(size = 8, color = "#999999", hjust = 0),
    axis.text.y = element_text(size = 8),
    axis.text.x = element_text(size = 8, angle = 45, hjust = 1),
    legend.position = "top",
    legend.title = element_text(size = 9),
    legend.key.width = unit(1.5, "cm"),
    panel.grid = element_blank(),
    plot.margin = margin(10, 80, 10, 10)
  )

p

```

## Venue activity over time by city

```{r}
#| label: city-timeseries
#| message: false
#| warning: false
#| fig.width: 10
#| fig.height: 6

city_daily <- schedule %>%
  filter(!is_training) %>%
  count(date, host_city, name = "events")

city_colors <- c(
  "Milano" = "#0085C7",
  "Cortina d'Ampezzo" = "#D4AF37",
  "Livigno" = "#009F3D",
  "Bormio" = "#EE334E",
  "Predazzo" = "#F4C300",
  "Tesero" = "#FF8C00",
  "Anterselva" = "#8B4513"
)

ggplot(city_daily, aes(x = date, y = events, color = host_city)) +
  geom_line(linewidth = 1, alpha = 0.8) +
  geom_point(size = 2) +
  scale_color_manual(values = city_colors, name = "Host City") +
  scale_x_date(date_breaks = "2 days", date_labels = "%b %d") +
  labs(
    title = "Daily competition events by host city",
    subtitle = "Milano sustains the highest event load throughout the Games",
    x = NULL,
    y = "Number of events",
    caption = "Source: TidyTuesday 2026-02-10"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14, color = "#00247D"),
    plot.subtitle = element_text(size = 10, color = "#555555"),
    plot.caption = element_text(size = 8, color = "#999999"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )

```

# Final thoughts and takeaways

The Milano-Cortina 2026 Winter Olympics are architecturally unique: rather than clustering events in a single alpine basin, these Games sprawl across the Italian Alps and down into the Po Valley. This analysis reveals several patterns in that geographic design:

**Milano is the engine room.** With four venues hosting ice sports — figure skating, short track, speed skating, and two hockey arenas — Milano sustains the highest daily event count throughout the Games. The city's venues are active nearly every day from opening to closing ceremony.

**Cortina's curling dominance is deceptive.** The Cortina Curling Olympic Stadium generates by far the most events of any venue, but its medal density is vanishingly low. The round-robin format means hundreds of sessions build to just a handful of medal moments — a sharp contrast to venues like the speed skating stadium or biathlon arena where a larger share of events produce medals.

**Mountain venues are specialists.** The alpine and Nordic venues — Bormio, Predazzo, Tesero, Anterselva — each host one or two disciplines and operate on tighter, more concentrated schedules. Their activity windows are shorter but intense.

**The mid-Games peak is universal.** Across nearly every venue, event density peaks around February 13–16, suggesting a deliberate scheduling strategy that builds momentum through the first week and then ramps into a climactic stretch before the closing ceremonies.

This geographic distribution presents both logistical challenges (athlete transport, spectator travel between cities separated by hours of mountain driving) and opportunities (multiple cities share the economic and cultural benefits of hosting). Whether this distributed model becomes a template for future Games remains to be seen.
