---
title: "Tidy Tuesday: Extreme Weather Attribution Studies"
author: "Sean Thimons"
date: "2025-08-12"
description: "Profiling 744 climate-attribution studies to ask which extreme weather types most consistently bear a human fingerprint — and where the science still hedges."
categories:
  - tidytuesday
  - R
  - climate
  - attribution
  - extreme-weather
---

```{r}
#| label: pseudo-hero-palette
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 1.35
#| out-width: 100%

if (!requireNamespace("pak", quietly = TRUE)) {
  install.packages("pak", repos = "https://cloud.r-project.org")
}

for (pkg in c("ggplot2", "paletteer", "tibble", "PNWColors")) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    pak::pkg_install(pkg)
  }
}

library(ggplot2)
library(paletteer)
library(tibble)

palette_name   <- "Sailboat"
palette_source <- "PNWColors"
palette_type   <- "discrete"
palette_id     <- paste0(palette_source, "::", palette_name)

hero_swatches <- if (palette_type %in% c("continuous", "diverging")) {
  as.character(paletteer::paletteer_c(palette_id, n = 5))
} else {
  rep_len(as.character(paletteer::paletteer_d(palette_id, n = 5)), 5)
}

swatch_df <- tibble(
  idx  = seq_along(hero_swatches),
  xmin = seq_along(hero_swatches) - 1,
  xmax = seq_along(hero_swatches),
  fill = hero_swatches
)

p_hero <- ggplot(swatch_df) +
  geom_rect(
    aes(xmin = xmin, xmax = xmax, ymin = 0, ymax = 1, fill = fill),
    color = NA,
    show.legend = FALSE
  ) +
  scale_fill_identity() +
  annotate(
    "text",
    x = 0,
    y = -0.22,
    hjust = 0,
    label = paste0("Palette: ", palette_name),
    fontface = "bold",
    size = 4.1
  ) +
  annotate(
    "text",
    x = 5,
    y = -0.22,
    hjust = 1,
    label = paste0("Source: ", palette_source),
    color = "grey35",
    size = 3.6
  ) +
  coord_cartesian(xlim = c(0, 5), ylim = c(-0.35, 1), clip = "off", expand = FALSE) +
  theme_void() +
  theme(plot.margin = margin(5.5, 0, 18, 0))

p_hero
```

# Preface

From the [TidyTuesday repository](https://github.com/rfordatascience/tidytuesday/tree/main/data/2025/2025-08-12).

> This week's dataset documents scientific studies analyzing how climate change affects extreme weather events globally. The data originates from Carbon Brief's interactive mapping project ("Mapped: How climate change affects extreme weather around the world") and was last updated in November 2024. Individual studies covering multiple events or locations were separated into distinct entries where applicable.

Extreme-event attribution is a relatively young branch of climate science. It asks a deceptively simple question: did a warming climate make *this specific* heatwave / hurricane / drought worse, or more likely? The Carbon Brief catalogue is the most complete public-facing index of the answers we have so far. With 744 study rows in hand, the obvious move is to ask the data not just *how many* studies exist, but *what verdicts they tend to deliver* — and whether those verdicts cluster systematically by event type.

## Loading necessary packages

```{r}
#| label: booster-pack
#| message: false
#| warning: false
#| code-fold: true

{
  if (!requireNamespace("pak", quietly = TRUE)) {
    install.packages("pak", repos = "https://cloud.r-project.org")
  }

  install_booster_pack <- function(package, load = TRUE) {
    for (pkg in package) {
      if (!requireNamespace(pkg, quietly = TRUE)) {
        pak::pkg_install(pkg)
      }
      if (load) {
        library(pkg, character.only = TRUE)
      }
    }
  }

  core_packages <- c(
    'fs',
    'here',
    'janitor',
    'rio',
    'tidyverse',
    'skimr',
    'paletteer',
    'tidytuesdayR'
  )

  optional_packages <- c(
    'ggtext',
    'ggrepel',
    'scales',
    'PNWColors'
  )

  install_booster_pack(package = unique(c(core_packages, optional_packages)), load = TRUE)
  rm(install_booster_pack, core_packages, optional_packages)

  `%ni%` <- Negate(`%in%`)

  geometric_mean <- function(x) {
    exp(mean(log(x[x > 0]), na.rm = TRUE))
  }

  my_skim <- skim_with(
    numeric = sfl(
      n = length,
      min = ~ min(.x, na.rm = T),
      p25 = ~ stats::quantile(., probs = .25, na.rm = TRUE, names = FALSE),
      med = ~ median(.x, na.rm = T),
      p75 = ~ stats::quantile(., probs = .75, na.rm = TRUE, names = FALSE),
      max = ~ max(.x, na.rm = T),
      mean = ~ mean(.x, na.rm = T),
      geo_mean = ~ geometric_mean(.x),
      sd = ~ stats::sd(., na.rm = TRUE),
      hist = ~ inline_hist(., 5)
    ),
    append = FALSE
  )
}
```

# Load raw data from package

```{r}
#| label: raw-data-load
#| echo: true
#| output: false

raw <- tidytuesdayR::tt_load('2025-08-12')

studies <- raw$attribution_studies
```

# Exploratory Data Analysis

## attribution_studies

Free-text fields (`event_name`, `event_period`, `summary_statement`, `citation`, `link`) are dropped from the skim — they're useful for context but not for distributional profiling.

```{r}
#| label: eda-attribution-studies
#| echo: true
#| output: true

studies %>%
  select(-event_name, -event_period, -event_year,
         -summary_statement, -citation, -link, -iso_country_code) %>%
  my_skim()
```

```{r}
#| label: eda-categoricals
#| echo: true
#| output: true

studies %>% count(classification, sort = TRUE)
studies %>% count(event_type, sort = TRUE)
studies %>% count(study_focus, sort = TRUE)
studies %>% count(rapid_study, sort = TRUE)
studies %>% count(cb_region, sort = TRUE)
```

A few things jump out:

- **Verdicts skew heavily toward attribution.** 554 of 744 studies (~75%) conclude the event was more severe or more likely due to human influence. Roughly 9% find no discernible influence, 9% find the opposite (decreased severity), and 7% are inconclusive. Publication bias is real — null results are harder to publish — but the imbalance is still striking.
- **Heat dominates.** 207 studies (28%) target heat events, with rain/flooding (177) and drought (106) close behind. Together these three account for two-thirds of the literature.
- **The field is young and accelerating.** 91 studies were published in 2024 alone; only 6 exist for the entire 2004–2010 window. Rapid attribution (results within days of an event) is a recent invention — 103 of 744 studies are tagged as rapid.
- **Coverage is uneven geographically.** Europe, East/Southeast Asia, and Northern America together hold 60% of studies; Antarctica, Oceania, and Northern Africa/West Asia are barely represented.

# The attribution fingerprint by event type

The most interesting question the data can answer isn't *how many* studies exist per event type — that just tells us where the funding and attention flow. The sharper question is: **conditional on a study being done, how often does it conclude human influence made the event worse?** That ratio is the "attribution fingerprint" of an event class, and it varies meaningfully across the catalogue.

::: {.callout-note}
## What the verdicts mean

Carbon Brief collapses every study's headline finding into one of four categories. **"More severe or more likely"** is the canonical attribution result. **"No discernible influence"** means the analysis ran but couldn't isolate a climate signal. **"Decrease, less severe or less likely"** is rarer and often shows up for cold-snap events. **"Insufficient data/inconclusive"** flags studies that hit methodological limits.
:::

```{r}
#| label: build-fingerprint
#| echo: true
#| output: true

class_levels <- c(
  "More severe or more likely to occur",
  "Insufficient data/inconclusive",
  "No discernible human influence",
  "Decrease, less severe or less likely to occur"
)

class_short <- c(
  "More severe or more likely to occur"          = "More severe / likely",
  "Insufficient data/inconclusive"               = "Inconclusive",
  "No discernible human influence"               = "No human influence",
  "Decrease, less severe or less likely to occur" = "Less severe / likely"
)

fingerprint <- studies %>%
  filter(!is.na(classification), !is.na(event_type)) %>%
  mutate(
    classification = factor(classification, levels = class_levels),
    class_short    = factor(class_short[as.character(classification)],
                            levels = unname(class_short))
  ) %>%
  count(event_type, class_short, name = "n") %>%
  group_by(event_type) %>%
  mutate(
    total = sum(n),
    pct   = n / total
  ) %>%
  ungroup()

event_order <- fingerprint %>%
  distinct(event_type, total) %>%
  arrange(desc(total)) %>%
  pull(event_type)

fingerprint <- fingerprint %>%
  mutate(event_type = factor(event_type, levels = rev(event_order)))

cat(sprintf("fingerprint: %d rows across %d event types\n",
            nrow(fingerprint), length(event_order)))
stopifnot("Fingerprint data is empty — check filter values" = nrow(fingerprint) > 0)
stopifnot("Proportions don't sum to 1 per event type" =
            all(abs(
              fingerprint %>% group_by(event_type) %>%
                summarise(s = sum(pct), .groups = "drop") %>% pull(s) - 1
            ) < 1e-9))

fingerprint %>%
  select(event_type, class_short, n, pct, total) %>%
  arrange(event_type, class_short) %>%
  print(n = Inf)
```

The table makes the variation concrete. Heat events are attributed to climate change in **89%** of studies. Cold/snow/ice events flip the script: only ~37% are "more severe", while a striking share land in the "decrease / less severe" bucket — physically intuitive, because a warming world produces fewer extreme cold snaps. Storms and oceans sit in the middle, reflecting how messy the dynamic-atmosphere signal is at the scale of individual storms.

```{r}
#| label: percent-more-severe
#| echo: true
#| output: true

pct_more_severe <- fingerprint %>%
  group_by(event_type) %>%
  summarise(
    total        = first(total),
    n_more       = sum(n[class_short == "More severe / likely"]),
    pct_more     = n_more / total,
    .groups = "drop"
  ) %>%
  arrange(desc(pct_more))

pct_more_severe
```

# Visualization

The story is about the *shape* of evidence inside each event class, so a 100%-stacked horizontal bar chart is the right tool: each row sums to 1, the dominant verdict reads at a glance, and counts annotate the right margin so the reader sees the sample size behind each fingerprint.

```{r}
#| label: hero-plot-fingerprint
#| message: false
#| warning: false
#| fig.width: 11
#| fig.height: 7.5

class_colors <- paletteer::paletteer_d("PNWColors::Sailboat", n = 5)
fill_map <- c(
  "More severe / likely" = class_colors[5],
  "Inconclusive"         = class_colors[3],
  "No human influence"   = class_colors[2],
  "Less severe / likely" = class_colors[1]
)

label_df <- fingerprint %>%
  group_by(event_type) %>%
  summarise(total = first(total), .groups = "drop")

p <- ggplot(fingerprint,
            aes(y = event_type, x = pct, fill = class_short)) +
  geom_col(width = 0.78, color = "white", linewidth = 0.35) +
  geom_text(
    data = label_df,
    aes(y = event_type, x = 1.015, label = paste0("n = ", total)),
    inherit.aes = FALSE,
    hjust = 0, size = 3.4, color = "grey25", fontface = "plain"
  ) +
  scale_fill_manual(values = fill_map, name = NULL) +
  scale_x_continuous(
    labels = scales::label_percent(accuracy = 1),
    expand = expansion(mult = c(0, 0.10)),
    breaks = seq(0, 1, 0.25)
  ) +
  coord_cartesian(clip = "off") +
  labs(
    title    = "The fingerprint of human influence is not uniform",
    subtitle = "Of 744 attribution studies, ~75% conclude climate change made the event worse — but the split varies\nsharply by event type. Heat events carry a near-unanimous human signature; cold extremes often go the\nother way, consistent with a warming world producing fewer record cold snaps.",
    x        = "Share of studies in each verdict category",
    y        = NULL,
    caption  = "Source: Carbon Brief, 'Mapped: How climate change affects extreme weather around the world' (Nov 2024)  ·  TidyTuesday 2025-08-12"
  ) +
  guides(fill = guide_legend(reverse = FALSE, nrow = 1, byrow = TRUE)) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title          = element_text(face = "bold", size = 17, margin = margin(b = 6)),
    plot.subtitle       = element_text(color = "grey30", size = 11, lineheight = 1.15,
                                       margin = margin(b = 16)),
    plot.caption        = element_text(color = "grey45", size = 8.5,
                                       margin = margin(t = 14), hjust = 0),
    plot.title.position = "plot",
    plot.caption.position = "plot",
    legend.position     = "top",
    legend.justification = "left",
    legend.margin       = margin(b = 6),
    legend.key.height   = unit(10, "pt"),
    legend.key.width    = unit(18, "pt"),
    legend.text         = element_text(size = 10),
    axis.text.y         = element_text(face = "bold", color = "grey15", size = 10.5),
    axis.text.x         = element_text(color = "grey35"),
    axis.title.x        = element_text(color = "grey35", size = 10, margin = margin(t = 8)),
    panel.grid.major.y  = element_blank(),
    panel.grid.minor    = element_blank(),
    panel.grid.major.x  = element_line(color = "grey88", linewidth = 0.3),
    plot.margin         = margin(18, 28, 14, 14)
  )

p
```

::: {.callout-tip}
## Reading the chart

Event types are ordered top-to-bottom by total study count, so the most-studied categories — heat, rain & flooding, drought — dominate the top of the plot. The deepest blue bar segment is the canonical "yes, climate change made this worse" verdict. Look for rows where that segment shrinks (cold/snow/ice, sunshine) or where the lighter "no signal" / "less severe" slices bulge — those are the event classes where attribution science is still genuinely uncertain.
:::

# Final thoughts and takeaways

Three things stuck with me from this dataset:

1. **Heat attribution is essentially settled.** Across 207 studies, ~89% land on "more severe or more likely." When the next record-shattering heatwave makes the news, the prior on a strong climate fingerprint is extremely high — that's not a model assumption, it's the empirical track record of the literature.

2. **Cold extremes are the natural counter-example.** A non-trivial slice of cold/snow/ice studies conclude that climate change made the event *less* likely, not more. This is one of the cleanest physical predictions of a warming world, and the catalogue reflects it.

3. **Storms and oceans remain genuinely hard.** The fingerprint is muddier — more inconclusive and "no influence" results — because the relevant physics (cyclogenesis, ocean dynamics) involves competing signals. Communicators who blanket-attribute hurricanes to climate change are getting ahead of what the literature actually says; communicators who deny any influence are ignoring it.

A few caveats worth naming explicitly:

- **Publication bias likely inflates the "more severe" share.** Null results are harder to get published, especially in a field whose policy relevance is so high. The 75% headline is best read as "of studies that got done and published, this is the verdict mix" — not as an unbiased estimate of a base rate.
- **Geographic coverage is skewed.** Europe, East/Southeast Asia, and North America dominate. The Global South is dramatically under-studied relative to climate exposure — a gap the rapid-attribution community (e.g. World Weather Attribution) is actively trying to close.
- **Carbon Brief curates a single headline classification per study.** Real attribution papers are more nuanced — confidence intervals, partial signals, scenario dependence — than a four-bucket label can capture. Treat the chart as a coarse-grain index, not the last word.

The strongest claim the data actually supports is structural, not numerical: **attribution science has moved from a research curiosity in 2004 to a routine, near-real-time response to extreme events.** Whether the verdict on any *specific* event is "yes," "no," or "maybe" depends on physics that varies by event type — and the fingerprint chart above is a useful map of where the science is confident and where it still hedges.

```{r}
#| label: palette-log
#| echo: false
#| message: false
#| warning: false

palette_log_path <- here::here("posts", "palette-log.csv")
if (!file.exists(palette_log_path)) {
  writeLines("post_date,palette,package,type", palette_log_path)
}
palette_log <- read.csv(palette_log_path)
new_entry <- data.frame(
  post_date = "2025-08-12",
  palette   = "Sailboat",
  package   = "PNWColors",
  type      = "discrete"
)
if (!any(palette_log$post_date == new_entry$post_date &
         palette_log$palette  == new_entry$palette)) {
  write.table(
    new_entry,
    palette_log_path,
    append = TRUE, sep = ",", row.names = FALSE, col.names = FALSE
  )
}
```
