---
title: "Tidy Tuesday: Selected British Literary Prizes"
author: "Sean Thimons"
date: "2025-10-28"
description: "Three decades of British literary prize data — examining gender representation, ethnic diversity trends, and the educational backgrounds that predict literary prestige."
categories: [tidytuesday, R, literature, diversity]
---

# Preface

From [TidyTuesday repository](https://github.com/rfordatascience/tidytuesday/blob/main/data/2025/2025-10-28/readme.md).

> This dataset from the Post45 Data Collective examines 32 years (1990–2022) of British literary award data, documenting author demographics alongside prize information — including gender identity, ethnicity, educational background, and whether authors were shortlisted or won.
>
> -   In which genres are women and ethnically diverse writers most frequently shortlisted or awarded?
> -   Have prizes demonstrated improvement in gender and ethnic representation?
> -   Does a connection exist between specific educational backgrounds and writer selection success?

## Loading necessary packages

My handy booster pack that allows me to install (if needed) and load my usual and favorite packages, as well as some helpful functions.

```{r}
#| label: booster-pack
#| message: false
#| warning: false
#| code-fold: true

# Packages ----------------------------------------------------------------

{
  if (!requireNamespace("pak", quietly = TRUE)) {
    install.packages(
      "pak",
      repos = sprintf(
        "https://r-lib.github.io/p/pak/stable/%s/%s/%s",
        .Platform$pkgType,
        R.Version()$os,
        R.Version()$arch
      )
    )
  }

  install_booster_pack <- function(package, load = TRUE) {
    for (pkg in package) {
      if (!requireNamespace(pkg, quietly = TRUE)) {
        pak::pkg_install(pkg)
      }
      if (load) {
        library(pkg, character.only = TRUE)
      }
    }
  }

  if (file.exists('packages.txt')) {
    packages <- read.table('packages.txt')
    install_booster_pack(package = packages$Package, load = FALSE)
    rm(packages)
  } else {
    booster_pack <- c(
      ### IO ----
      'fs',
      'here',
      'janitor',
      'rio',
      'tidyverse',

      ### EDA ----
      'skimr',

      ### Plot ----
      'ggrepel',
      'ggtext',
      'scales',

      ### Misc ----
      'tidytuesdayR'
    )

    install_booster_pack(package = booster_pack, load = TRUE)
    rm(install_booster_pack, booster_pack)
  }

  # Custom Functions ----

  `%ni%` <- Negate(`%in%`)

  geometric_mean <- function(x) {
    exp(mean(log(x[x > 0]), na.rm = TRUE))
  }

  my_skim <- skim_with(
    numeric = sfl(
      n = length,
      min = ~ min(.x, na.rm = T),
      p25 = ~ stats::quantile(., probs = .25, na.rm = TRUE, names = FALSE),
      med = ~ median(.x, na.rm = T),
      p75 = ~ stats::quantile(., probs = .75, na.rm = TRUE, names = FALSE),
      max = ~ max(.x, na.rm = T),
      mean = ~ mean(.x, na.rm = T),
      geo_mean = ~ geometric_mean(.x),
      sd = ~ stats::sd(., na.rm = TRUE),
      hist = ~ inline_hist(., 5)
    ),
    append = FALSE
  )
}
```

# Load raw data from package

```{r}
#| label: raw-data-load
#| echo: true
#| output: false

raw <- tidytuesdayR::tt_load('2025-10-28')

prizes <- raw$prizes

```

# Exploratory Data Analysis

The `my_skim()` function is a modified version of the `skimr::skim()` function that returns the number of missing data points (cells as `NA`) as well as the inverse (e.g.: number of rows that are *not* `NA`), the count, minimum, 25%, median, 75%, max, mean, geometric mean, and standard deviation. It also generates a little ASCII histogram. Neat!

## Literary Prizes

```{r}
#| label: eda-prizes
#| echo: true
#| output: true

prizes %>%
  my_skim(.)

```

```{r}
#| label: eda-overview
#| echo: true
#| output: true

prizes %>%
  count(prize_alias, sort = TRUE)

prizes %>%
  count(person_role, sort = TRUE)

prizes %>%
  count(gender, sort = TRUE)

prizes %>%
  count(ethnicity_macro, sort = TRUE)

prizes %>%
  count(prize_genre, sort = TRUE)

```

# Diversity in Literary Prizes

## Gender Representation Over Time

```{r}
#| label: gender-trend
#| echo: true
#| output: true

gender_by_year <- prizes %>%
  filter(!is.na(gender), gender %in% c("woman", "man")) %>%
  group_by(prize_year, gender) %>%
  summarize(n = n(), .groups = "drop") %>%
  group_by(prize_year) %>%
  mutate(pct = n / sum(n)) %>%
  ungroup()

# 5-year rolling window
gender_rolling <- gender_by_year %>%
  filter(gender == "woman") %>%
  arrange(prize_year) %>%
  mutate(pct_rolling = zoo::rollmean(pct, k = 5, fill = NA, align = "right"))

```

## Ethnic Diversity Trends

```{r}
#| label: ethnicity-trend
#| echo: true
#| output: true

ethnicity_by_year <- prizes %>%
  filter(!is.na(ethnicity_macro)) %>%
  mutate(diverse = ethnicity_macro %ni% c("White British", "Non-UK White", "Irish")) %>%
  group_by(prize_year) %>%
  summarize(
    n = n(),
    n_diverse = sum(diverse),
    pct_diverse = n_diverse / n,
    .groups = "drop"
  )

ethnicity_by_year %>%
  filter(prize_year >= 2015) %>%
  arrange(prize_year)

```

## Winners vs. Shortlisted by Gender

```{r}
#| label: win-rate-gender
#| echo: true
#| output: true

win_rate <- prizes %>%
  filter(!is.na(gender), gender %in% c("woman", "man")) %>%
  group_by(gender, person_role) %>%
  summarize(n = n(), .groups = "drop") %>%
  group_by(gender) %>%
  mutate(pct = n / sum(n)) %>%
  ungroup()

win_rate

```

## Educational Background

```{r}
#| label: education
#| echo: true
#| output: true

prizes %>%
  filter(!is.na(degree_institution)) %>%
  count(degree_institution, sort = TRUE) %>%
  head(15)

prizes %>%
  filter(!is.na(highest_degree)) %>%
  count(highest_degree, sort = TRUE)

```
## Visualizing Diversity in British Letters

```{r}
#| label: viz-gender-trend
#| message: false
#| warning: false
#| fig.width: 10
#| fig.height: 8

# Subdued literary palette
gender_cols <- c("woman" = "#8E44AD", "man" = "#2C3E50")

ggplot(gender_by_year, aes(x = prize_year, y = pct, fill = gender)) +
  geom_area(alpha = 0.7) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "#888888") +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_x_continuous(breaks = seq(1990, 2022, 4)) +
  scale_fill_manual(values = gender_cols, name = "Gender") +
  labs(
    title = "Gender Balance in British Literary Prizes (1990–2022)",
    subtitle = "Share of shortlisted and winning authors by gender | Dashed line = parity",
    x = "Year",
    y = "Share of Nominations + Wins",
    caption = "Source: TidyTuesday 2025-10-28 | Post45 Data Collective"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 17, color = "#2C3E50"),
    plot.subtitle = element_text(size = 11, color = "#555555"),
    plot.caption = element_text(size = 9, color = "#888888"),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )

```

```{r}
#| label: viz-ethnicity-trend
#| message: false
#| warning: false
#| fig.width: 10
#| fig.height: 6

ggplot(ethnicity_by_year, aes(x = prize_year, y = pct_diverse)) +
  geom_col(fill = "#D35400", alpha = 0.7, width = 0.8) +
  geom_smooth(method = "loess", se = TRUE, color = "#8E44AD", linewidth = 1, alpha = 0.2) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_x_continuous(breaks = seq(1990, 2022, 4)) +
  labs(
    title = "Ethnic Diversity in British Literary Prizes",
    subtitle = "Percentage of non-white shortlisted/winning authors per year | LOESS trend",
    x = "Year",
    y = "% Ethnically Diverse Authors",
    caption = "Source: TidyTuesday 2025-10-28 | Post45 Data Collective"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 17, color = "#2C3E50"),
    plot.subtitle = element_text(size = 11, color = "#555555"),
    plot.caption = element_text(size = 9, color = "#888888"),
    panel.grid.minor = element_blank()
  )

```

# Final thoughts and takeaways

British literary prizes have slowly but measurably diversified over three decades. Gender representation has approached parity in recent years, with women regularly making up half or more of shortlists — a meaningful shift from the male-dominated early 1990s. Ethnic diversity shows a more uneven trajectory, with notable gains in some years but inconsistent progress overall.

The educational background data reveals an uncomfortable truth about literary prestige: Oxbridge graduates remain heavily overrepresented among prize nominees and winners. This suggests that access to literary networks, mentorship, and publishing connections may matter as much as raw talent — a structural advantage that diversity initiatives in prize selection alone cannot fully address.

::: {.callout-note}
The researchers emphasize that this data captures broad patterns and "is not intended as definitive." Demographic classifications — particularly around ethnicity and gender — are based on publicly available information and may not reflect individuals' self-identification.
:::
