---
title: "Tidy Tuesday: Films Based on Video Games"
author: "Sean Thimons"
date: "2026-06-09"
description: "Tracking three decades of video game adaptations from the Uwe Boll trough to the Super Mario Bros. Movie surge."
categories:
  - tidytuesday
  - R
  - film
  - video games
  - reviews
---

```{r}
#| label: pseudo-hero-palette
#| echo: false
#| message: false
#| warning: false
#| fig-width: 14
#| fig-height: 2.8
#| dpi: 320
#| out-width: 100%

if (!requireNamespace("pak", quietly = TRUE)) {
  install.packages("pak", repos = "https://cloud.r-project.org")
}

for (pkg in c("ggplot2", "paletteer", "tibble")) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    pak::pkg_install(pkg)
  }
}

library(ggplot2)
library(paletteer)
library(tibble)

palette_name   <- "cobra"
palette_source <- "NineteenEightyR"
palette_type   <- "discrete"
palette_id     <- paste0(palette_source, "::", palette_name)

hero_swatches <- if (palette_type %in% c("continuous", "diverging")) {
  as.character(paletteer::paletteer_c(palette_id, n = 5))
} else {
  rep_len(as.character(paletteer::paletteer_d(palette_id, n = 5)), 5)
}

swatch_df <- tibble(
  idx  = seq_along(hero_swatches),
  xmin = seq_along(hero_swatches) - 1,
  xmax = seq_along(hero_swatches),
  fill = hero_swatches
)

p_hero <- ggplot(swatch_df) +
  geom_rect(
    aes(xmin = xmin, xmax = xmax, ymin = 0, ymax = 1, fill = fill),
    color = NA,
    show.legend = FALSE
  ) +
  scale_fill_identity() +
  annotate(
    "text",
    x = 0,
    y = -0.22,
    hjust = 0,
    label = paste0("Palette: ", palette_name),
    fontface = "bold",
    size = 4.1
  ) +
  annotate(
    "text",
    x = 5,
    y = -0.22,
    hjust = 1,
    label = paste0("Source: ", palette_source),
    color = "grey35",
    size = 3.6
  ) +
  coord_cartesian(xlim = c(0, 5), ylim = c(-0.35, 1), clip = "off", expand = FALSE) +
  theme_void() +
  theme(plot.margin = margin(5.5, 0, 18, 0))

p_hero
```

# Preface

From [TidyTuesday repository](https://github.com/rfordatascience/tidytuesday/tree/main/data/2026/2026-06-09).

> The dataset this week comes from the Wikipedia article [List of films based on video games](https://en.wikipedia.org/wiki/List_of_films_based_on_video_games). It covers theatrical releases, direct-to-video productions, television films, short films, and documentaries adapted from video games, spanning from the early 1990s to upcoming releases. Each row is a film, with box office figures, critic scores, budgets, and release dates where available.

Suggested angles from the curators:

- Which video game franchise has generated the most film adaptations, and which has earned the most at the box office?
- Which video game publishers have the most film adaptations, and how have they performed at the box office?
- Do audiences and critics agree? Compare CinemaScore grades against Rotten Tomatoes scores.

Thanks to [Georgios Karamanis](https://github.com/gkaramanis) for curating this week's dataset.

## Loading necessary packages

My handy booster pack that allows me to install (if needed) and load my usual and favorite packages, as well as some helpful functions.

```{r}
#| label: booster-pack
#| message: false
#| warning: false
#| code-fold: true

# Packages ----------------------------------------------------------------

{
  # Install pak if it's not already installed
  if (!requireNamespace("pak", quietly = TRUE)) {
    install.packages("pak", repos = "https://cloud.r-project.org")
  }

  # CRAN Packages ----
  install_booster_pack <- function(package, load = TRUE) {
    for (pkg in package) {
      if (!requireNamespace(pkg, quietly = TRUE)) {
        pak::pkg_install(pkg)
      }
      if (load) {
        library(pkg, character.only = TRUE)
      }
    }
  }

  core_packages <- c(
    'fs',
    'here',
    'janitor',
    'rio',
    'tidyverse',
    'skimr',
    'paletteer',
    'tidytuesdayR'
  )

  optional_packages <- c('ggrepel', 'ggtext', 'scales')

  install_booster_pack(package = unique(c(core_packages, optional_packages)), load = TRUE)
  rm(install_booster_pack, core_packages, optional_packages)

  # Custom Functions ----

  `%ni%` <- Negate(`%in%`)

  geometric_mean <- function(x) {
    exp(mean(log(x[x > 0]), na.rm = TRUE))
  }

  my_skim <- skim_with(
    numeric = sfl(
      n = length,
      min = ~ min(.x, na.rm = T),
      p25 = ~ stats::quantile(., probs = .25, na.rm = TRUE, names = FALSE),
      med = ~ median(.x, na.rm = T),
      p75 = ~ stats::quantile(., probs = .75, na.rm = TRUE, names = FALSE),
      max = ~ max(.x, na.rm = T),
      mean = ~ mean(.x, na.rm = T),
      geo_mean = ~ geometric_mean(.x),
      sd = ~ stats::sd(., na.rm = TRUE),
      hist = ~ inline_hist(., 5)
    ),
    append = FALSE
  )
}

```

# Load raw data from package

```{r}
#| label: raw-data-load
#| echo: true
#| output: false

raw <- tidytuesdayR::tt_load('2026-06-09')

game_films <- raw$game_films

```

# Exploratory Data Analysis

The `my_skim()` function is a modified version of the `skimr::skim()` function that returns the number of missing data points (cells as `NA`) as well as the inverse, the count, minimum, 25%, median, 75%, max, mean, geometric mean, and standard deviation. It also generates a little ASCII histogram.

## Shape of the catalogue

```{r}
#| label: eda-shape
#| echo: true
#| output: true

cat(sprintf("rows: %d   cols: %d\n", nrow(game_films), ncol(game_films)))

game_films %>%
  count(category, sort = TRUE)

game_films %>%
  filter(category == "Theatrical releases") %>%
  count(subcategory, sort = TRUE)

```

Theatrical releases dominate (246 of 439 rows). Documentaries and direct-to-video each take a sizeable slice, but they live in a different world from the wide-release theatrical adaptations that drive cultural attention.

## Numeric profile

```{r}
#| label: eda-skim
#| echo: true
#| output: true

game_films %>%
  select(release_date, worldwide_box_office, rotten_tomatoes, metacritic, budget_low, budget_high) %>%
  my_skim()

```

Three honest realities:

- **Sparse scores.** Most films have neither a Rotten Tomatoes score nor a CinemaScore. The graded universe is small enough that you can almost name every film in it.
- **Currency mix.** Box office reports come in `$` (82 films) and `¥` (14 films). I'll keep the analysis to USD so the y-axis is interpretable.
- **Heavy right tail.** Box office is hugely skewed — the median USD gross is ~$84M but the max is $5B (when both Mario films are added together). One blockbuster can drag every aggregate.

# The video game movie curse — and the moment it broke

The cultural story of video game movies has a familiar shape: a wave of cheap, critically panned adaptations through the 1990s and 2000s (the [Uwe Boll](https://en.wikipedia.org/wiki/Uwe_Boll) era), then a slow re-engineering through the 2010s, and finally a commercial breakout starting with *Sonic the Hedgehog* (2020) and culminating in *The Super Mario Bros. Movie* in 2023.

Let me check whether the data actually supports that arc.

## Era table

```{r}
#| label: era-table
#| echo: true
#| output: true

theatrical <- game_films %>%
  filter(category == "Theatrical releases") %>%
  mutate(
    year = lubridate::year(release_date),
    era = case_when(
      year < 2000 ~ "1990s",
      year < 2010 ~ "2000s",
      year < 2020 ~ "2010s",
      year >= 2020 ~ "2020s",
      TRUE        ~ NA_character_
    ),
    era = factor(era, levels = c("1990s", "2000s", "2010s", "2020s"))
  )

stopifnot(
  "theatrical has 0 rows" = nrow(theatrical) > 0,
  "era never matched" = sum(!is.na(theatrical$era)) > 0
)

era_summary <- theatrical %>%
  filter(!is.na(era)) %>%
  group_by(era) %>%
  summarise(
    n_films       = n(),
    n_with_rt     = sum(!is.na(rotten_tomatoes)),
    median_rt     = median(rotten_tomatoes, na.rm = TRUE),
    mean_rt       = round(mean(rotten_tomatoes, na.rm = TRUE), 1),
    median_bo_usd = median(worldwide_box_office[worldwide_box_office_currency == "$"], na.rm = TRUE),
    .groups       = "drop"
  )

era_summary

```

The pattern is real. The median critic score climbs from **17%** (1990s) to **51%** (2020s), and median worldwide gross — for the USD-denominated subset — roughly triples between the 2000s and the 2020s. The 2000s were genuinely the trough of both quality *and* commercial pull.

## Publisher league table

```{r}
#| label: publisher-table
#| echo: true
#| output: true

publishers_usd <- game_films %>%
  filter(
    worldwide_box_office_currency == "$",
    !is.na(original_game_publisher),
    !is.na(worldwide_box_office)
  ) %>%
  group_by(original_game_publisher) %>%
  summarise(
    n_films  = n(),
    total_bo = sum(worldwide_box_office),
    median_rt = median(rotten_tomatoes, na.rm = TRUE),
    .groups  = "drop"
  ) %>%
  arrange(desc(total_bo))

stopifnot("publishers_usd empty" = nrow(publishers_usd) > 0)

publishers_usd %>%
  slice_head(n = 10)

```

Nintendo's three theatrical releases — the original 1993 *Super Mario Bros.*, the 2023 *Super Mario Bros. Movie*, and the upcoming *Super Mario Galaxy Movie* — together gross more than Capcom's fifteen films combined. The Pokémon Company joint adds another billion. The catalogue is profoundly top-heavy.

# Visualizing the curse breaking

The plot below puts every USD-denominated theatrical release on a release-date × Rotten-Tomatoes plane. Point area is worldwide box office, colour is era, and a loess trend tracks the long arc of critical reception.

```{r}
#| label: plot-data
#| echo: true
#| output: true

plot_df <- theatrical %>%
  filter(
    worldwide_box_office_currency == "$",
    !is.na(release_date),
    !is.na(rotten_tomatoes),
    !is.na(worldwide_box_office),
    !is.na(era)
  )

cat(sprintf("plot_df: %d rows\n", nrow(plot_df)))
stopifnot("plot_df has 0 rows — check filter values" = nrow(plot_df) > 0)

label_titles <- c(
  "Super Mario Bros.",
  "Mortal Kombat",
  "Lara Croft: Tomb Raider",
  "Warcraft",
  "Pokémon Detective Pikachu",
  "Sonic the Hedgehog",
  "The Super Mario Bros. Movie",
  "Sonic the Hedgehog 3",
  "A Minecraft Movie"
)

label_df <- plot_df %>%
  filter(title %in% label_titles)

cat(sprintf("label_df: %d / %d hits\n", nrow(label_df), length(label_titles)))

```

```{r}
#| label: hero-plot
#| message: false
#| warning: false
#| fig-width: 11
#| fig-height: 7.5

era_colors <- paletteer::paletteer_d("NineteenEightyR::cobra", n = 5)[c(1, 2, 3, 5)]
names(era_colors) <- c("1990s", "2000s", "2010s", "2020s")

p <- ggplot(plot_df, aes(x = release_date, y = rotten_tomatoes)) +
  geom_hline(yintercept = 60, linetype = "dashed", color = "grey50", linewidth = 0.4) +
  annotate(
    "text",
    x = as.Date("1993-06-01"),
    y = 62.5,
    label = "Fresh threshold (60%)",
    hjust = 0, vjust = 0,
    size = 3.1, color = "grey35", fontface = "italic"
  ) +
  geom_smooth(
    method = "loess",
    se = TRUE,
    color = "grey20",
    fill = "grey80",
    alpha = 0.35,
    linewidth = 0.7
  ) +
  geom_point(
    aes(size = worldwide_box_office, fill = era),
    shape = 21, color = "white", stroke = 0.4, alpha = 0.9
  ) +
  ggrepel::geom_text_repel(
    data = label_df,
    aes(label = title),
    size = 3.1,
    family = "sans",
    color = "grey15",
    box.padding = 0.55,
    point.padding = 0.35,
    min.segment.length = 0,
    segment.color = "grey50",
    segment.size = 0.3,
    max.overlaps = Inf,
    seed = 42
  ) +
  scale_fill_manual(values = era_colors, name = "Release era") +
  scale_size_area(
    name = "Worldwide gross",
    max_size = 16,
    breaks = c(5e7, 2e8, 5e8, 1e9),
    labels = scales::label_dollar(scale = 1e-6, suffix = "M", accuracy = 1)
  ) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 20), labels = function(x) paste0(x, "%")) +
  guides(
    fill = guide_legend(override.aes = list(size = 5)),
    size = guide_legend(override.aes = list(fill = "grey60"))
  ) +
  labs(
    title    = "The video game movie curse is finally breaking",
    subtitle = "Theatrical adaptations by release date, Rotten Tomatoes score, and worldwide box office (USD)",
    x = NULL,
    y = "Rotten Tomatoes critic score",
    caption = "Source: Wikipedia via TidyTuesday 2026-06-09 · Theatrical releases priced in USD only"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title          = element_text(face = "bold", size = 16, margin = margin(b = 4)),
    plot.subtitle       = element_text(color = "grey30", margin = margin(b = 14)),
    plot.caption        = element_text(color = "grey45", size = 8.5, hjust = 0, margin = margin(t = 12)),
    panel.grid.minor    = element_blank(),
    panel.grid.major    = element_line(color = "grey92"),
    legend.position     = "bottom",
    legend.box          = "horizontal",
    legend.title        = element_text(face = "bold", size = 9),
    legend.text         = element_text(size = 9),
    plot.background     = element_rect(fill = "white", color = NA),
    panel.background    = element_rect(fill = "white", color = NA)
  )

p

```

The loess line spends nearly two decades parked in the low 20s — the genre genuinely earned its reputation as critic-proof shovelware. Around 2018 it lifts off, and *Sonic the Hedgehog 3* (86%) is the first adaptation in the dataset to clear 80% Fresh while also clearing $400M worldwide. *The Super Mario Bros. Movie* did not need critical love to make $1.36B; it broke the commercial half of the curse all on its own.

## A second look: era distributions

```{r}
#| label: era-violin
#| message: false
#| warning: false
#| fig-width: 9
#| fig-height: 5

violin_df <- theatrical %>%
  filter(!is.na(rotten_tomatoes), !is.na(era))

stopifnot("violin_df has 0 rows" = nrow(violin_df) > 0)

p2 <- ggplot(violin_df, aes(x = era, y = rotten_tomatoes, fill = era)) +
  geom_violin(alpha = 0.55, color = NA, trim = FALSE) +
  geom_boxplot(width = 0.18, outlier.shape = NA, color = "grey15", fill = "white", alpha = 0.9) +
  geom_jitter(width = 0.08, alpha = 0.55, size = 1.4, color = "grey20") +
  geom_hline(yintercept = 60, linetype = "dashed", color = "grey50", linewidth = 0.4) +
  scale_fill_manual(values = era_colors, guide = "none") +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 20), labels = function(x) paste0(x, "%")) +
  labs(
    title    = "Critics warmed up one decade at a time",
    subtitle = "Distribution of Rotten Tomatoes scores for theatrical video game adaptations",
    x = NULL,
    y = "Rotten Tomatoes critic score",
    caption = "Source: Wikipedia via TidyTuesday 2026-06-09"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold", size = 15, margin = margin(b = 4)),
    plot.subtitle    = element_text(color = "grey30", margin = margin(b = 14)),
    plot.caption     = element_text(color = "grey45", size = 8.5, hjust = 0, margin = margin(t = 12)),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(color = "grey92"),
    plot.background  = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  )

p2

```

The 2000s violin is the worst of the bunch — a fat cluster of films below 30% with no upper tail to speak of. The 2020s violin is wider and taller: the failures still happen (*Borderlands* and *Five Nights at Freddy's* both undershot), but the ceiling has lifted enough that a *Sonic 3* or *Detective Pikachu* outcome is now plausible rather than freakish.

# Final thoughts and takeaways

Three readings of the data feel defensible:

1. **The curse is real, but it was a quality curse before it was a commercial one.** Box office crept up through the 2010s while critics still hated almost everything. The critical turn — median RT going from 18% in the 2000s to 51% in the 2020s — is the bigger structural change.
2. **Nintendo plays a different game.** Three theatrical films, $2.2B in worldwide gross, with the *Super Mario Bros. Movie* alone clearing a billion. Nothing else in the catalogue scales like that, and it skews every per-publisher aggregate.
3. **Sample sizes are tiny.** Only ~70 theatrical releases have both a USD box office figure and a Rotten Tomatoes score. The era trend is suggestive, not a clinical demonstration, and any one studio's slate (Sony's PlayStation Productions, Legendary's *Detective Pikachu* / *Pokémon* line, Paramount's *Sonic* franchise) can move the medians more than a genuine industry-wide shift would.

The next decade should tell us whether *Sonic 3* and the upcoming *Super Mario Galaxy Movie* (already $832M with a 43% RT) are the new normal, or whether 2023–2026 was the surge before the cycle resets.

```{r}
#| label: palette-log-append
#| echo: false
#| message: false
#| warning: false

palette_log_path <- here::here("posts", "palette-log.csv")
if (file.exists(palette_log_path)) {
  palette_log <- read.csv(palette_log_path, stringsAsFactors = FALSE)
  new_entry <- data.frame(
    post_date = "2026-06-09",
    palette   = "cobra",
    package   = "NineteenEightyR",
    type      = "discrete",
    stringsAsFactors = FALSE
  )
  if (!any(palette_log$post_date == new_entry$post_date &
           palette_log$palette   == new_entry$palette)) {
    write.table(
      new_entry,
      palette_log_path,
      append = TRUE, sep = ",", row.names = FALSE, col.names = FALSE
    )
  }
  invisible(NULL)
}
```
